### 2020-03-25 paper/paper-ldpred2.tex
init

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}

%\renewcommand{\arraystretch}{1.8}

\usepackage{xr}
\externaldocument{paper-ldpred2-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
\author{Florian Priv\'e,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

\noindent Contacts:
\begin{itemize}
\item \url{florian.prive.21@gmail.com}
\item \url{bjv@econ.au.dk}
\end{itemize}

\clearpage

\abstract{

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Material and Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage
\vspace*{5em}

\section*{Software and code availability}

%All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
%R packages bigsnpr and bigutilsr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
%A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.

\section*{Acknowledgements}

This research has been conducted using the UK Biobank Resource under Application Number 41181.

F.P. and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof. John McGrath), and also acknowledge the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).

\section*{Declaration of Interests}

The authors declare no competing interests.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2020-03-28 paper/paper-ldpred2.tex
start intro

@@
 
-\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
+\title{LDpred2: better, faster, stronger}
 \author{Florian Priv\'e,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
@@
 
-\clearpage
+\vspace*{4em}
 
@@
 
+The use of polygenic scores (PGS) is widespread now.
+There are high hopes for using these scores, which sum up the information of many genetic variants into a single score, into clinical practice. 
+Predictive performance of PGS is inexorably bounded by heritability, the proportion of phenotypic variance that is attributable to genetics, so that their possibility of being used as a diagnostic tool for diseases is limited.
+Nevertheless, it should be possible to use them to stratify the population into groups of different risk levels, to be able to identify individuals at high risk for a given disease [CITATION]. 
+These people should then be monitored more closely to detect the disease and intervene sooner to maximize chances of remission (e.g.\ in the case of cancer).
+PGS are also extensively used in epidemiology and economics as predictive variables of interest.
+For example, a recently derived PGS for education attainment has been the most predictive variable in social sciences so far [CITATION].
+
+LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a Linkage Disequilibrium (LD) matrix only \cite[]{vilhjalmsson2015modeling}.
+It assumes there is a proportion $p$ of variants that are causal.
+However, LDpred currently has several limitations that may result in limited predictive performance.
+The non-infinitesimal version of LDpred, a Gibbs sampler, is particularly sensitive to model misspecification when applied to summary statistics with large sample sizes.
+It may also be very unstable in long-range LD regions such as the human leukocyte antigen (HLA) region of chromosome 6.
+This issue has lead to the removal of this region from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is a shame since this region of the genome contains several disease-associated variants, particularly with autoimmune diseases and psychiatric disorders [CITATIONS].
+LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as in its model specification.
+
+Here, we present LDpred2, a new version of LDpred, that aims to cope with the issues presented before.
+We show that [DESCRIBE RESULTS].
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Results}
+
+\subsection{Overview of Methods}
 

### 2020-03-31 paper/paper-ldpred2.tex
save

@@
 \title{LDpred2: better, faster, stronger}
-\author{Florian Priv\'e,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
 
@@
 \noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
+\noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 Here, we present LDpred2, a new version of LDpred, that aims to cope with the issues presented before.
+We provide a faster and more robust implementation of LDpred in R package bigsnpr \cite[]{prive2017efficient}.
+We also provide two new options: a ``sparse'' option, where LDpred2 truly fit some effects to be zero, therefore providing a sparse vector of effects; and an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the (SNP) heritability $h^2$, therefore for which there is no need for validation data to choose hyper-parameters performing best.
 We show that [DESCRIBE RESULTS].
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Material and Methods}
 
@@
 
-\subsection{Overview of Methods}
+\subsection*{Overview of Methods}
 
@@
 
-%\clearpage
-\vspace*{5em}
+\section{Online Methods}
+
+\subsection*{Overview of LDpred model}
+
+LDpred assumes the following model for effect sizes,
+\begin{equation}\label{eq:model}
+\beta_j \sim \left\{
+\begin{array}{ll}
+\mathcal N\left(0, \dfrac{h^2}{M p}\right) & \mbox{with probably $p$,} \\
+0 & \mbox{otherwise,}
+\end{array}
+\right.
+\end{equation}
+where $p$ is the proportion of causal variants, $M$ the number of variants and $h^2$ the (SNP) heritability.
+\cite{vilhjalmsson2015modeling} estimates $h^2$ using constrained LD score regression (intercept fixed to 1) and recommend testing a grid of hyper-parameter values for $p$ (e.g.\ 1, 0.3, 0.1, 0.03, 0.01, etc.).
+
+\newcommand{\hmpn}{\dfrac{h^2}{M p} + \dfrac{1}{N}}
+\newcommand{\phmpn}{\dfrac{p}{\sqrt{\hmpn}}}
+\newcommand{\betahmpn}{\dfrac{\tilde{\beta}_j^2}{\hmpn}}
+
+To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler [REALLY A GS?]. They compute the probability that variant $j$ is causal as 
+\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
+which we rewrite as
+\begin{equation}
+\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.
+\end{equation}
+
+Then, $\beta_j$ is sampled according to
+\begin{equation}\label{eg:beta_random}
+\beta_j ~|~ \tilde{\beta}_j \sim \left\{
+\begin{array}{ll}
+\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{N h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{N h^2}}\dfrac{1}{N}\right) & \mbox{with probably $\bar{p}_j$,} \\
+0 & \mbox{otherwise.}
+\end{array}
+\right.
+\end{equation}
+
+[TO COMPLETE + CLEARER -> SHOULD SHOW THAT IT IS ITERATIVE]
+
+\subsection*{New LDpred2 models}
+
+LDpred2 comes with two extensions of the LDpred model.
+
+The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h2$ using constrained LD score regression \cite[]{bulik2015ld}.
+This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
+To estimate $p$, we count the number of causal variants (i.e.\ $S = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:beta_random}). 
+We can assume that $S \sim B(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the conjugate prior $p \sim \text{Beta}(1 + S, 1 + M - S)$.
+Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$. Instead, we estimate $h^2 = \beta^T K \beta$, where $K$ is the correlation matrix.
+
+The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e. that some resulting effects are exactly 0.
+When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j = 0$.
+Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the $p$ and $h^2$ estimates from LDpred2-auto.
+
+\subsection*{New strategy for local correlation}
+
+LDpred has a window size parameter that needs to be set; for a given variant, correlations with other variants outside of this window are assumed to be 0.
+The recommended value for this window (in number of variants) is to use the total number of variants divided by 3000, which corresponds to a window radius of around 2 Mb \cite[]{vilhjalmsson2015modeling}.
+After investigating, we come to the conclusion that this window size is not large enough. Indeed, the human leukocyte antigen (HLA) region of chromosome 6 is 8 Mb long \cite[]{price2008long}. 
+Using a window of 8Mb would be computationally and memory inefficient.
+Instead, we propose to use genetic distances. Genome-wide, 1 Mb corresponds on average to 1 cM. Yet, the HLA region is only 3 cM long (vs.\ 8 Mb long).
+Therefore, genetic distances enable to capture the same LD at a smaller distance.
+We provide function \texttt{snp\_asGeneticPos()} in package bignspr to easily interpolate physical position to genetic positions.
+We recommend to use genetic positions and to use a size parameter of 3 cM when computing the correlation between variants for LDpred2. 
+Note that, in the code, we use \texttt{size = 3 / 1000} as parameter \texttt{size} is internally multiplied by 1000 in functions of package bigsnpr.
+
+\subsection*{New strategy for other hyper-parameters}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+%\vspace*{5em}
 

### 2020-04-08 paper/paper-ldpred2.tex
save

@@
 
+\subsection*{From marginal effects to joint effects}
+
+In this section, we explain how we can obtain joint effects from summary statistics (marginal effects) and a correlation matrix.
+Let us denote by $S$ the diagonal matrix with standard deviations of the $p$ variants, $C_n = I_n - 1 1^T / n$ the centering matrix, $G$ the genotype matrix of $n$ individuals and $p$ variants, and $y$ the phenotype vector for $n$ individuals.
+Then, the marginal effects (assuming no other covariates than the intercept) are given by 
+\[\hat{\beta}_{marg} = S^{-2} G^T C_n y ~,\]
+while the joint effects are obtained by solving
+\[\begin{bmatrix} \hat{\alpha} \\ \hat{\beta}_{joint} \end{bmatrix} = \left(\begin{bmatrix} 1 & G \end{bmatrix}^T \begin{bmatrix} 1 & G \end{bmatrix}\right)^{-1} \begin{bmatrix} 1 & G \end{bmatrix}^T y ~.\]
+Using the Woodburry formula, we get 
+\[\hat{\beta}_{joint} = (G^T C_n G)^{-1} G^T C_n y ~.\]
+We further note that the correlation matrix of $G$ is \(K =  S^{-1} G^T C_n G  S^{-1}\).
+Then we get $\hat{\beta}_{joint} = S^{-1} K^{-1} S \hat{\beta}_{marg}$.
+
+For one marginal effect $\hat{\beta}$, let us denote by $y^*$ and $g^*$ the corresponding phenotype and genotype (residualized from $K$ covariates, e.g.\ centering them).
+Then,
+$\text{var}(\hat{\beta}) = \dfrac{(y^* - \hat{\beta} g^*)^T (y^* - \hat{\beta} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
+Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{sd}(\hat{\beta}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\beta}\right) \approx \dfrac{\hat{\beta}}{\text{sd}(\hat{\beta})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.
+
+Let us go back to the formula we derived before: $\hat{\beta}_{joint} = S^{-1} K^{-1} S \hat{\beta}_{marg}$.
+As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$.
+Then it justifies the use of Z-scores ($\hat{\beta} / \text{sd}(\hat{\beta})$) divided by $\sqrt{N}$ as input for LDpred.
+Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{sd}(\hat{\beta}) ~ \sqrt{N}\right)$.
+Note that LDpred (v1) and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.
+
 \subsection*{Overview of LDpred model}
@@
 Instead, we propose to use genetic distances. Genome-wide, 1 Mb corresponds on average to 1 cM. Yet, the HLA region is only 3 cM long (vs.\ 8 Mb long).
-Therefore, genetic distances enable to capture the same LD at a smaller distance.
+Therefore, genetic distances enable to capture the same LD using a globally smaller window.
 We provide function \texttt{snp\_asGeneticPos()} in package bignspr to easily interpolate physical position to genetic positions.
@@
 
-\subsection*{New strategy for other hyper-parameters}
+\subsection*{New strategy for running LDpred2}
+
+We recommend to run LDpred2 per chromosome.
+Even if it is possible to run LDpred2 genome-wide, this approach has two limitations.
+First, it can be memory and computationally demanding to do so.
+For around one million (1M) variants, storing the 1M$\times$1M sparse correlation matrix takes more than 32 GB of memory.
+Doubling to 2M variants would require 128 GB of RAM to store the matrix.
+Second, as noted in \cite{prive2019making}, it may be beneficial to assume that architecture of traits may be different for different chromosomes.
+For example, autoimmune diseases clearly show larger effects on chromosome 6. compared to other chromosomes.
+Assuming the same model for genetic effects genome-wide could results in severe model misspecification, which would lead to suboptimal predictive performance.
+Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combining the results.
+
+\subsection*{Simulation analyses}
+
+We use the UK Biobank data for both real data analyses and simulations.
+We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled).
+We restrict variants to HapMap3 variants.
+This results in 337,475 individuals and 1,194,574 variants.
+We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and computing correlations between variants. 
+We use 300,000 other individuals for running logistic GWAS to create summary statistics.
+We use the remaining 27,475 individuals as test set for evaluating models.
+
+\subsection*{Real data analyses}
+
+We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 327,475 individuals as test set.
+We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
+We defined phenotypes as done in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section).
+
+%% TABLE 1 HERE
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|l|c|c|c|c|}
+		\hline
+		Trait & GWAS citation & GWAS sample size & GWAS \#variants \\
+		\hline
+		Breast cancer (BRCA) & \cite{michailidou2017association} & 137,045 / 119,078 & 11,792,542 \\
+		Rheumatoid arthritis (RA) & \cite{okada2014genetics} & ~~29,880 / ~~73,758 & ~~9,739,303 \\
+		Type 1 diabetes (T1D) & \cite{censin2017childhood} & ~~~~~5913 / ~~~~~8828  & ~~8,996,866 \\
+		Type 2 diabetes (T2D) & \cite{scott2017expanded} & ~~26,676 / 132,532 & 12,056,346 \\
+		Prostate cancer (PRCA) & \cite{schumacher2018association} & ~~79,148 / ~~61,106 & 20,370,946 \\
+		Depression (MDD) & \cite{wray2018genome} & ~~59,851 / 113,154 & 13,554,550 \\
+		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & ~~60,801 / 123,504 & ~~9,455,778 \\
+		Asthma & \cite{demenais2018multiancestry} & ~~19,954 / 107,715 & ~~2,001,280 \\
+		\hline
+	\end{tabular}
+	\caption{Summary of external GWAS summary statistics used. The GWAS sample size is the number of cases / controls in the GWAS. \label{tab:sumstats}}
+\end{table}
 
@@
 
-%All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
-%R packages bigsnpr and bigutilsr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
-%A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.
+All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}.
+R package bigsnpr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
+A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
 

### 2020-04-14 paper/paper-ldpred2.tex
save

@@
 \usepackage{graphicx}
-\graphicspath{{figures/}}
+\graphicspath{{../figures/}}
 \DeclareMathOperator*{\argmin}{\arg\!\min}
@@
 
-\subsection*{Overview of Methods}
+\subsection*{Overview of methods}
+
+Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
+LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation test; 3) LDpred3-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a truly sparse version of LDpred; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameter to optimize.
+As a recall, LDpred1 has two options: LDpred1-inf and LDpred1-grid, where only $p$ is optimized.
+
+We compare the two versions of LDpred using 6 simulation scenarios to understand in which cases it is important to use the new version.
+We also compare these two versions in 8 binary phenotypes of major interest using published external summary statistics and the UK Biobank data.
+
+\subsection*{Simulations}
+
+Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
+6 simulation scenarios are used, each repeated 10 times.
+In the first 4 simulations scenarios, a heritability of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
+In these scenarios, infinitesimal models are performing similarly.
+When testing a grid of parameters, LDpred2 performed substantially better than LDpred1 in the cases where the number of causal variations is smaller, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3).
+For example, in simulations with 300 causal variants, AUC of 73.5\% is obtained with LDpred1 while an AUC of 81.9\% is obtained with LDpred2.
+In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
+In HLA scenarios, i.e.\ where causal variants are sampled in a region with variants at high LD, LDpred1 perform poorly.
+In these HLA scenarios, all LDpred2 models perform well, although LDpred2-auto perform slightly worse than other models.
+
+%% FIGURE 1 HERE
+\begin{figure}[h]
+\centerline{\includegraphics[width=0.95\textwidth]{AUC-simu}}
+\caption{Results of the 6 simulation scenarios. 
+Scenarios are named with 4 parts, separated by underscore: 1) ``all'' means that causal variants were randomly sampled anywhere on the genome while ``HLA'' means that they were sampled in the HLA region of chromosome 6 (25-35 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
+Bars present the mean and 95\% CI of $10^4$ non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
+%See corresponding values in table S1.
+}
+\label{fig:AUC-simu}
+\end{figure}
+
+\subsection*{Real data}
+
+Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr).
+8 binary phenotypes are used, summarized in table \ref{tab:sumstats}.
+For T1D, T2D, BRCA, PRCA and MDD, all LDpred2 models perform much better than LDpred1.
+For example, for BRCA, AUC improves from 58.5\% with LDpred1 to 65\% with LDpred2, and from 55\% to 75\% for T1D.
+For Asthma, CAD and RA, predictive performances of LDpred1 and LDpred2 are similar.
+
+As in simulations, the sparse version of LDpred2 performs equally well as the original version, with the exception of PRCA, for which it performs poorly.
+For CAD, PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly.
+
+%% FIGURE 2 HERE
+\begin{figure}[h]
+\centerline{\includegraphics[width=0.85\textwidth]{AUC-real}}
+\caption{Results of the real data applications using published external summary statistics.
+Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
+%See corresponding values in table S2.
+}
+\label{fig:AUC-real}
+\end{figure}
 
@@
 
-\section{Online Methods}
+\section{Methods}
+
+\subsection*{Simulation analyses}
+
+We use the UK Biobank data for both real data analyses and simulations.
+We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled).
+We restrict variants to HapMap3 variants.
+This results in 337,475 individuals and 1,194,574 variants.
+We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and computing correlations between variants (LD matrix). 
+We use 300,000 other individuals for running logistic GWAS to create summary statistics.
+We use the remaining 27,475 individuals as test set for evaluating models.
+
+\subsection*{Real data analyses}
+
+We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 327,475 individuals as test set.
+We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
+We defined phenotypes as done in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section).
+
+%% TABLE 1 HERE
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|l|c|c|c|c|}
+		\hline
+		Trait & GWAS citation & GWAS sample size & GWAS \#variants \\
+		\hline
+		Breast cancer (BRCA) & \cite{michailidou2017association} & 137,045 / 119,078 & 11,792,542 \\
+		Rheumatoid arthritis (RA) & \cite{okada2014genetics} & ~~29,880 / ~~73,758 & ~~9,739,303 \\
+		Type 1 diabetes (T1D) & \cite{censin2017childhood} & ~~~~~5913 / ~~~~~8828  & ~~8,996,866 \\
+		Type 2 diabetes (T2D) & \cite{scott2017expanded} & ~~26,676 / 132,532 & 12,056,346 \\
+		Prostate cancer (PRCA) & \cite{schumacher2018association} & ~~79,148 / ~~61,106 & 20,370,946 \\
+		Depression (MDD) & \cite{wray2018genome} & ~~59,851 / 113,154 & 13,554,550 \\
+		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & ~~60,801 / 123,504 & ~~9,455,778 \\
+		Asthma & \cite{demenais2018multiancestry} & ~~19,954 / 107,715 & ~~2,001,280 \\
+		\hline
+	\end{tabular}
+	\caption{Summary of external GWAS summary statistics used. The GWAS sample size is the number of cases / controls in the GWAS. \label{tab:sumstats}}
+\end{table}
 
@@
 
-\subsection*{Simulation analyses}
-
-We use the UK Biobank data for both real data analyses and simulations.
-We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled).
-We restrict variants to HapMap3 variants.
-This results in 337,475 individuals and 1,194,574 variants.
-We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and computing correlations between variants. 
-We use 300,000 other individuals for running logistic GWAS to create summary statistics.
-We use the remaining 27,475 individuals as test set for evaluating models.
-
-\subsection*{Real data analyses}
-
-We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 327,475 individuals as test set.
-We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We defined phenotypes as done in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section).
-
-%% TABLE 1 HERE
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|l|c|c|c|c|}
-		\hline
-		Trait & GWAS citation & GWAS sample size & GWAS \#variants \\
-		\hline
-		Breast cancer (BRCA) & \cite{michailidou2017association} & 137,045 / 119,078 & 11,792,542 \\
-		Rheumatoid arthritis (RA) & \cite{okada2014genetics} & ~~29,880 / ~~73,758 & ~~9,739,303 \\
-		Type 1 diabetes (T1D) & \cite{censin2017childhood} & ~~~~~5913 / ~~~~~8828  & ~~8,996,866 \\
-		Type 2 diabetes (T2D) & \cite{scott2017expanded} & ~~26,676 / 132,532 & 12,056,346 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & ~~79,148 / ~~61,106 & 20,370,946 \\
-		Depression (MDD) & \cite{wray2018genome} & ~~59,851 / 113,154 & 13,554,550 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & ~~60,801 / 123,504 & ~~9,455,778 \\
-		Asthma & \cite{demenais2018multiancestry} & ~~19,954 / 107,715 & ~~2,001,280 \\
-		\hline
-	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used. The GWAS sample size is the number of cases / controls in the GWAS. \label{tab:sumstats}}
-\end{table}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-04-14 paper/paper-ldpred2.tex
save

@@
 
+Here we present LDpred2, a new version of LDpred. 
+LDpred is widely used and has the potential to provide polygenic models with good predictive performance \cite{khera2018genome}. 
+Yet, it has some instability issues.
+We propose a new version to solve these instability issues and try to improve other points as well.
+
+We show that LDpred2 is much more stable and provides higher prediction than LDpred1.
+We demonstrate that LDpred1 has some large issues due to long-range LD regions such as the HLA region.
+Indeed, both LDpred1 and LDpred1-inf perform poorly in the simulations where causal variants are the HLA region.
+In constrast, LDpred2 perform very well. We hypothesize that LDpred1 does not use a window size that is large enough to account for long-range LD such as in the HLA region.
+In LDpred2, we use a window size of 3 cM, which is much larger and enable LDpred2 to work well even when causal variants are in long-range LD regions.
+We hypothesize that using a larger LD window explains the 5 real traits where LDpred2 performs much better than LDpred1, especially because LDpred2-inf performs better than LDpred1-inf.
+We strongly discourage against removing these regions as suggested by others as these regions, especially the HLA region, contain lots of variants associated with many traits such as autoimmune diseases and psychiatric disorders, and are therefore very useful for prediction.
+
+In LDpred2, we also test more values for $p$ (17 instead of 7 by default) and for $h^2$ (3 instead of 1).
+When testing the grid of hyper-parameters of $p$ and $h2$, we also allow for testing an option to enable sparse models in LDpred2 (see after).
+Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1, we use 5 times as many iterations, use a larger window size for computing correlation, yet it is LDpred2 is still faster than LDpred1.
+LDpred2 is faster than LDpred1 because we provide a very efficient parallel implementation in C++.
+The parallelization is performed on the grid of parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see methods).
+As an example, it takes less than two hours to run LDpred2-inf, LDpred2-grid (108 hyper-parameters values) and LDpred2-auto for a chromosome with 100K variants and using 8 cores, and only 15 minutes for a chromosome with 20K variants.
+Pre-computing the LD matrix for a chromosome with 100K variants takes 11 minutes using 16 cores.
+
+LDpred2 also comes with two new models.
+When the sparse option is enabled, it provide models that are truly sparse, as compared to LDpred1 which outputs very small non-zero effect sizes \cite{cecile2019polygenic}.
+It can provide sparse effect sizes as much as 98\% for small $p$, while keeping predictive performance as good non-sparse models, as opposed to if very small effects were simply discarded \cite[]{bolli2019software}.
+This sparse model perform equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs equally well as the non-sparse model.
+As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, but does not perform well for half of the real traits analyzed here.
+More investigations need to perform to understand the poor results of LDpred2-auto for some real traits.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We use the remaining 27,475 individuals as test set for evaluating models.
+
+We simulate binary phenotypes with a heritability of $h^2 = 0.4$ (or 0.3)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}. 
+We vary the number of causal variants (300, 3000, 30K, or 300K) to match a range of genetic architectures from low to high polygenicity.
+Liability scores are computed from a model with additive effects only: we compute the liability score of the i-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G_{i,j}} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
+Causal variants are chosen randomly anywhere on the genome. 
+We make two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25-35 Mb).
+Both parts of the $y_i$ are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
+This is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
+Each simulation scenario is repeated 10 times and averages of the Area Under the ROC Curve (AUC) are reported.
 

### 2020-04-14 paper/paper-ldpred2.tex
save

@@
 The use of polygenic scores (PGS) is widespread now.
-There are high hopes for using these scores, which sum up the information of many genetic variants into a single score, into clinical practice. 
-Predictive performance of PGS is inexorably bounded by heritability, the proportion of phenotypic variance that is attributable to genetics, so that their possibility of being used as a diagnostic tool for diseases is limited.
-Nevertheless, it should be possible to use them to stratify the population into groups of different risk levels, to be able to identify individuals at high risk for a given disease [CITATION]. 
+A PGS sums up the information of many genetic variants into a single score.
+There are high hopes for using these scores into clinical practice. 
+Predictive performance of PGS is inexorably bounded by heritability, the proportion of phenotypic variance that is attributable to genetics, therefore using them as a diagnostic tool for diseases is limited.
+Nevertheless, it should be possible to use them to stratify the population into groups of different risk levels, to be able to identify individuals at high risk for a given disease \cite[]{pashayan2015implications,willoughby2019genetic}.
 These people should then be monitored more closely to detect the disease and intervene sooner to maximize chances of remission (e.g.\ in the case of cancer).
 PGS are also extensively used in epidemiology and economics as predictive variables of interest.
-For example, a recently derived PGS for education attainment has been the most predictive variable in social sciences so far [CITATION].
+For example, a recently derived PGS for education attainment has been one of the most predictive variables in behavioural sciences so far \cite[]{allegrini2019genomic}.
 
@@
 It may also be very unstable in long-range LD regions such as the human leukocyte antigen (HLA) region of chromosome 6.
-This issue has lead to the removal of this region from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is a shame since this region of the genome contains several disease-associated variants, particularly with autoimmune diseases and psychiatric disorders [CITATIONS].
+This issue has lead to the removal of this region from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is a shame since this region of the genome contains several disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
 LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as in its model specification.
@@
 We provide a faster and more robust implementation of LDpred in R package bigsnpr \cite[]{prive2017efficient}.
-We also provide two new options: a ``sparse'' option, where LDpred2 truly fit some effects to be zero, therefore providing a sparse vector of effects; and an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the (SNP) heritability $h^2$, therefore for which there is no need for validation data to choose hyper-parameters performing best.
-We show that [DESCRIBE RESULTS].
+We also provide two new options: a ``sparse'' option, where LDpred2 truly fit some effects to be zero, therefore providing a sparse vector of effects; and an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, therefore for which there is no need for validation data to choose hyper-parameters performing best.
+We show that LDpred2 provides higher predictive performance than LDpred1, especially when there are causal variants in long-range LD regions or when the proportion of causal variants is small.
+We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to fit a sparse model without losing prediction accuracy.
 
@@
 Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
-LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation test; 3) LDpred3-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a truly sparse version of LDpred; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameter to optimize.
-As a recall, LDpred1 has two options: LDpred1-inf and LDpred1-grid, where only $p$ is optimized.
+LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation test; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a truly sparse version of LDpred; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to optimize.
+As a recall, LDpred1 has two options: LDpred1-grid where only $p$ is optimized, and LDpred1-inf.
 
 We compare the two versions of LDpred using 6 simulation scenarios to understand in which cases it is important to use the new version.
-We also compare these two versions in 8 binary phenotypes of major interest using published external summary statistics and the UK Biobank data.
+We also compare these two versions of LDpred using 8 binary phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
 
@@
 6 simulation scenarios are used, each repeated 10 times.
-In the first 4 simulations scenarios, a heritability of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
-In these scenarios, infinitesimal models are performing similarly.
-When testing a grid of parameters, LDpred2 performed substantially better than LDpred1 in the cases where the number of causal variations is smaller, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3).
-For example, in simulations with 300 causal variants, AUC of 73.5\% is obtained with LDpred1 while an AUC of 81.9\% is obtained with LDpred2.
+In the first 4 simulation scenarios, a heritability of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
+In these scenarios, infinitesimal models perform similarly. 
+When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is smaller, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3).
+For example, in simulations with 300 causal variants, a mean AUC of 73.5\% is obtained with LDpred1 while a value of 81.9\% is obtained with LDpred2.
 In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
-In HLA scenarios, i.e.\ where causal variants are sampled in a region with variants at high LD, LDpred1 perform poorly.
-In these HLA scenarios, all LDpred2 models perform well, although LDpred2-auto perform slightly worse than other models.
+In HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in high LD, LDpred1 performs poorly.
+In these two HLA scenarios, all LDpred2 models perform well, although LDpred2-auto performs slightly worse than other models.
 
@@
 \caption{Results of the 6 simulation scenarios. 
-Scenarios are named with 4 parts, separated by underscore: 1) ``all'' means that causal variants were randomly sampled anywhere on the genome while ``HLA'' means that they were sampled in the HLA region of chromosome 6 (25-35 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
+Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25-35 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
 Bars present the mean and 95\% CI of $10^4$ non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
@@
 Here we present LDpred2, a new version of LDpred. 
-LDpred is widely used and has the potential to provide polygenic models with good predictive performance \cite{khera2018genome}. 
+LDpred is widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
 Yet, it has some instability issues.
@@
 We demonstrate that LDpred1 has some large issues due to long-range LD regions such as the HLA region.
-Indeed, both LDpred1 and LDpred1-inf perform poorly in the simulations where causal variants are the HLA region.
-In constrast, LDpred2 perform very well. We hypothesize that LDpred1 does not use a window size that is large enough to account for long-range LD such as in the HLA region.
+Indeed, LDpred1 performs poorly in the simulations where causal variants are in the HLA region.
+In constrast, LDpred2 performs very well. We hypothesize that LDpred1 does not use a window size that is large enough to account for long-range LD such as in the HLA region.
 In LDpred2, we use a window size of 3 cM, which is much larger and enable LDpred2 to work well even when causal variants are in long-range LD regions.
-We hypothesize that using a larger LD window explains the 5 real traits where LDpred2 performs much better than LDpred1, especially because LDpred2-inf performs better than LDpred1-inf.
-We strongly discourage against removing these regions as suggested by others as these regions, especially the HLA region, contain lots of variants associated with many traits such as autoimmune diseases and psychiatric disorders, and are therefore very useful for prediction.
+%We hypothesize that using a larger LD window explains the 5 real traits where LDpred2 performs much better than LDpred1, especially because LDpred2-inf performs better than LDpred1-inf.
+We strongly discourage against removing these regions as suggested by others.
+Indeed, these regions, especially the HLA region, contain lots of variants associated with many traits, and are therefore very useful for prediction.
 
 In LDpred2, we also test more values for $p$ (17 instead of 7 by default) and for $h^2$ (3 instead of 1).
-When testing the grid of hyper-parameters of $p$ and $h2$, we also allow for testing an option to enable sparse models in LDpred2 (see after).
-Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1, we use 5 times as many iterations, use a larger window size for computing correlation, yet it is LDpred2 is still faster than LDpred1.
-LDpred2 is faster than LDpred1 because we provide a very efficient parallel implementation in C++.
-The parallelization is performed on the grid of parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see methods).
-As an example, it takes less than two hours to run LDpred2-inf, LDpred2-grid (108 hyper-parameters values) and LDpred2-auto for a chromosome with 100K variants and using 8 cores, and only 15 minutes for a chromosome with 20K variants.
+When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see after).
+Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1, we use 5 times as many iterations in the Gibbs sampler, we use a larger window size for computing correlation, yet LDpred2 is still faster than LDpred1.
+LDpred2 is faster than LDpred1 because we provide an efficient parallel implementation in C++.
+The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
+As an example, it takes less than two hours to run LDpred2-inf, LDpred2-grid (108 hyper-parameter values) and LDpred2-auto for a chromosome with 100K variants and using 8 cores.
 Pre-computing the LD matrix for a chromosome with 100K variants takes 11 minutes using 16 cores.
@@
 LDpred2 also comes with two new models.
-When the sparse option is enabled, it provide models that are truly sparse, as compared to LDpred1 which outputs very small non-zero effect sizes \cite{cecile2019polygenic}.
-It can provide sparse effect sizes as much as 98\% for small $p$, while keeping predictive performance as good non-sparse models, as opposed to if very small effects were simply discarded \cite[]{bolli2019software}.
-This sparse model perform equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs equally well as the non-sparse model.
+When the sparse option is enabled, it provides models that are truly sparse, as compared to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
+It can provide sparsity in effect sizes as much as 98\% for small $p$, while keeping predictive performance as good as non-sparse models, as opposed to if very small effects were simply discarded \cite[]{bolli2019software}.
+This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs equally well as the non-sparse model.
 As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, but does not perform well for half of the real traits analyzed here.
-More investigations need to perform to understand the poor results of LDpred2-auto for some real traits.
+More investigations need to be performed to understand the poor results of LDpred2-auto for these particular real traits.
+
+[TALK ABOUT QUALITY CONTROL ON AF?]
+
+[TALK ABOUT DIFFERENT SETS OF VARIANTS?]
 
@@
 
-We use the UK Biobank data for both real data analyses and simulations.
+We use the UK Biobank data for both real data analyses and simulations \cite[]{bycroft2018uk}.
 We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled).
-We restrict variants to HapMap3 variants.
+We restrict variants to HapMap3 variants (considered to be well imputed).
 This results in 337,475 individuals and 1,194,574 variants.
-We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and computing correlations between variants (LD matrix). 
+We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and for computing correlations between variants (LD matrix). 
 We use 300,000 other individuals for running logistic GWAS to create summary statistics.
@@
 Both parts of the $y_i$ are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
-This is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
+Such simulation of phenotypes based on real genotypes is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
 Each simulation scenario is repeated 10 times and averages of the Area Under the ROC Curve (AUC) are reported.
@@
 We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We defined phenotypes as done in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section).
+We defined phenotypes as in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section).
 
@@
 In this section, we explain how we can obtain joint effects from summary statistics (marginal effects) and a correlation matrix.
-Let us denote by $S$ the diagonal matrix with standard deviations of the $p$ variants, $C_n = I_n - 1 1^T / n$ the centering matrix, $G$ the genotype matrix of $n$ individuals and $p$ variants, and $y$ the phenotype vector for $n$ individuals.
+Let us denote by $S$ the diagonal matrix with standard deviations of the $m$ variants, $C_n = I_n - 1 1^T / n$ the centering matrix, $G$ the genotype matrix of $n$ individuals and $m$ variants, and $y$ the phenotype vector for $n$ individuals.
 Then, the marginal effects (assuming no other covariates than the intercept) are given by 
@@
 where $p$ is the proportion of causal variants, $M$ the number of variants and $h^2$ the (SNP) heritability.
-\cite{vilhjalmsson2015modeling} estimates $h^2$ using constrained LD score regression (intercept fixed to 1) and recommend testing a grid of hyper-parameter values for $p$ (e.g.\ 1, 0.3, 0.1, 0.03, 0.01, etc.).
+\cite{vilhjalmsson2015modeling} estimates $h^2$ using constrained LD score regression (intercept fixed to 1) and recommend testing a grid of hyper-parameter values for $p$ (1, 0.3, 0.1, 0.03, 0.01, 0.003 and 0.001).
 
@@
 
-The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h2$ using constrained LD score regression \cite[]{bulik2015ld}.
+The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h^2$ using constrained LD score regression \cite[]{bulik2015ld}.
 This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
@@
 
-The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e. that some resulting effects are exactly 0.
+The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ that some resulting effects are exactly 0.
 When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j = 0$.
@@
 The recommended value for this window (in number of variants) is to use the total number of variants divided by 3000, which corresponds to a window radius of around 2 Mb \cite[]{vilhjalmsson2015modeling}.
-After investigating, we come to the conclusion that this window size is not large enough. Indeed, the human leukocyte antigen (HLA) region of chromosome 6 is 8 Mb long \cite[]{price2008long}. 
+We have come to the conclusion that this window size is not large enough. Indeed, the human leukocyte antigen (HLA) region of chromosome 6 is 8 Mb long \cite[]{price2008long}. 
 Using a window of 8Mb would be computationally and memory inefficient.
@@
 Therefore, genetic distances enable to capture the same LD using a globally smaller window.
-We provide function \texttt{snp\_asGeneticPos()} in package bignspr to easily interpolate physical position to genetic positions.
+We provide function \texttt{snp\_asGeneticPos} in package bignspr to easily interpolate physical position to genetic positions.
 We recommend to use genetic positions and to use a size parameter of 3 cM when computing the correlation between variants for LDpred2. 
-Note that, in the code, we use \texttt{size = 3 / 1000} as parameter \texttt{size} is internally multiplied by 1000 in functions of package bigsnpr.
+Note that, in the code, we use \texttt{size = 3 / 1000} since parameter \texttt{size} is internally multiplied by 1000 in functions of package bigsnpr.
+To prevent storing very small correlations, we also set to 0 all correlations with a p-value larger than \texttt{alpha = 0.9} (when testing that the correlation is different from 0).
 
@@
 Second, as noted in \cite{prive2019making}, it may be beneficial to assume that architecture of traits may be different for different chromosomes.
-For example, autoimmune diseases clearly show larger effects on chromosome 6. compared to other chromosomes.
+For example, autoimmune diseases clearly show larger effects on chromosome 6 compared to other chromosomes.
 Assuming the same model for genetic effects genome-wide could results in severe model misspecification, which would lead to suboptimal predictive performance.
-Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combining the results.
+Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results.
 
@@
 
+[TODO] 
 All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}.

### 2020-04-18 paper/paper-ldpred2.tex
Julyan's comments

@@
 This issue has lead to the removal of this region from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is a shame since this region of the genome contains several disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
-LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as in its model specification.
+LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as expected by its model specification.
 
-Here, we present LDpred2, a new version of LDpred, that aims to cope with the issues presented before.
+Here, we present LDpred2, a new version of LDpred, that aims to cope with these issues.
 We provide a faster and more robust implementation of LDpred in R package bigsnpr \cite[]{prive2017efficient}.
-We also provide two new options: a ``sparse'' option, where LDpred2 truly fit some effects to be zero, therefore providing a sparse vector of effects; and an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, therefore for which there is no need for validation data to choose hyper-parameters performing best.
-We show that LDpred2 provides higher predictive performance than LDpred1, especially when there are causal variants in long-range LD regions or when the proportion of causal variants is small.
+We also provide two new options in LDpred2.
+First, we provide a ``sparse'' option, where LDpred2 truly fits some effects to zero, therefore providing a sparse vector of effects.
+Second, we also provide an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, therefore for which there is no need for validation data to choose hyper-parameters performing best.
+We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions or when the proportion of causal variants is small.
 We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to fit a sparse model without losing prediction accuracy.
@@
 LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation test; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a truly sparse version of LDpred; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to optimize.
-As a recall, LDpred1 has two options: LDpred1-grid where only $p$ is optimized, and LDpred1-inf.
+As a recall, LDpred v1 has two options: LDpred1-grid where only $p$ is optimized, and LDpred1-inf.
 
-We compare the two versions of LDpred using 6 simulation scenarios to understand in which cases it is important to use the new version.
-We also compare these two versions of LDpred using 8 binary phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
+We compare the two versions of LDpred using six simulation scenarios to understand in which cases it is important to use the new version.
+We also compare these two versions of LDpred using eight binary phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
+Area Under the ROC Curve (AUC) values are reported.
 
@@
 Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
-6 simulation scenarios are used, each repeated 10 times.
-In the first 4 simulation scenarios, a heritability of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
+six simulation scenarios are used, each repeated 10 times.
+In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
 In these scenarios, infinitesimal models perform similarly. 
@@
 \centerline{\includegraphics[width=0.95\textwidth]{AUC-simu}}
-\caption{Results of the 6 simulation scenarios. 
+\caption{Results of the six simulation scenarios. 
 Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25-35 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
@@
 Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr).
-8 binary phenotypes are used, summarized in table \ref{tab:sumstats}.
+eight binary phenotypes are used, summarized in table \ref{tab:sumstats}.
 For T1D, T2D, BRCA, PRCA and MDD, all LDpred2 models perform much better than LDpred1.
@@
 We show that LDpred2 is much more stable and provides higher prediction than LDpred1.
-We demonstrate that LDpred1 has some large issues due to long-range LD regions such as the HLA region.
+We demonstrate that LDpred1 has defects, particularly when handling long-range LD regions such as the HLA region.
 Indeed, LDpred1 performs poorly in the simulations where causal variants are in the HLA region.
-In constrast, LDpred2 performs very well. We hypothesize that LDpred1 does not use a window size that is large enough to account for long-range LD such as in the HLA region.
-In LDpred2, we use a window size of 3 cM, which is much larger and enable LDpred2 to work well even when causal variants are in long-range LD regions.
+In contrast, LDpred2 performs very well. We hypothesize that LDpred1 does not use a window size that is large enough to account for long-range LD such as in the HLA region.
+In LDpred2, we use a window size of 3 cM, which is much larger than the default value used in LDpred1 and which enables LDpred2 to work well even when causal variants are in long-range LD regions.
 %We hypothesize that using a larger LD window explains the 5 real traits where LDpred2 performs much better than LDpred1, especially because LDpred2-inf performs better than LDpred1-inf.
-We strongly discourage against removing these regions as suggested by others.
+We strongly discourage against removing these regions as sometimes suggested in the literature. 
 Indeed, these regions, especially the HLA region, contain lots of variants associated with many traits, and are therefore very useful for prediction.
@@
 When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see after).
-Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1, we use 5 times as many iterations in the Gibbs sampler, we use a larger window size for computing correlation, yet LDpred2 is still faster than LDpred1.
-LDpred2 is faster than LDpred1 because we provide an efficient parallel implementation in C++.
+Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1, we use 5 times as many iterations in the Gibbs sampler, we use a larger window size for computing correlation, yet LDpred2 is still as fast as LDpred1.
+LDpred2 is fast because we provide an efficient parallel implementation in C++.
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
@@
 LDpred2 also comes with two new models.
-When the sparse option is enabled, it provides models that are truly sparse, as compared to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
+When the sparse option is enabled, it provides models that truly encourage sparsity, as compared to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
 It can provide sparsity in effect sizes as much as 98\% for small $p$, while keeping predictive performance as good as non-sparse models, as opposed to if very small effects were simply discarded \cite[]{bolli2019software}.
@@
 We simulate binary phenotypes with a heritability of $h^2 = 0.4$ (or 0.3)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}. 
-We vary the number of causal variants (300, 3000, 30K, or 300K) to match a range of genetic architectures from low to high polygenicity.
-Liability scores are computed from a model with additive effects only: we compute the liability score of the i-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G_{i,j}} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
+We vary the number of causal variants (300, 3000, 30,000, or 300,000) to match a range of genetic architectures from low to high polygenicity.
+Liability scores are computed from a model with additive effects only: we compute the liability score of the $i$-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G_{i,j}} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon_i$ follows a Gaussian distribution $N(0, 1 - h^2)$.
 Causal variants are chosen randomly anywhere on the genome. 
 We make two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25-35 Mb).
-Both parts of the $y_i$ are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
+Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
 Such simulation of phenotypes based on real genotypes is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
@@
 Then, the marginal effects (assuming no other covariates than the intercept) are given by 
-\[\hat{\beta}_{marg} = S^{-2} G^T C_n y ~,\]
-while the joint effects are obtained by solving
-\[\begin{bmatrix} \hat{\alpha} \\ \hat{\beta}_{joint} \end{bmatrix} = \left(\begin{bmatrix} 1 & G \end{bmatrix}^T \begin{bmatrix} 1 & G \end{bmatrix}\right)^{-1} \begin{bmatrix} 1 & G \end{bmatrix}^T y ~.\]
+\[\hat{\beta}_{\text{marg}} = S^{-2} G^T C_n y ~,\]
+while the intercept $\hat{\alpha}$ and the joint effects are obtained by solving
+\[\begin{bmatrix} \hat{\alpha} \\ \hat{\beta}_{\text{joint}} \end{bmatrix} = \left(\begin{bmatrix} 1 & G \end{bmatrix}^T \begin{bmatrix} 1 & G \end{bmatrix}\right)^{-1} \begin{bmatrix} 1 & G \end{bmatrix}^T y ~.\]
 Using the Woodburry formula, we get 
-\[\hat{\beta}_{joint} = (G^T C_n G)^{-1} G^T C_n y ~.\]
+\[\hat{\beta}_{\text{joint}} = (G^T C_n G)^{-1} G^T C_n y ~.\]
 We further note that the correlation matrix of $G$ is \(K =  S^{-1} G^T C_n G  S^{-1}\).
-Then we get $\hat{\beta}_{joint} = S^{-1} K^{-1} S \hat{\beta}_{marg}$.
+Then we get $\hat{\beta}_{\text{joint}} = S^{-1} K^{-1} S \hat{\beta}_{\text{marg}}$.
 
@@
 
-Let us go back to the formula we derived before: $\hat{\beta}_{joint} = S^{-1} K^{-1} S \hat{\beta}_{marg}$.
+Let us go back to the formula we derived before: $\hat{\beta}_{\text{joint}} = S^{-1} K^{-1} S \hat{\beta}_{\text{marg}}$.
 As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$.
@@
 Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{sd}(\hat{\beta}) ~ \sqrt{N}\right)$.
-Note that LDpred (v1) and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.
+Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.
 
@@
 \begin{array}{ll}
-\mathcal N\left(0, \dfrac{h^2}{M p}\right) & \mbox{with probably $p$,} \\
+\mathcal N\left(0, \dfrac{h^2}{M p}\right) & \mbox{with probability $p$,} \\
 0 & \mbox{otherwise,}
@@
 \begin{array}{ll}
-\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{N h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{N h^2}}\dfrac{1}{N}\right) & \mbox{with probably $\bar{p}_j$,} \\
+\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{N h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{N h^2}}\dfrac{1}{N}\right) & \mbox{with probability $\bar{p}_j$,} \\
 0 & \mbox{otherwise.}
@@
 This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
-To estimate $p$, we count the number of causal variants (i.e.\ $S = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:beta_random}). 
-We can assume that $S \sim B(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the conjugate prior $p \sim \text{Beta}(1 + S, 1 + M - S)$.
+To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:beta_random}). 
+We can assume that $m_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + m_c, 1 + M - m_c)$.
 Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$. Instead, we estimate $h^2 = \beta^T K \beta$, where $K$ is the correlation matrix.
@@
 For example, autoimmune diseases clearly show larger effects on chromosome 6 compared to other chromosomes.
-Assuming the same model for genetic effects genome-wide could results in severe model misspecification, which would lead to suboptimal predictive performance.
+Assuming the same model for genetic effects genome-wide could result in severe model misspecification, which would lead to suboptimal predictive performance.
 Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results.

### 2020-04-19 paper/paper-ldpred2.tex
save

@@
 Let us denote by $S$ the diagonal matrix with standard deviations of the $m$ variants, $C_n = I_n - 1 1^T / n$ the centering matrix, $G$ the genotype matrix of $n$ individuals and $m$ variants, and $y$ the phenotype vector for $n$ individuals.
-Then, the marginal effects (assuming no other covariates than the intercept) are given by 
-\[\hat{\beta}_{\text{marg}} = S^{-2} G^T C_n y ~,\]
-while the intercept $\hat{\alpha}$ and the joint effects are obtained by solving
-\[\begin{bmatrix} \hat{\alpha} \\ \hat{\beta}_{\text{joint}} \end{bmatrix} = \left(\begin{bmatrix} 1 & G \end{bmatrix}^T \begin{bmatrix} 1 & G \end{bmatrix}\right)^{-1} \begin{bmatrix} 1 & G \end{bmatrix}^T y ~.\]
+
+When solving a joint model with all variants and an intercept ${\alpha}$, the joint effects ${\gamma}_{\text{joint}}$ are obtained by solving
+\[\begin{bmatrix} \hat{\alpha} \\ \hat{\gamma}_{\text{joint}} \end{bmatrix} = \left(\begin{bmatrix} 1 & G \end{bmatrix}^T \begin{bmatrix} 1 & G \end{bmatrix}\right)^{-1} \begin{bmatrix} 1 & G \end{bmatrix}^T y ~.\]
 Using the Woodburry formula, we get 
-\[\hat{\beta}_{\text{joint}} = (G^T C_n G)^{-1} G^T C_n y ~.\]
-We further note that the correlation matrix of $G$ is \(K =  S^{-1} G^T C_n G  S^{-1}\).
-Then we get $\hat{\beta}_{\text{joint}} = S^{-1} K^{-1} S \hat{\beta}_{\text{marg}}$.
+\[\hat{\gamma}_{\text{joint}} = (G^T C_n G)^{-1} G^T C_n y ~.\]
+When fitting each variant separately in GWAS, the marginal effects (assuming no covariate) simplify to 
+\[\hat{\gamma}_{\text{marg}} = \dfrac{1}{n-1} S^{-2} G^T C_n y ~.\] 
+We further note that the correlation matrix of $G$ is \(K =  \dfrac{1}{n-1} S^{-1} G^T C_n G  S^{-1}\).
+Then we get \[\hat{\gamma}_{\text{joint}} = S^{-1} K^{-1} S \hat{\gamma}_{\text{marg}} ~.\]
 
-For one marginal effect $\hat{\beta}$, let us denote by $y^*$ and $g^*$ the corresponding phenotype and genotype (residualized from $K$ covariates, e.g.\ centering them).
+For one marginal effect $\hat{\gamma}$, let us denote by $y^*$ and $g^*$ the corresponding phenotype and genotype (residualized from $K$ covariates, e.g.\ centering them).
 Then,
-$\text{var}(\hat{\beta}) = \dfrac{(y^* - \hat{\beta} g^*)^T (y^* - \hat{\beta} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
-Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{sd}(\hat{\beta}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\beta}\right) \approx \dfrac{\hat{\beta}}{\text{sd}(\hat{\beta})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.
+$\text{var}(\hat{\gamma}) = \dfrac{(y^* - \hat{\gamma} g^*)^T (y^* - \hat{\gamma} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
+Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{sd}(\hat{\gamma}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\gamma}\right) \approx \dfrac{\hat{\gamma}}{\text{sd}(\hat{\gamma})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.
 
-Let us go back to the formula we derived before: $\hat{\beta}_{\text{joint}} = S^{-1} K^{-1} S \hat{\beta}_{\text{marg}}$.
+Let us go back to the formula we derived before: $\hat{\gamma}_{\text{joint}} = S^{-1} K^{-1} S \hat{\gamma}_{\text{marg}}$. 
+Note that $\gamma$ are the effects on the allele scale while we denote by $\beta = S \gamma$ the effects of the scaled genotypes.
 As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$.
-Then it justifies the use of Z-scores ($\hat{\beta} / \text{sd}(\hat{\beta})$) divided by $\sqrt{N}$ as input for LDpred.
-Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{sd}(\hat{\beta}) ~ \sqrt{N}\right)$.
+Then it justifies the use of Z-scores ($\hat{\gamma} ~/~ \text{sd}(\hat{\gamma})$) divided by $\sqrt{N}$ as input for LDpred.
+Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{sd}(\hat{\gamma}) ~ \sqrt{N}\right)$.
 Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.
@@
 \begin{equation}\label{eq:model}
-\beta_j \sim \left\{
+\beta_j = S \gamma_j \sim \left\{
 \begin{array}{ll}

### 2020-04-22 paper/paper-ldpred2.tex
Bjarni's suggestions

@@
 \DeclareMathOperator*{\argmin}{\arg\!\min}
+\usepackage{algorithm} 
+\usepackage{algpseudocode}
 
@@
 \title{LDpred2: better, faster, stronger}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,3,}*}$}
 
@@
 \noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
+\noindent$^{\text{\sf 3}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 
-The use of polygenic scores (PGS) is widespread now.
-A PGS sums up the information of many genetic variants into a single score.
-There are high hopes for using these scores into clinical practice. 
-Predictive performance of PGS is inexorably bounded by heritability, the proportion of phenotypic variance that is attributable to genetics, therefore using them as a diagnostic tool for diseases is limited.
-Nevertheless, it should be possible to use them to stratify the population into groups of different risk levels, to be able to identify individuals at high risk for a given disease \cite[]{pashayan2015implications,willoughby2019genetic}.
-These people should then be monitored more closely to detect the disease and intervene sooner to maximize chances of remission (e.g.\ in the case of cancer).
-PGS are also extensively used in epidemiology and economics as predictive variables of interest.
+In recent years the use of polygenic scores (PGS) has become widespread.
+A PGS aggregates (risk) effects across many genetic variants into a single predictive score.
+These scores have proven useful for studying the genetic architecture and relationships between diseases and traits \cite[]{purcell2009common,kong2018nature}.
+Moreover, there are high hopes for using these scores in clinical practice to improve disease risk estimates and prognostic accuracies.
+The heritability, i.e.\ the proportion of phenotypic variance that is attributable to genetics, determines an upper limit on the predictive performance of PGS and thus their value as a diagnostic tool.
+Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
+PGS are also extensively used in epidemiology and economics as predictive variables of interest [TODO BJARNI CITATIONS]. 
 For example, a recently derived PGS for education attainment has been one of the most predictive variables in behavioural sciences so far \cite[]{allegrini2019genomic}.
@@
 It assumes there is a proportion $p$ of variants that are causal.
-However, LDpred currently has several limitations that may result in limited predictive performance.
+However, LDpred has several limitations that may result in limited predictive performance.
 The non-infinitesimal version of LDpred, a Gibbs sampler, is particularly sensitive to model misspecification when applied to summary statistics with large sample sizes.
-It may also be very unstable in long-range LD regions such as the human leukocyte antigen (HLA) region of chromosome 6.
-This issue has lead to the removal of this region from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is a shame since this region of the genome contains several disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
-LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as expected by its model specification.
+It is also unstable in long-range LD regions such as the human leukocyte antigen (HLA) region of chromosome 6. 
+This issue has lead to the removal of such regions from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is unfortunate since this region of the genome contains many known disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
+%LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as expected by its model specification.
 
-Here, we present LDpred2, a new version of LDpred, that aims to cope with these issues.
-We provide a faster and more robust implementation of LDpred in R package bigsnpr \cite[]{prive2017efficient}.
+Here, we present LDpred2, a new version of LDpred that addresses these issues while markedly improving its computational efficiency.
+We provide this faster and more robust implementation of LDpred in R package bigsnpr \cite[]{prive2017efficient}.
 We also provide two new options in LDpred2.
 First, we provide a ``sparse'' option, where LDpred2 truly fits some effects to zero, therefore providing a sparse vector of effects.
-Second, we also provide an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, therefore for which there is no need for validation data to choose hyper-parameters performing best.
-We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions or when the proportion of causal variants is small.
-We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to fit a sparse model without losing prediction accuracy.
+Second, we also provide an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, and therefore does not require validation data to tune hyper-parameters.
+We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions, or when the proportion of causal variants is small.
+We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing prediction accuracy.
 
@@
 LDpred is widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
-Yet, it has some instability issues.
-We propose a new version to solve these instability issues and try to improve other points as well.
+Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
+We propose a new version of LDpred to solve these instability issues and improve its computational efficiency.
 
@@
 
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.85\textwidth]{QC-CAD}}
+	\caption{Results of the real data applications using published external summary statistics.
+		Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
+		%See corresponding values in table S2.
+	}
+	\label{fig:QC-CAD}
+\end{figure}
 [TALK ABOUT QUALITY CONTROL ON AF?]
@@
 \end{equation}
+Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $N \tilde{\beta}_j^2$ is large.
 
@@
 \end{equation}
+
+%\begin{minipage}{.7\textwidth}
+\begin{algorithm}[H]
+\caption{LDpred} 
+\begin{algorithmic}
+	\For {$i=1,\ldots,N_{iter}$}
+	\For {$j=1,\ldots,M$} \Comment{put some comments here}
+	\State Run ... \Comment{put some comments here}
+	\State Compute... \Comment{put some comments here}
+	\EndFor
+	\State Optimize ...
+	\State $\theta_{old}\leftarrow\theta$
+	\EndFor
+\end{algorithmic} 
+\end{algorithm}
+%\end{minipage}
 

### 2020-04-22 paper/paper-ldpred2.tex
Add algo

@@
 
-To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler [REALLY A GS?]. They compute the probability that variant $j$ is causal as 
+To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.
+
+First, they estimate the residualized marginal effect for variant $j$ as [SHOULD VERIFY]
+\begin{equation}\label{eq:beta_res}
+\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{K}_{\text{-}j, j}
+\end{equation}
+where $\boldsymbol{K}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix and $\boldsymbol{\hat{\beta}}$ are the marginal effect sizes.
+Then, they compute the probability that variant $j$ is causal as 
 \[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
@@
 \begin{equation}
-\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.
+\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.\label{eq:postp}
 \end{equation}
@@
 Then, $\beta_j$ is sampled according to
-\begin{equation}\label{eg:beta_random}
+\begin{equation}\label{eg:random_beta}
 \beta_j ~|~ \tilde{\beta}_j \sim \left\{
@@
 
+Therefore, the posterior mean of $\beta_j$ is given by 
+\begin{equation}\label{eg:post_beta}
+\omega_j = \dfrac{\bar{p}_j \tilde{\beta}_j}{1 + \dfrac{M p}{N h^2}} ~.
+\end{equation}
+
 %\begin{minipage}{.7\textwidth}
 \begin{algorithm}[H]
-\caption{LDpred} 
-\begin{algorithmic}
-	\For {$i=1,\ldots,N_{iter}$}
-	\For {$j=1,\ldots,M$} \Comment{put some comments here}
-	\State Run ... \Comment{put some comments here}
-	\State Compute... \Comment{put some comments here}
+\caption{LDpred, with hyper-parameters $p$ and $h^2$}\label{algo:ldpred}
+\begin{algorithmic}[1]
+	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
+	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$
+	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
+	\For {each variant $j$} \Comment{All variants}
+	\State Compute $\tilde{\beta}_j$ according to \eqref{eq:beta_res}
+	\State Compute $\bar{p}_j$ according to \eqref{eq:postp}
+	\State Sample $\beta_j$ according to \eqref{eg:random_beta}
+	\State Compute $\omega_j$ according to \eqref{eg:post_beta}
 	\EndFor
-	\State Optimize ...
-	\State $\theta_{old}\leftarrow\theta$
+	\State Compute $\boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
+	\If {$k > N_\text{burn-in}$} 
+	\State Increment $\boldsymbol{\Omega}$ with $\boldsymbol{\omega}$ 
+	\EndIf
 	\EndFor
+	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$
+	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
 \end{algorithmic} 
@@
 
-[TO COMPLETE + CLEARER -> SHOULD SHOW THAT IT IS ITERATIVE]
+[CHANGE NOTATION OF SUMSTATS / SD -> SE?]
 
@@
 This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
-To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:beta_random}). 
+To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
 We can assume that $m_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + m_c, 1 + M - m_c)$.
 Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$. Instead, we estimate $h^2 = \beta^T K \beta$, where $K$ is the correlation matrix.
+These parameters $p$ and $h^2$ are updated after the inner loop in algorithm \ref{algo:ldpred}, then these new values are used in the next iteration of the outer loop.
 
 The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ that some resulting effects are exactly 0.
-When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j = 0$.
+When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j$ and $\omega_j$ to 0 (lines 6-8 of algorithm \ref{algo:ldpred}).
 Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the $p$ and $h^2$ estimates from LDpred2-auto.

### 2020-04-24 paper/paper-ldpred2.tex
save

@@
 \end{figure}
+
 [TALK ABOUT QUALITY CONTROL ON AF?]
@@
 [TALK ABOUT DIFFERENT SETS OF VARIANTS?]
+
+[NEED TO UNDERSTAND WHY AUTO DOES NOT WORK FOR SOME TRAITS -> CAD BECAUSE DIFFERENT N // T1D PROBABLY BECAUSE OF SMALL SAMPLE SIZE // ASTHMA? AND PRCA?]
+
+[TALK ABOUT MODEL FOR BINARY DATA]
 
@@
 $\text{var}(\hat{\gamma}) = \dfrac{(y^* - \hat{\gamma} g^*)^T (y^* - \hat{\gamma} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
-Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{sd}(\hat{\gamma}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\gamma}\right) \approx \dfrac{\hat{\gamma}}{\text{sd}(\hat{\gamma})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.
+Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{se}(\hat{\gamma}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\gamma}\right) \approx \dfrac{\hat{\gamma}}{\text{se}(\hat{\gamma})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.
 
@@
 As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$.
-Then it justifies the use of Z-scores ($\hat{\gamma} ~/~ \text{sd}(\hat{\gamma})$) divided by $\sqrt{N}$ as input for LDpred.
-Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{sd}(\hat{\gamma}) ~ \sqrt{N}\right)$.
+Then it justifies the use of Z-scores ($\hat{\gamma} ~/~ \text{se}(\hat{\gamma})$) divided by $\sqrt{N}$ as input for LDpred.
+Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{se}(\hat{\gamma}) ~ \sqrt{N}\right)$.
 Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.
@@
 To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.
-
 First, they estimate the residualized marginal effect for variant $j$ as [SHOULD VERIFY]
@@
 \begin{algorithmic}[1]
-	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
+	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
 	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$
@@
 	\EndFor
-	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$
-	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{sd}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
+	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$ \Comment{Average of all $\boldsymbol{\omega}$ after burn-in}
+	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
 \end{algorithmic} 
@@
 
-[CHANGE NOTATION OF SUMSTATS / SD -> SE?]
+[CHANGE NOTATION OF SUMSTATS?]
 
@@
 Second, as noted in \cite{prive2019making}, it may be beneficial to assume that architecture of traits may be different for different chromosomes.
-For example, autoimmune diseases clearly show larger effects on chromosome 6 compared to other chromosomes.
+For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
 Assuming the same model for genetic effects genome-wide could result in severe model misspecification, which would lead to suboptimal predictive performance.

### 2020-04-26 paper/paper-ldpred2.tex
save

@@
 \abstract{	
-
+In recent years the use of polygenic scores has become widespread.
+LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a Linkage Disequilibrium (LD) matrix only.
+However, LDpred has several limitations that may result in limited predictive performance.
+Here, we present LDpred2, a new version of LDpred that addresses these issues therefore providing higher predictive performance, while markedly improving its computational efficiency.
+We provide this faster and more robust implementation of LDpred in R package bigsnpr, as well as two new options in LDpred2: ``sparse'' and ``auto''.
 }
@@
 It is also unstable in long-range LD regions such as the human leukocyte antigen (HLA) region of chromosome 6. 
-This issue has lead to the removal of such regions from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is unfortunate since this region of the genome contains many known disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
+This issue has led to the removal of such regions from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is unfortunate since this region of the genome contains many known disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
 %LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as expected by its model specification.
@@
 Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
-LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation test; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a truly sparse version of LDpred; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to optimize.
+LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to tune.
 As a recall, LDpred v1 has two options: LDpred1-grid where only $p$ is optimized, and LDpred1-inf.
 
-We compare the two versions of LDpred using six simulation scenarios to understand in which cases it is important to use the new version.
+We compare the two versions of LDpred using six simulation scenarios to understand in which cases it is important to use the new version of LDpred.
 We also compare these two versions of LDpred using eight binary phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
@@
 Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
-six simulation scenarios are used, each repeated 10 times.
+Six simulation scenarios are used, each repeated 10 times.
 In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome. 
@@
 Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr).
-eight binary phenotypes are used, summarized in table \ref{tab:sumstats}.
-For T1D, T2D, BRCA, PRCA and MDD, all LDpred2 models perform much better than LDpred1.
-For example, for BRCA, AUC improves from 58.5\% with LDpred1 to 65\% with LDpred2, and from 55\% to 75\% for T1D.
-For Asthma, CAD and RA, predictive performances of LDpred1 and LDpred2 are similar.
+Eight binary phenotypes are used, summarized in table \ref{tab:sumstats}.
+For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1.
+These improvements are also significant for MDD and CAD, albeit to a lesser extend.
+For example, for BRCA, AUC improves from 58.5\% with LDpred1 to 64.4\% with LDpred2, and from 54.9\% to 74.3\% for T1D.
+For Asthma and RA, predictive performances of LDpred1 and LDpred2 are similar.
 
-As in simulations, the sparse version of LDpred2 performs equally well as the original version, with the exception of PRCA, for which it performs poorly.
-For CAD, PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly.
+As in simulations, the sparse version of LDpred2 performs equally well as the original version.
+For PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly.
 
@@
 When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see after).
-Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1, we use 5 times as many iterations in the Gibbs sampler, we use a larger window size for computing correlation, yet LDpred2 is still as fast as LDpred1.
+Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1.
+We also use 5 times as many iterations in the Gibbs sampler and a larger window size for computing correlations between variants, yet LDpred2 is still as fast as LDpred1.
 LDpred2 is fast because we provide an efficient parallel implementation in C++.
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-As an example, it takes less than two hours to run LDpred2-inf, LDpred2-grid (108 hyper-parameter values) and LDpred2-auto for a chromosome with 100K variants and using 8 cores.
+As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (108 hyper-parameter values parallelized over 8 cores) and LDpred2-auto for a chromosome with 100K variants.
 Pre-computing the LD matrix for a chromosome with 100K variants takes 11 minutes using 16 cores.
@@
 It can provide sparsity in effect sizes as much as 98\% for small $p$, while keeping predictive performance as good as non-sparse models, as opposed to if very small effects were simply discarded \cite[]{bolli2019software}.
-This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs equally well as the non-sparse model.
-As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, but does not perform well for half of the real traits analyzed here.
-More investigations need to be performed to understand the poor results of LDpred2-auto for these particular real traits.
+This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs better or equally well as the non-sparse model.
+As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, but does not perform well for some of the real traits analyzed here.
+This is expected for T1D because it is mainly composed of large effects in the HLA region (see similar simulation results) and because summary statistics have a small sample size (5913 cases and 8828 controls only).
+Yet, we do not know yet why LDpred2-auto performs poorly specifically for PRCA.
+More investigations need to be performed to understand these poor results, and to find if this issue can be fixed.
+Solutions that come to mind include for example running multiple chains with different initial values of $p$ and keeping the one with the best convergence properties.
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.85\textwidth]{QC-CAD}}
-	\caption{Results of the real data applications using published external summary statistics.
-		Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
-		%See corresponding values in table S2.
-	}
-	\label{fig:QC-CAD}
-\end{figure}
+%[TALK ABOUT QUALITY CONTROL ON AF?]
 
-[TALK ABOUT QUALITY CONTROL ON AF?]
+%[TALK ABOUT DIFFERENT SETS OF VARIANTS?]
 
-[TALK ABOUT DIFFERENT SETS OF VARIANTS?]
-
-[NEED TO UNDERSTAND WHY AUTO DOES NOT WORK FOR SOME TRAITS -> CAD BECAUSE DIFFERENT N // T1D PROBABLY BECAUSE OF SMALL SAMPLE SIZE // ASTHMA? AND PRCA?]
-
-[TALK ABOUT MODEL FOR BINARY DATA]
+%[TALK ABOUT MODEL FOR BINARY DATA]
 
@@
 We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled).
-We restrict variants to HapMap3 variants (considered to be well imputed).
+We restrict variants to HapMap3 variants.
 This results in 337,475 individuals and 1,194,574 variants.
@@
 \begin{algorithm}[H]
-\caption{LDpred, with hyper-parameters $p$ and $h^2$}\label{algo:ldpred}
+\caption{LDpred, with hyper-parameters $p$ and $h^2$, and summary statistics $\boldsymbol{\hat{\gamma}}$, $\boldsymbol{\text{se}(\hat{\gamma})}$ and $\boldsymbol{N}$}\label{algo:ldpred}
 \begin{algorithmic}[1]
@@
 
-[CHANGE NOTATION OF SUMSTATS?]
+%[CHANGE NOTATION OF SUMSTATS?]
 
@@
 The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h^2$ using constrained LD score regression \cite[]{bulik2015ld}.
-This makes LDpred2-auto a method free of hyper-parameters which can therefore directly be applied to data without the need of a validation dataset to choose best-performing hyper-parameters.
+This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly to data without the need of a validation dataset to choose best-performing hyper-parameters.
 To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
 We can assume that $m_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + m_c, 1 + M - m_c)$.
-Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$. Instead, we estimate $h^2 = \beta^T K \beta$, where $K$ is the correlation matrix.
+Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$.
+Instead, we estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$, where $\boldsymbol{K}$ is the correlation matrix.
 These parameters $p$ and $h^2$ are updated after the inner loop in algorithm \ref{algo:ldpred}, then these new values are used in the next iteration of the outer loop.
 
-The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ that some resulting effects are exactly 0.
+The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ some resulting effects are exactly 0.
 When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j$ and $\omega_j$ to 0 (lines 6-8 of algorithm \ref{algo:ldpred}).
-Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the $p$ and $h^2$ estimates from LDpred2-auto.
+Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the estimates of $p$ and $h^2$ from LDpred2-auto.
 
@@
 Therefore, genetic distances enable to capture the same LD using a globally smaller window.
-We provide function \texttt{snp\_asGeneticPos} in package bignspr to easily interpolate physical position to genetic positions.
+We provide function \texttt{snp\_asGeneticPos} in package bignspr to easily interpolate physical positions (in bp) to genetic positions (in cM).
 We recommend to use genetic positions and to use a size parameter of 3 cM when computing the correlation between variants for LDpred2. 
@@
 Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results.
+Combining results from all chromosomes requires using some appropriate scaling (see tutorial).
 
@@
 
-[TODO] 
-All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}.
-R package bigsnpr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
+%[TODO] 
+%All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}.
+The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
 A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.

### 2020-04-28 paper/paper-ldpred2.tex
include comments

@@
 \abstract{	
-In recent years the use of polygenic scores has become widespread.
-LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a Linkage Disequilibrium (LD) matrix only.
-However, LDpred has several limitations that may result in limited predictive performance.
-Here, we present LDpred2, a new version of LDpred that addresses these issues therefore providing higher predictive performance, while markedly improving its computational efficiency.
-We provide this faster and more robust implementation of LDpred in R package bigsnpr, as well as two new options in LDpred2: ``sparse'' and ``auto''.
+Polygenic scores have become a central tool in human genetics research.
+LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
+However, LDpred has limitations that may result in limited predictive performance.
+Here we present LDpred2, a new version of LDpred that addresses these issues.
+We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns parameters from data.
+We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
+LDpred2 is implemented in R package bigsnpr.
 }
@@
 Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
-PGS are also extensively used in epidemiology and economics as predictive variables of interest [TODO BJARNI CITATIONS]. 
+PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{horsdal2019association}. 
 For example, a recently derived PGS for education attainment has been one of the most predictive variables in behavioural sciences so far \cite[]{allegrini2019genomic}.
@@
 Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
-LDpred2 has 4 options: 1) LDpred2-inf, an infinitesimal model similar to the one in LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to tune.
+LDpred2 has 4 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to tune.
 As a recall, LDpred v1 has two options: LDpred1-grid where only $p$ is optimized, and LDpred1-inf.
 
-We compare the two versions of LDpred using six simulation scenarios to understand in which cases it is important to use the new version of LDpred.
-We also compare these two versions of LDpred using eight binary phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
+We compare the two versions of LDpred using six simulation scenarios to understand the expected impact of using the new version of LDpred.
+We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
 Area Under the ROC Curve (AUC) values are reported.
@@
 In these scenarios, infinitesimal models perform similarly. 
-When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is smaller, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3).
+When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is small, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3).
 For example, in simulations with 300 causal variants, a mean AUC of 73.5\% is obtained with LDpred1 while a value of 81.9\% is obtained with LDpred2.
 In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
-In HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in high LD, LDpred1 performs poorly.
-In these two HLA scenarios, all LDpred2 models perform well, although LDpred2-auto performs slightly worse than other models.
+As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in high LD.
+In contrast, all LDpred2 models perform well in these two HLA scenarios, although LDpred2-auto performs slightly worse than other LDpred2 models.
 
@@
 \caption{Results of the six simulation scenarios. 
-Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25-35 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
-Bars present the mean and 95\% CI of $10^4$ non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
+Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25.5-33.5 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
+Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
 %See corresponding values in table S1.
@@
 Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr).
-Eight binary phenotypes are used, summarized in table \ref{tab:sumstats}.
+Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
 For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1.
@@
 \caption{Results of the real data applications using published external summary statistics.
-Bars present AUC values on the test set of UKBB (mean and 95\% CI from $10^4$ bootstrap samples).
+Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
 %See corresponding values in table S2.
@@
 Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
-We propose a new version of LDpred to solve these instability issues and improve its computational efficiency.
+We therefore implemented a new version of LDpred to solve these instability issues and improve its computational efficiency.
 
@@
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (108 hyper-parameter values parallelized over 8 cores) and LDpred2-auto for a chromosome with 100K variants.
-Pre-computing the LD matrix for a chromosome with 100K variants takes 11 minutes using 16 cores.
+As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (108 hyper-parameter values parallelized over 8 cores) and LDpred2-auto for a chromosome with 100,000 variants.
+It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
 
@@
 Yet, we do not know yet why LDpred2-auto performs poorly specifically for PRCA.
-More investigations need to be performed to understand these poor results, and to find if this issue can be fixed.
+More investigations need to be performed to understand these poor results of LDpred2-auto in these two cases, and to find if this issue can be fixed.
 Solutions that come to mind include for example running multiple chains with different initial values of $p$ and keeping the one with the best convergence properties.
+More stringent quality control on summary statistics might help as well.
 
@@
 This results in 337,475 individuals and 1,194,574 variants.
-We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and for computing correlations between variants (LD matrix). 
+We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and for computing correlations between variants (LD matrix $\boldsymbol{R}$). 
 We use 300,000 other individuals for running logistic GWAS to create summary statistics.
@@
 We vary the number of causal variants (300, 3000, 30,000, or 300,000) to match a range of genetic architectures from low to high polygenicity.
-Liability scores are computed from a model with additive effects only: we compute the liability score of the $i$-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G_{i,j}} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon_i$ follows a Gaussian distribution $N(0, 1 - h^2)$.
+Liability scores are computed from a model with additive effects only: we compute the liability score of the $i$-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G}_{i,j} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G}_{i,j}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon_i$ follows a Gaussian distribution $N(0, 1 - h^2)$.
 Causal variants are chosen randomly anywhere on the genome. 
-We make two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25-35 Mb).
+We make two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25.5-33.5 Mb).
 Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
@@
 
-In this section, we explain how we can obtain joint effects from summary statistics (marginal effects) and a correlation matrix.
-Let us denote by $S$ the diagonal matrix with standard deviations of the $m$ variants, $C_n = I_n - 1 1^T / n$ the centering matrix, $G$ the genotype matrix of $n$ individuals and $m$ variants, and $y$ the phenotype vector for $n$ individuals.
+In this section, we explain how we can obtain joint effects from summary statistics (marginal effects) and a correlation matrix $\boldsymbol{R}$.
+Let us denote by $\boldsymbol{S}$ the diagonal matrix with standard deviations of the $m$ variants, $\boldsymbol{C_n} = \boldsymbol{I_n} - \boldsymbol{1} \boldsymbol{1}^T / n$ the centering matrix, $\boldsymbol{G}$ the genotype matrix of $n$ individuals and $m$ variants, and $\boldsymbol{y}$ the phenotype vector for $n$ individuals.
 
-When solving a joint model with all variants and an intercept ${\alpha}$, the joint effects ${\gamma}_{\text{joint}}$ are obtained by solving
-\[\begin{bmatrix} \hat{\alpha} \\ \hat{\gamma}_{\text{joint}} \end{bmatrix} = \left(\begin{bmatrix} 1 & G \end{bmatrix}^T \begin{bmatrix} 1 & G \end{bmatrix}\right)^{-1} \begin{bmatrix} 1 & G \end{bmatrix}^T y ~.\]
+When solving a joint model with all variants and an intercept ${\alpha}$, the joint effects $\boldsymbol{\gamma}_{\text{joint}}$ are obtained by solving
+\[\begin{bmatrix} \hat{\alpha} \\ \boldsymbol{\hat{\gamma}_{\text{joint}}} \end{bmatrix} = \left(\begin{bmatrix} \boldsymbol{1} & \boldsymbol{G} \end{bmatrix}^T \begin{bmatrix} \boldsymbol{1} & \boldsymbol{G} \end{bmatrix}\right)^{-1} \begin{bmatrix} \boldsymbol{1} & \boldsymbol{G} \end{bmatrix}^T \boldsymbol{y} ~.\]
 Using the Woodburry formula, we get 
-\[\hat{\gamma}_{\text{joint}} = (G^T C_n G)^{-1} G^T C_n y ~.\]
+\[\boldsymbol{\hat{\gamma}_{\text{joint}}} = (\boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{G})^{-1} \boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{y} ~.\]
 When fitting each variant separately in GWAS, the marginal effects (assuming no covariate) simplify to 
-\[\hat{\gamma}_{\text{marg}} = \dfrac{1}{n-1} S^{-2} G^T C_n y ~.\] 
-We further note that the correlation matrix of $G$ is \(K =  \dfrac{1}{n-1} S^{-1} G^T C_n G  S^{-1}\).
-Then we get \[\hat{\gamma}_{\text{joint}} = S^{-1} K^{-1} S \hat{\gamma}_{\text{marg}} ~.\]
+\[\boldsymbol{\hat{\gamma}_{\text{marg}}} = \dfrac{1}{n-1} \boldsymbol{S}^{-2} \boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{y} ~.\] 
+We further note that the correlation matrix of $\boldsymbol{G}$ is \(\boldsymbol{R} =  \dfrac{1}{n-1} \boldsymbol{S}^{-1} \boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{G}  \boldsymbol{S}^{-1}\).
+Then we get 
+\begin{equation}\label{eq:joint}
+\boldsymbol{\hat{\gamma}_{\text{joint}}} = \boldsymbol{S}^{-1} \boldsymbol{R}^{-1} \boldsymbol{S} \boldsymbol{\hat{\gamma}_{\text{marg}}} ~.
+\end{equation}
+In practice, the correlation matrix $\boldsymbol{R}$ is usually not available but is computed from another dataset. 
+Also note that $\boldsymbol{\gamma}$ are the effects on the allele scale while we denote by $\boldsymbol{\beta} = \boldsymbol{S} \boldsymbol{\gamma}$ the effects of the scaled genotypes.
 
-For one marginal effect $\hat{\gamma}$, let us denote by $y^*$ and $g^*$ the corresponding phenotype and genotype (residualized from $K$ covariates, e.g.\ centering them).
+For the marginal effect $\hat{\gamma}_j$ of variant $j$, let us denote by $\boldsymbol{\breve{y}}$ and $\boldsymbol{\breve{G}_j}$ the vectors of phenotypes and genotypes for variant $j$ residualized from $K$ covariates, e.g.\ centering them.
 Then,
-$\text{var}(\hat{\gamma}) = \dfrac{(y^* - \hat{\gamma} g^*)^T (y^* - \hat{\gamma} g^*)}{(N - K - 1) ~ g^{*T} g^*} \approx \dfrac{y^{*T} y^*}{N ~ g^{*T} g^*} \approx \dfrac{\text{var}(y)}{N ~ \text{var}(g)}$.
-Thus, we can derive $\text{sd}(g) \approx \dfrac{\text{sd}(y)}{\text{se}(\hat{\gamma}) ~ \sqrt{N}}$ and $\left(\text{sd}(g) ~ \hat{\gamma}\right) \approx \dfrac{\hat{\gamma}}{\text{se}(\hat{\gamma})} \dfrac{\text{sd}(y)}{\sqrt{N}}$.
-
-Let us go back to the formula we derived before: $\hat{\gamma}_{\text{joint}} = S^{-1} K^{-1} S \hat{\gamma}_{\text{marg}}$. 
-Note that $\gamma$ are the effects on the allele scale while we denote by $\beta = S \gamma$ the effects of the scaled genotypes.
-As $\text{sd}(y)$ is the same for all variants, it is cancelled out by $S^{-1}$ and $S$.
-Then it justifies the use of Z-scores ($\hat{\gamma} ~/~ \text{se}(\hat{\gamma})$) divided by $\sqrt{N}$ as input for LDpred.
-Then, the effect sizes that LDpred outputs need to be scaled by multiplying by $\left(\text{se}(\hat{\gamma}) ~ \sqrt{N}\right)$.
-Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{sd}(y) = 1$ only.
+\[\left(\text{se}(\hat{\gamma}_j)\right)^2 = \dfrac{(\boldsymbol{\breve{y}} - \hat{\gamma}_j \boldsymbol{\breve{G}_j})^T (\boldsymbol{\breve{y}} - \hat{\gamma}_j \boldsymbol{\breve{G}_j})}{(n - K - 1) ~ \boldsymbol{\breve{G}_j}^T \boldsymbol{\breve{G}_j}} \approx \dfrac{\boldsymbol{\breve{y}}^T \boldsymbol{\breve{y}}}{n ~ \boldsymbol{\breve{G}_j}^T \boldsymbol{\breve{G}_j}} \approx \dfrac{\text{var}(\boldsymbol{y})}{n ~ \text{var}(\boldsymbol{g_j})} ~.\]
+Thus, we can derive $\text{sd}(\boldsymbol{g_j}) \approx \dfrac{\text{sd}(\boldsymbol{y})}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n}}$ and $\left(\text{sd}(\boldsymbol{g_j}) ~ \hat{\gamma}_j\right) \approx \dfrac{\hat{\gamma}_j}{\text{se}(\hat{\gamma}_j)} \dfrac{\text{sd}(\boldsymbol{y})}{\sqrt{n}}$.
+Let us go back to equation \ref{eq:joint}. 
+As $\text{sd}(\boldsymbol{y})$ is the same for all variants, it is cancelled out by $\boldsymbol{S}^{-1}$ and $\boldsymbol{S}$, therefore we can assume that $\text{var}(\boldsymbol{y}) = 1$.
+Then it justifies the use of the Z-scores ($\hat{\gamma}_j ~/~ \text{se}(\hat{\gamma}_j)$) divided by $\sqrt{n}$ as input for LDpred (first line of algorithm \ref{algo:ldpred}).
+Then, the effect sizes that LDpred outputs need to be scaled back by multiplying by $\left(\text{se}(\hat{\gamma}_j) ~ \sqrt{n}\right)$ (last line of algorithm \ref{algo:ldpred}).
+Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{var}(\boldsymbol{y}) = 1$ only.
 
@@
 \begin{equation}\label{eq:model}
-\beta_j = S \gamma_j \sim \left\{
+\beta_j = S_{j,j} \gamma_j \sim \left\{
 \begin{array}{ll}
@@
 
-\newcommand{\hmpn}{\dfrac{h^2}{M p} + \dfrac{1}{N}}
+\newcommand{\hmpn}{\dfrac{h^2}{M p} + \dfrac{1}{n}}
 \newcommand{\phmpn}{\dfrac{p}{\sqrt{\hmpn}}}
@@
 To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.
-First, they estimate the residualized marginal effect for variant $j$ as [SHOULD VERIFY]
+First, the residualized marginal effect for variant $j$ is computed as
 \begin{equation}\label{eq:beta_res}
-\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{K}_{\text{-}j, j}
+\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{R}_{\text{-}j, j}
 \end{equation}
-where $\boldsymbol{K}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix and $\boldsymbol{\hat{\beta}}$ are the marginal effect sizes.
-Then, they compute the probability that variant $j$ is causal as 
-\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{N}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{N}}\right\rbrace}~,\]
+where $\boldsymbol{R}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix, $\boldsymbol{\hat{\beta}}$ is the vector of marginal effect sizes, and $\boldsymbol{{\beta}}$ is the vector of current effect sizes in the Gibbs sampler.
+Then, the probability that variant $j$ is causal is computed as 
+\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{n}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{n}}\right\rbrace}~,\]
 which we rewrite as
 \begin{equation}
-\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{N h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{N \tilde{\beta}_j^2}{1 + \dfrac{Mp}{N h^2}}\right\rbrace}~.\label{eq:postp}
+\bar{p}_j = \dfrac{1}{1 + \dfrac{1-p}{p} \sqrt{1 + \dfrac{n h^2}{Mp}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{n \tilde{\beta}_j^2}{1 + \dfrac{Mp}{n h^2}}\right\rbrace}~.\label{eq:postp}
 \end{equation}
-Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $N \tilde{\beta}_j^2$ is large.
+Computing $\bar{p}_j$ using the second expression is important to avoid numerical issues when $\left(n \tilde{\beta}_j^2\right)$ is large.
 
@@
 \begin{array}{ll}
-\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{N h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{N h^2}}\dfrac{1}{N}\right) & \mbox{with probability $\bar{p}_j$,} \\
+\mathcal N\left(\dfrac{1}{1 + \dfrac{M p}{n h^2}} \tilde{\beta}_j, \dfrac{1}{1 + \dfrac{M p}{n h^2}}\dfrac{1}{n}\right) & \mbox{with probability $\bar{p}_j$,} \\
 0 & \mbox{otherwise.}
@@
 \begin{equation}\label{eg:post_beta}
-\omega_j = \dfrac{\bar{p}_j \tilde{\beta}_j}{1 + \dfrac{M p}{N h^2}} ~.
+\omega_j = \dfrac{\bar{p}_j \tilde{\beta}_j}{1 + \dfrac{M p}{n h^2}} ~.
 \end{equation}
@@
 \begin{algorithm}[H]
-\caption{LDpred, with hyper-parameters $p$ and $h^2$, and summary statistics $\boldsymbol{\hat{\gamma}}$, $\boldsymbol{\text{se}(\hat{\gamma})}$ and $\boldsymbol{N}$}\label{algo:ldpred}
+\caption{LDpred, with hyper-parameters $p$ and $h^2$, LD matrix $\boldsymbol{R}$ and summary statistics $\boldsymbol{\hat{\gamma}}$, $\boldsymbol{\text{se}(\hat{\gamma})}$ and $\boldsymbol{n}$}\label{algo:ldpred}
 \begin{algorithmic}[1]
-	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}}$ \Comment{See previous section}
-	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$
+	\State $\boldsymbol{\hat{\beta}} \gets \dfrac{\boldsymbol{\hat{\gamma}}}{\boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{n}}}$ \Comment{Initialization of scaled marginal effects (see previous section)}
+	\State $\boldsymbol{\Omega} \gets \boldsymbol{0}$ \Comment{Initialization of posterior means}
 	\For {$k=1,\ldots,N_\text{burn-in} + N_\text{iter}$} \Comment{Gibbs iterations}
@@
 	\EndFor
-	\State Compute $\boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
+	\State Compute $\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
 	\If {$k > N_\text{burn-in}$} 
-	\State Increment $\boldsymbol{\Omega}$ with $\boldsymbol{\omega}$ 
+	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} + \boldsymbol{\omega}$ 
 	\EndIf
@@
 	\State $\boldsymbol{\Omega} \gets \boldsymbol{\Omega} /  N_\text{iter}$ \Comment{Average of all $\boldsymbol{\omega}$ after burn-in}
-	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{N}}$ \Comment{See previous section}
+	\State Return $\boldsymbol{\Omega} \cdot  \boldsymbol{\text{se}(\hat{\gamma})} \cdot \sqrt{\boldsymbol{n}}$ \Comment{Return posterior means, scaled back (see previous section)}
 \end{algorithmic} 
@@
 This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly to data without the need of a validation dataset to choose best-performing hyper-parameters.
-To estimate $p$, we count the number of causal variants (i.e.\ $m_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
-We can assume that $m_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + m_c, 1 + M - m_c)$.
+To estimate $p$, we count the number of causal variants (i.e.\ $M_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
+We can assume that $M_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + M_c, 1 + M - M_c)$.
 Due to complexity reasons, we could not derive a Bayesian estimator of $h^2$.
-Instead, we estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{K} \boldsymbol{\beta}$, where $\boldsymbol{K}$ is the correlation matrix.
+Instead, we estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix.
 These parameters $p$ and $h^2$ are updated after the inner loop in algorithm \ref{algo:ldpred}, then these new values are used in the next iteration of the outer loop.

### 2020-05-26 paper/paper-ldpred2.tex
save

@@
 \usepackage{graphicx}
-\usepackage{subcaption}
 \usepackage{amsmath, amsthm, amssymb}
@@
 
-\usepackage{xr}
-\externaldocument{paper-ldpred2-supp}
+%\usepackage{xr}
+%\externaldocument{paper-ldpred2-supp}
 
@@
 Area Under the ROC Curve (AUC) values are reported.
+Finally, we compare LDpred2 to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
 
@@
 \centerline{\includegraphics[width=0.85\textwidth]{AUC-real}}
-\caption{Results of the real data applications using published external summary statistics.
+\caption{Results of LDpred2 and LDpred1 for the real data applications using published external summary statistics.
 Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
@@
 \label{fig:AUC-real}
+\end{figure}
+
+When comparing to other methods, we only report the best LDpred2 model (using the validation set). 
+LDpred2 performs best for Asthma and TD2 while lassosum performs best for BRCA, MDD and PRCA (Figure \ref{fig:AUC-all}).
+On average, methods perform very similarly.
+
+[TODO: COMPLETE]
+
+%% FIGURE 3 HERE
+\begin{figure}[h]
+\centerline{\includegraphics[width=0.85\textwidth]{AUC-all}}
+\caption{Results of LDpred2, C+T, SCT, lassosum and PRS-CS in the real data applications using published external summary statistics.
+Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
+%See corresponding values in table S2.
+}
+\label{fig:AUC-all}
 \end{figure}
@@
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (108 hyper-parameter values parallelized over 8 cores) and LDpred2-auto for a chromosome with 100,000 variants.
+As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (102 hyper-parameter values parallelized over 8 cores) and LDpred2-auto for a chromosome with 100,000 variants.
 It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
@@
 We use the UK Biobank data for both real data analyses and simulations \cite[]{bycroft2018uk}. 
-We restrict individuals to White-British individuals used in the PC computation of the UK Biobank (i.e.\ unrelated and quality-controlled).
-We restrict variants to HapMap3 variants.
-This results in 337,475 individuals and 1,194,574 variants.
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have pass some quality control.
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
+We restrict variants to HapMap3 variants used in PRS-CS \cite[]{ge2019polygenic}.
+This results in 362,320 individuals and 1,117,493 variants.
 We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and for computing correlations between variants (LD matrix $\boldsymbol{R}$). 
 We use 300,000 other individuals for running logistic GWAS to create summary statistics.
-We use the remaining 27,475 individuals as test set for evaluating models.
+We use the remaining 52,320 individuals as test set for evaluating models.
 
@@
 
-We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 327,475 individuals as test set.
+We use the same data as in the simulation analyses. We use the same 10,000 individuals as validation set, and use the remaining 352,320 individuals as test set.
 We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
@@
 \end{table}
+
+In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
+Then, we compare of the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
+For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}.
+For C+T, we choose the best set of parameters on each chromosome individually and combine them using a proper scaling as done for LDpred2 (see section \ref{sec:bychr}).
+For lassosum, we use the default grid of hyper-parameters $s$ and $\lambda$. 
+Finally, for PRS-CS, we use the grid $\{10^{-6}, 10^{-5}, \dots, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters $a$ and $b$.
 
@@
 Note that, in the code, we use \texttt{size = 3 / 1000} since parameter \texttt{size} is internally multiplied by 1000 in functions of package bigsnpr.
-To prevent storing very small correlations, we also set to 0 all correlations with a p-value larger than \texttt{alpha = 0.9} (when testing that the correlation is different from 0).
 
-\subsection*{New strategy for running LDpred2}
+\subsection*{New strategy for running LDpred2}\label{sec:bychr}
 
@@
 Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results.
-Combining results from all chromosomes requires using some appropriate scaling (see tutorial).
+
+To choose the best model from LDpred2-grid, one can choose the best model according to their preferred criterion (e.g. max AUC). 
+Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC. 
+Using the regression also enables adjusting for covariates in this step.
+Combining results from all chromosomes requires using some appropriate scaling.
+Therefore, we scale the effect sizes by the regresion coefficient to be on a proper scaling when adding it to other chromosomes' scores.
 

### 2020-05-29 paper/paper-ldpred2.tex
save

@@
 We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
+We then show that LDpred2 compares favorably to other polygenic score methods.
 LDpred2 is implemented in R package bigsnpr.
@@
 In these scenarios, infinitesimal models perform similarly. 
-When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is small, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.5e-4 and 2.5e-3).
-For example, in simulations with 300 causal variants, a mean AUC of 73.5\% is obtained with LDpred1 while a value of 81.9\% is obtained with LDpred2.
+When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is small, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.7e-4 and 2.7e-3).
+For example, in simulations with 300 causal variants, a mean AUC of 73.5\% is obtained with LDpred1 while a value of 81.9\% is obtained with LDpred2 [TODO: VERIF NEW NUMBERS].
 In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
-As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in high LD.
+As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in long-range LD. 
 In contrast, all LDpred2 models perform well in these two HLA scenarios, although LDpred2-auto performs slightly worse than other LDpred2 models.
@@
 %% FIGURE 1 HERE
-\begin{figure}[h]
+\begin{figure}[htbp]
 \centerline{\includegraphics[width=0.95\textwidth]{AUC-simu}}
@@
 As in simulations, the sparse version of LDpred2 performs equally well as the original version.
-For PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly.
+For PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly [TODO: CHANGE].
 
 %% FIGURE 2 HERE
-\begin{figure}[h]
+\begin{figure}[htbp]
 \centerline{\includegraphics[width=0.85\textwidth]{AUC-real}}
@@
 
-When comparing to other methods, we only report the best LDpred2 model (using the validation set). 
-LDpred2 performs best for Asthma and TD2 while lassosum performs best for BRCA, MDD and PRCA (Figure \ref{fig:AUC-all}).
-On average, methods perform very similarly.
+When comparing to other methods, we only report the best LDpred2 model (chosen using the validation set). 
+LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).
 
@@
 %% FIGURE 3 HERE
-\begin{figure}[h]
+\begin{figure}[htbp]
 \centerline{\includegraphics[width=0.85\textwidth]{AUC-all}}
@@
 Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
-We therefore implemented a new version of LDpred to solve these instability issues and improve its computational efficiency.
+We have therefore implemented a new version of LDpred to solve these instability issues and improve its computational efficiency.
 
-We show that LDpred2 is much more stable and provides higher prediction than LDpred1.
+We show that LDpred2 is much more stable and provides higher predictive performance than LDpred1.
 We demonstrate that LDpred1 has defects, particularly when handling long-range LD regions such as the HLA region.
@@
 
-In LDpred2, we also test more values for $p$ (17 instead of 7 by default) and for $h^2$ (3 instead of 1).
-When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see after).
+In LDpred2, we also test more values for $p$ (21 instead of 7 by default) and for $h^2$ (3 instead of 1).
+When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see below).
 Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1.
-We also use 5 times as many iterations in the Gibbs sampler and a larger window size for computing correlations between variants, yet LDpred2 is still as fast as LDpred1.
+We also use a larger window size for computing correlations between variants, yet LDpred2 is still as fast as LDpred1.
 LDpred2 is fast because we provide an efficient parallel implementation in C++.
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (102 hyper-parameter values parallelized over 8 cores) and LDpred2-auto for a chromosome with 100,000 variants.
+As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (102 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
 It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
@@
 This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs better or equally well as the non-sparse model.
-As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, but does not perform well for some of the real traits analyzed here.
-This is expected for T1D because it is mainly composed of large effects in the HLA region (see similar simulation results) and because summary statistics have a small sample size (5913 cases and 8828 controls only).
-Yet, we do not know yet why LDpred2-auto performs poorly specifically for PRCA.
-More investigations need to be performed to understand these poor results of LDpred2-auto in these two cases, and to find if this issue can be fixed.
-Solutions that come to mind include for example running multiple chains with different initial values of $p$ and keeping the one with the best convergence properties.
-More stringent quality control on summary statistics might help as well.
+As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, and also performs very well in real data applications, provided that some quality control is performed on summary statistics (see Supplementary Note).
 
-%[TALK ABOUT QUALITY CONTROL ON AF?]
+In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
+When comparing to recently derived methods that showed higher predictive performance as compared to the previous version of LDpred, LDpred2 performs favorably, providing the highest predictive performance for half of the real traits tested, and very good predictive performance for the other half.
+Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
+Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.
+Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, which may break some of the assumptions behind their respective models.
+Selection of variants is therefore a direction of future research for us.
 
-%[TALK ABOUT DIFFERENT SETS OF VARIANTS?]
-
-%[TALK ABOUT MODEL FOR BINARY DATA]
+[TALK ABOUT QUALITY CONTROL ON AF? -> ON SD, NEED A SUPP NOTE] 
+[ALSO TALK ABOUT MODEL FOR BINARY DATA]
 
@@
 We use the UK Biobank data for both real data analyses and simulations \cite[]{bycroft2018uk}. 
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have pass some quality control.
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have passed some quality control.
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
@@
 Causal variants are chosen randomly anywhere on the genome. 
-We make two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25.5-33.5 Mb).
+We design two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25.5-33.5 Mb).
 Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
@@
 We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We defined phenotypes as in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section).
+We defined phenotypes as in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section). [TODO: TALK ABOUT QC HERE]
 
@@
 In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
-Then, we compare of the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
+Then, we compare the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
 For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}.
@@
 As $\text{sd}(\boldsymbol{y})$ is the same for all variants, it is cancelled out by $\boldsymbol{S}^{-1}$ and $\boldsymbol{S}$, therefore we can assume that $\text{var}(\boldsymbol{y}) = 1$. 
-Then it justifies the use of the Z-scores ($\hat{\gamma}_j ~/~ \text{se}(\hat{\gamma}_j)$) divided by $\sqrt{n}$ as input for LDpred (first line of algorithm \ref{algo:ldpred}).
+It justifies the use of the Z-scores ($\hat{\gamma}_j ~/~ \text{se}(\hat{\gamma}_j)$) divided by $\sqrt{n}$ as input for LDpred (first line of algorithm \ref{algo:ldpred}).
 Then, the effect sizes that LDpred outputs need to be scaled back by multiplying by $\left(\text{se}(\hat{\gamma}_j) ~ \sqrt{n}\right)$ (last line of algorithm \ref{algo:ldpred}).
 Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{var}(\boldsymbol{y}) = 1$ only.
+In the case of binary phenotypes, the approximation of the standard deviations is slightly different (see Supplementary Note).
 
@@
 	\EndFor
-	\State Compute $\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$ (estimate of $h^2$) to check divergence \Comment{Return 0s if divergence}
 	\If {$k > N_\text{burn-in}$} 
@@
 
-LDpred has a window size parameter that needs to be set; for a given variant, correlations with other variants outside of this window are assumed to be 0.
+There is a window size parameter that needs to be set in LDpred; for a given variant, correlations with other variants outside of this window are assumed to be 0.
 The recommended value for this window (in number of variants) is to use the total number of variants divided by 3000, which corresponds to a window radius of around 2 Mb \cite[]{vilhjalmsson2015modeling}.
@@
 We recommend to use genetic positions and to use a size parameter of 3 cM when computing the correlation between variants for LDpred2. 
-Note that, in the code, we use \texttt{size = 3 / 1000} since parameter \texttt{size} is internally multiplied by 1000 in functions of package bigsnpr.
+Note that, in the code, we use \texttt{size = 3 / 1000} since parameter \texttt{size} is internally multiplied by 1000 in the bigsnpr functions.
 
@@
 
-We recommend to run LDpred2 per chromosome.
+We run LDpred2 per chromosome.
 Even if it is possible to run LDpred2 genome-wide, this approach has two limitations.
@@
 Doubling to 2M variants would require 128 GB of RAM to store the matrix.
-Second, as noted in \cite{prive2019making}, it may be beneficial to assume that architecture of traits may be different for different chromosomes.
+Second, as noted in \cite{prive2019making}, it may be beneficial to assume that the architecture of traits is different for different chromosomes.
 For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
-Assuming the same model for genetic effects genome-wide could result in severe model misspecification, which would lead to suboptimal predictive performance.
+Assuming the same model for genetic effects genome-wide could result in model misspecification, which would lead to suboptimal predictive performance.
 Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results. 
-
-To choose the best model from LDpred2-grid, one can choose the best model according to their preferred criterion (e.g. max AUC). 
+To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
 Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC. 
-Using the regression also enables adjusting for covariates in this step.
-Combining results from all chromosomes requires using some appropriate scaling.
-Therefore, we scale the effect sizes by the regresion coefficient to be on a proper scaling when adding it to other chromosomes' scores.
+Using the regression would also enable adjusting for covariates in this step.
 
@@
 
-%[TODO] 
-%All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}.
+[TODO]
+
+All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}.
 The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
@@
 
+Authors thank Yixuan Qiu for pointing to matrix-free solvers.
 This research has been conducted using the UK Biobank Resource under Application Number 41181.
+Authors would also like to thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
+
+\section*{Funding}
 

### 2020-05-31 paper/paper-ldpred2.tex
save

@@
 We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
-We then show that LDpred2 compares favorably to other polygenic score methods.
+We then show that LDpred2 compares favorably to other polygenic score methods developed more recently.
 LDpred2 is implemented in R package bigsnpr.
@@
 When testing a grid of hyper-parameters, LDpred2 performs substantially better than LDpred1 in the cases where the number of causal variants is small, i.e.\ in the case of 300 or 3000 causal variants ($p$=2.7e-4 and 2.7e-3).
-For example, in simulations with 300 causal variants, a mean AUC of 73.5\% is obtained with LDpred1 while a value of 81.9\% is obtained with LDpred2 [TODO: VERIF NEW NUMBERS].
+For example, in simulations with 300 causal variants, a mean AUC of 73.6\% is obtained with LDpred1 while a value of 81.7\% is obtained with LDpred2.
 In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
-As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled in a region with variants in long-range LD. 
+As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled from a long-range LD region. 
 In contrast, all LDpred2 models perform well in these two HLA scenarios, although LDpred2-auto performs slightly worse than other LDpred2 models.
@@
 %% FIGURE 1 HERE
-\begin{figure}[htbp]
-\centerline{\includegraphics[width=0.95\textwidth]{AUC-simu}}
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.85\textwidth]{AUC-simu}}
 \caption{Results of the six simulation scenarios. 
@@
 Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
-For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1.
-These improvements are also significant for MDD and CAD, albeit to a lesser extend.
-For example, for BRCA, AUC improves from 58.5\% with LDpred1 to 64.4\% with LDpred2, and from 54.9\% to 74.3\% for T1D.
+For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1. These improvements are also significant for MDD and CAD, albeit to a lesser extend. 
+For example, for BRCA, AUC improves from 58.9\% with LDpred1 to 64.6\% with LDpred2, and from 57.4\% to 77.0\% for T1D. 
 For Asthma and RA, predictive performances of LDpred1 and LDpred2 are similar.
-
 As in simulations, the sparse version of LDpred2 performs equally well as the original version.
-For PRCA and TD1, LDpred2-auto perform much worse than LDpred2, otherwise it performs similarly [TODO: CHANGE].
+Moreover, LDpred2-auto also perform  very well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
+When comparing to other PGS methods, we only report the best LDpred2 model (chosen using the validation set). 
+LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).
 
@@
 \end{figure}
-
-When comparing to other methods, we only report the best LDpred2 model (chosen using the validation set). 
-LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).
-
-[TODO: COMPLETE]
 
@@
 Here we present LDpred2, a new version of LDpred. 
-LDpred is widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
+LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
 Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
@@
 In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
-When comparing to recently derived methods that showed higher predictive performance as compared to the previous version of LDpred, LDpred2 performs favorably, providing the highest predictive performance for half of the real traits tested, and very good predictive performance for the other half.
+When comparing to recently derived methods that showed higher predictive performance compared to the previous version of LDpred, LDpred2 performs favorably, providing the highest predictive performance for half of the real traits tested, and very good predictive performance for the other half.
 Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
 Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.
-Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, which may break some of the assumptions behind their respective models.
+Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, or detrimental as it may break some of the assumptions behind their respective models.
 Selection of variants is therefore a direction of future research for us.
-
-[TALK ABOUT QUALITY CONTROL ON AF? -> ON SD, NEED A SUPP NOTE] 
-[ALSO TALK ABOUT MODEL FOR BINARY DATA]
 
@@
 We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We defined phenotypes as in \cite{prive2019making}. For details, please refer to our R code (Software and code availability section). [TODO: TALK ABOUT QC HERE]
+We defined phenotypes as in \cite{prive2019making}. Some quality control is applied to summary statistics (see Supplementary Note).
+For more details, please refer to our R code (Software and code availability section).
 
@@
 For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}.
-For C+T, we choose the best set of parameters on each chromosome individually and combine them using a proper scaling as done for LDpred2 (see section \ref{sec:bychr}).
 For lassosum, we use the default grid of hyper-parameters $s$ and $\lambda$. 
@@
 
-%\begin{minipage}{.7\textwidth}
 \begin{algorithm}[H]
@@
 \end{algorithm}
-%\end{minipage}
-
-%[CHANGE NOTATION OF SUMSTATS?]
 
@@
 Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC.
-Using the regression would also enable adjusting for covariates in this step.
 
@@
 
-[TODO]
+[TODO: EXPORT CODE FROM CLUSTER]
 
@@
 
+\clearpage
+
+\section*{Supplementary Note: quality control of summary statistics}
+
+For summary statistics of binary traits derived from a logistic regression, 
+\begin{equation}\label{eq:approx-sd}
+\text{sd}(\boldsymbol{g_j}) \approx \dfrac{2}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n_\text{eff}}} ~,
+\end{equation}
+where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. Anyway, we recommend to verify this assumption and to perform some quality control.
+Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
+However, in real data applications, where summary statistics comes from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
+Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set.
+We removed variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
+
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=0.7\textwidth]{sd-approx-simu}}
+\caption{In simulations, standard deviations derived from summary statistics based on equation \eqref{eq:approx-sd} versus the standard deviations of genotypes of individuals in the validation set.}
+\label{fig:sd-approx-simu}
+\end{figure}
+
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=0.9\textwidth]{sd-approx-BRCA}}
+\caption{Standard deviations derived from summary statistics of breast cancer based on equation \eqref{eq:approx-sd} versus the standard deviations of genotypes of individuals in the validation set.
+Coloring shows the quality control applied in this paper.}
+\label{fig:sd-approx-brca}
+\end{figure}
+
 \end{document}

### 2020-06-01 paper/paper-ldpred2.tex
supp note

@@
 
-\clearpage
-
-\section*{Supplementary Note: quality control of summary statistics}
-
-For summary statistics of binary traits derived from a logistic regression, 
-\begin{equation}\label{eq:approx-sd}
-\text{sd}(\boldsymbol{g_j}) \approx \dfrac{2}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n_\text{eff}}} ~,
-\end{equation}
-where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. Anyway, we recommend to verify this assumption and to perform some quality control.
-Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
-However, in real data applications, where summary statistics comes from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
-Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set.
-We removed variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
-
-\begin{figure}[htbp]
-\centerline{\includegraphics[width=0.7\textwidth]{sd-approx-simu}}
-\caption{In simulations, standard deviations derived from summary statistics based on equation \eqref{eq:approx-sd} versus the standard deviations of genotypes of individuals in the validation set.}
-\label{fig:sd-approx-simu}
-\end{figure}
-
-\begin{figure}[htbp]
-\centerline{\includegraphics[width=0.9\textwidth]{sd-approx-BRCA}}
-\caption{Standard deviations derived from summary statistics of breast cancer based on equation \eqref{eq:approx-sd} versus the standard deviations of genotypes of individuals in the validation set.
-Coloring shows the quality control applied in this paper.}
-\label{fig:sd-approx-brca}
-\end{figure}
-
 \end{document}

### 2020-06-21 paper/paper-ldpred2.tex
save

@@
 We then show that LDpred2 compares favorably to other polygenic score methods developed more recently.
+LDpred2-auto, which automatically estimates the parameters of heritability and proportion of causal variants, performs particularly well, provided some quality control is performed.
 LDpred2 is implemented in R package bigsnpr.
@@
 Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
-PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{horsdal2019association}. 
+PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association}. 
 For example, a recently derived PGS for education attainment has been one of the most predictive variables in behavioural sciences so far \cite[]{allegrini2019genomic}.
@@
 We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing prediction accuracy.
+Moreover, LDpred2-auto, which does not require any validation set, perform very well, provided some quality control is performed.
 
@@
 \centerline{\includegraphics[width=0.85\textwidth]{AUC-simu}}
-\caption{Results of the six simulation scenarios. 
+\caption{Six variants of LDpred are compared (two variants of LDpred1 in magenta and orange, and four variants of LDpred2 in blue and green colors) in the six simulation scenarios. 
 Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25.5-33.5 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
@@
 Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
-For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1. These improvements are also significant for MDD and CAD, albeit to a lesser extend. 
+For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1.
 For example, for BRCA, AUC improves from 58.9\% with LDpred1 to 64.6\% with LDpred2, and from 57.4\% to 77.0\% for T1D.
-For Asthma and RA, predictive performances of LDpred1 and LDpred2 are similar.
+These improvements are also significant for MDD and CAD, albeit to a lesser extend. 
+For Asthma and RA, predictive performance of LDpred1 and LDpred2 are similar.
 As in simulations, the sparse version of LDpred2 performs equally well as the original version.
 Moreover, LDpred2-auto also perform  very well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
-When comparing to other PGS methods, we only report the best LDpred2 model (chosen using the validation set). 
+When comparing to other PGS methods, we only report the best LDpred2 model (chosen using the validation set) and LDpred2-auto. 
 LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).
+LDpred2-auto usually performs even better than the LDpred-best (Figure \ref{fig:AUC-all}).
+This means that choosing the best LDpred model per chromosome leads to overfitting in the validation set, which does not transfer favorably when evaluating prediction performance in the test set.
 
@@
 \centerline{\includegraphics[width=0.85\textwidth]{AUC-real}}
-\caption{Results of LDpred2 and LDpred1 for the real data applications using published external summary statistics.
+\caption{Six variants of LDpred are compared (two variants of LDpred1 in magenta and orange, and four variants of LDpred2 in blue and green colors) in the real data applications using published external summary statistics.
 Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
@@
 \centerline{\includegraphics[width=0.85\textwidth]{AUC-all}}
-\caption{Results of LDpred2, C+T, SCT, lassosum and PRS-CS in the real data applications using published external summary statistics.
+\caption{LDpred2 is compared with C+T, SCT, lassosum and PRS-CS in the real data applications using published external summary statistics.
 Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
@@
 When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see below).
-Overall, we test a grid of 102 different values in LDpred2 instead of 7 in LDpred1.
+Overall, we test a grid of 126 different values in LDpred2 instead of 7 in LDpred1.
 We also use a larger window size for computing correlations between variants, yet LDpred2 is still as fast as LDpred1.
@@
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (102 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
+Efficient parallelization is possible because we developed an on-disk sparse matrix format accessed via memory mapping in R package bigsparser. 
+As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
 It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
@@
 As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, and also performs very well in real data applications, provided that some quality control is performed on summary statistics (see Supplementary Note).
+This quality control aims at ensuring that effects are transferable from the summary statistics to the data where PGS are computed.
 
 In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
-When comparing to recently derived methods that showed higher predictive performance compared to the previous version of LDpred, LDpred2 performs favorably, providing the highest predictive performance for half of the real traits tested, and very good predictive performance for the other half.
+When comparing to recently derived methods that showed higher predictive performance compared to the previous version of LDpred, LDpred2 performs favorably. 
+Indeed, LDpred2 (especially LDpred2-auto) provides the highest predictive performance for almost all the real traits tested, and very good predictive performance for the others.
 Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
@@
 Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, or detrimental as it may break some of the assumptions behind their respective models.
-Selection of variants is therefore a direction of future research for us.
+For now, we recommend using the same variants used in PRS-CS and used here.
+For example, we do not think using several million variants would improve predictive performance of LDpred2.
+Investigating selection of variants is a direction of future research for us.
 
@@
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
-We restrict variants to HapMap3 variants used in PRS-CS \cite[]{ge2019polygenic}.
+We restrict variants to the HapMap3 variants used in PRS-CS \cite[]{ge2019polygenic}.
 This results in 362,320 individuals and 1,117,493 variants.
@@
 Then,
-\[\left(\text{se}(\hat{\gamma}_j)\right)^2 = \dfrac{(\boldsymbol{\breve{y}} - \hat{\gamma}_j \boldsymbol{\breve{G}_j})^T (\boldsymbol{\breve{y}} - \hat{\gamma}_j \boldsymbol{\breve{G}_j})}{(n - K - 1) ~ \boldsymbol{\breve{G}_j}^T \boldsymbol{\breve{G}_j}} \approx \dfrac{\boldsymbol{\breve{y}}^T \boldsymbol{\breve{y}}}{n ~ \boldsymbol{\breve{G}_j}^T \boldsymbol{\breve{G}_j}} \approx \dfrac{\text{var}(\boldsymbol{y})}{n ~ \text{var}(\boldsymbol{g_j})} ~.\]
-Thus, we can derive $\text{sd}(\boldsymbol{g_j}) \approx \dfrac{\text{sd}(\boldsymbol{y})}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n}}$ and $\left(\text{sd}(\boldsymbol{g_j}) ~ \hat{\gamma}_j\right) \approx \dfrac{\hat{\gamma}_j}{\text{se}(\hat{\gamma}_j)} \dfrac{\text{sd}(\boldsymbol{y})}{\sqrt{n}}$.
+\[\left(\text{se}(\hat{\gamma}_j)\right)^2 = \dfrac{(\boldsymbol{\breve{y}} - \hat{\gamma}_j \boldsymbol{\breve{G}_j})^T (\boldsymbol{\breve{y}} - \hat{\gamma}_j \boldsymbol{\breve{G}_j})}{(n - K - 1) ~ \boldsymbol{\breve{G}_j}^T \boldsymbol{\breve{G}_j}} \approx \dfrac{\boldsymbol{\breve{y}}^T \boldsymbol{\breve{y}}}{n ~ \boldsymbol{\breve{G}_j}^T \boldsymbol{\breve{G}_j}} \approx \dfrac{\text{var}(\boldsymbol{y})}{n ~ \text{var}(\boldsymbol{G_j})} ~.\]
+The first approximation is possible because $\hat{\gamma}_j$ is expected to be small, while the second approximation assumes that the effects from covariates are small. 
+Thus, we can derive $\text{sd}(\boldsymbol{G_j}) \approx \dfrac{\text{sd}(\boldsymbol{y})}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n}}$ and then $\left(\text{sd}(\boldsymbol{G_j}) ~ \hat{\gamma}_j\right) \approx \dfrac{\hat{\gamma}_j}{\text{se}(\hat{\gamma}_j)} \dfrac{\text{sd}(\boldsymbol{y})}{\sqrt{n}}$.
 Let us go back to equation \ref{eq:joint}. 
@@
 \begin{equation}\label{eq:beta_res}
-\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta}^T_{\text{-}j} \boldsymbol{R}_{\text{-}j, j}
+\tilde{\beta}_j = \hat{\beta}_j - \boldsymbol{\beta_{-j}}^T \boldsymbol{R_{-j, j}}
 \end{equation}
-where $\boldsymbol{R}_{\text{-}j, j}$ is the $j$-th column without the  $j$-th row of the correlation matrix, $\boldsymbol{\hat{\beta}}$ is the vector of marginal effect sizes, and $\boldsymbol{{\beta}}$ is the vector of current effect sizes in the Gibbs sampler.
+where $\boldsymbol{R_{-j, j}}$ is the $j$-th column without the  $j$-th row of the correlation matrix, $\boldsymbol{\hat{\beta}}$ is the vector of marginal effect sizes, $\boldsymbol{{\beta}}$ is the vector of current effect sizes in the Gibbs sampler, and $\boldsymbol{\beta_{-j}}$ is $\boldsymbol{\beta}$ without the $j$-th element.
 Then, the probability that variant $j$ is causal is computed as 
@@
 
-Therefore, the posterior mean of $\beta_j$ is given by 
+Therefore, the posterior mean of $\beta_j ~|~ \tilde{\beta}_j$ is given by 
 \begin{equation}\label{eg:post_beta}
@@
 
-The first extension consists in estimating $p$ and $h^2$ in the model, as opposed to testing several values of $p$ and estimating $h^2$ using constrained LD score regression \cite[]{bulik2015ld}.
+The first extension consists in estimating $p$ and $h^2$ within the model, as opposed to testing several values of $p$ and estimating $h^2$ using constrained LD score regression \cite[]{bulik2015ld}.
 This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly to data without the need of a validation dataset to choose best-performing hyper-parameters.
@@
 Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the estimates of $p$ and $h^2$ from LDpred2-auto.
+
+When running LDpred2-grid, we test a grid of hyper-parameters with $p$ from a sequence of 21 values from $10^{-5}$ to $1$ on a log-scale; $h^2$ within $\{0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$, where $h^2_\text{LDSC}$ is the heritability estimate for a constrained LD score regression \cite{bulik2015ld}; and whether sparsity is enabled or not.
+In total, this grid is of size $21 \times 3 \times 2 = 126$.
+When running LDpred2-auto, we run it 30 times with $h^2_\text{LDSC}$ as initial value for $h^2$ and a sequence of 30 values from $10^{-4}$ to $0.9$ on a log-scale as initial values for $p$.
+Running many Gibbs chains aims at checking that models did not diverge (and hopefully converged).
+As a criterion for non-divergence, we compute the standard deviations of the resulting predictors from the 30 models, and keep only those within 3 median absolute deviations from their median.
 
@@
 
-[TODO: EXPORT CODE FROM CLUSTER]
+%[TODO: EXPORT CODE FROM CLUSTER]
 

### 2020-06-22 paper/paper-ldpred2.tex
save

@@
 LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
-However, LDpred has limitations that may result in limited predictive performance.
+However, LDpred has limitations that may reduce its predictive performance.
 Here we present LDpred2, a new version of LDpred that addresses these issues.
-We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns parameters from data.
+We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns the two LDpred parameters from data.
 We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
-We then show that LDpred2 compares favorably to other polygenic score methods developed more recently.
-LDpred2-auto, which automatically estimates the parameters of heritability and proportion of causal variants, performs particularly well, provided some quality control is performed.
+We then show that LDpred2 compares favorably to other polygenic score methods recently developed.
+LDpred2-auto, which automatically estimates the two LDpred parameters, performs particularly well, provided some quality control is performed on the summary statistics.
 LDpred2 is implemented in R package bigsnpr.
@@
 Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
-PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association}. 
+PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association,barth2020genetic}. 
 For example, a recently derived PGS for education attainment has been one of the most predictive variables in behavioural sciences so far \cite[]{allegrini2019genomic}.
 
-LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a Linkage Disequilibrium (LD) matrix only \cite[]{vilhjalmsson2015modeling}.
-It assumes there is a proportion $p$ of variants that are causal.
-However, LDpred has several limitations that may result in limited predictive performance.
-The non-infinitesimal version of LDpred, a Gibbs sampler, is particularly sensitive to model misspecification when applied to summary statistics with large sample sizes.
+LDpred is a popular method for deriving polygenic scores based on summary statistics and a Linkage Disequilibrium (LD) matrix only \cite[]{vilhjalmsson2015modeling}.
+However, LDpred has several limitations that may reduce its predictive performance.
+The non-infinitesimal version of LDpred, which assumes there is a proportion $p$ of variants that are causal, is a Gibbs sampler and is particularly sensitive to model misspecification when applied to summary statistics with large sample sizes.
 It is also unstable in long-range LD regions such as the human leukocyte antigen (HLA) region of chromosome 6. 
@@
 We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing prediction accuracy.
-Moreover, LDpred2-auto, which does not require any validation set, perform very well, provided some quality control is performed.
+Moreover, LDpred2-auto, which does not require any validation set, performs very well, provided some quality control is performed on the summary statistics.
 
@@
 Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
-LDpred2 has 4 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ and therefore is free of hyper-parameters to tune.
-As a recall, LDpred v1 has two options: LDpred1-grid where only $p$ is optimized, and LDpred1-inf.
+LDpred2 has 4 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ from data and therefore is free of hyper-parameters to tune.
+As a recall, LDpred v1 has two options: LDpred1-inf, and LDpred1-grid where only $p$ is optimized and $h^2$ is estimated from constrained LD score regression \cite[]{bulik2015ld}.
 
@@
 In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
-As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are sampled from a long-range LD region. 
-In contrast, all LDpred2 models perform well in these two HLA scenarios, although LDpred2-auto performs slightly worse than other LDpred2 models.
+As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are located in a long-range LD region. 
+In contrast, all LDpred2 models perform well in these two HLA scenarios, but LDpred2-auto performs slightly worse than other LDpred2 models in these extreme scenarios.
 
@@
 For Asthma and RA, predictive performance of LDpred1 and LDpred2 are similar.
-As in simulations, the sparse version of LDpred2 performs equally well as the original version.
-Moreover, LDpred2-auto also perform  very well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
+As in simulations, the sparse version of LDpred2 performs similarly to the non-sparse version.
+Moreover, LDpred2-auto also performs well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
 When comparing to other PGS methods, we only report the best LDpred2 model (chosen using the validation set) and LDpred2-auto. 
 LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).
-LDpred2-auto usually performs even better than the LDpred-best (Figure \ref{fig:AUC-all}).
+We note that LDpred2-auto usually performs even better than LDpred-best (Figure \ref{fig:AUC-all}).
 This means that choosing the best LDpred model per chromosome leads to overfitting in the validation set, which does not transfer favorably when evaluating prediction performance in the test set.
@@
 
-Here we present LDpred2, a new version of LDpred. 
-LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
-Yet, it has some instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
-We have therefore implemented a new version of LDpred to solve these instability issues and improve its computational efficiency.
-
-We show that LDpred2 is much more stable and provides higher predictive performance than LDpred1.
-We demonstrate that LDpred1 has defects, particularly when handling long-range LD regions such as the HLA region.
-Indeed, LDpred1 performs poorly in the simulations where causal variants are in the HLA region.
-In contrast, LDpred2 performs very well. We hypothesize that LDpred1 does not use a window size that is large enough to account for long-range LD such as in the HLA region.
-In LDpred2, we use a window size of 3 cM, which is much larger than the default value used in LDpred1 and which enables LDpred2 to work well even when causal variants are in long-range LD regions.
+The previous version of LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
+Yet, it has instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
+We have therefore implemented a new version of LDpred, LDpred2, to address these issues and improve its computational efficiency.
+We show that LDpred2 is more stable and provides higher predictive performance than LDpred1, particularly when handling long-range LD regions or less polygenic traits.
+We hypothesize that LDpred1 does not use a LD window size that is large enough to account for long-range LD such as in the HLA region.
+In LDpred2, we use a window size of 3 cM, which is larger than the default value used in LDpred1.
+This enables LDpred2 to work well even when causal variants are located in long-range LD regions.
 %We hypothesize that using a larger LD window explains the 5 real traits where LDpred2 performs much better than LDpred1, especially because LDpred2-inf performs better than LDpred1-inf.
-We strongly discourage against removing these regions as sometimes suggested in the literature. 
+We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved}. 
 Indeed, these regions, especially the HLA region, contain lots of variants associated with many traits, and are therefore very useful for prediction.
 
-In LDpred2, we also test more values for $p$ (21 instead of 7 by default) and for $h^2$ (3 instead of 1).
+In LDpred2, we also expand the grid of hyper-parameters examined with now more values for $p$ (21 instead of 7 by default in LDpred1) and for $h^2$ (3 instead of 1).
 When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see below).
 Overall, we test a grid of 126 different values in LDpred2 instead of 7 in LDpred1.
-We also use a larger window size for computing correlations between variants, yet LDpred2 is still as fast as LDpred1.
-LDpred2 is fast because we provide an efficient parallel implementation in C++.
+We also use a larger window size for computing correlations between variants (see above), yet LDpred2 is still as fast as LDpred1.
+The efficiency of LDpred2 is achieved through an efficient parallel implementation in C++.
 The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-Efficient parallelization is possible because we developed an on-disk sparse matrix format accessed via memory mapping in R package bigsparser. 
-As an example, it takes around two hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
+Efficient parallelization is possible because we developed an on-disk sparse matrix format accessed using memory mapping in R package bigsparser. 
+As an example, it takes around less than three hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
 It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
 
-LDpred2 also comes with two new models.
-When the sparse option is enabled, it provides models that truly encourage sparsity, as compared to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
-It can provide sparsity in effect sizes as much as 98\% for small $p$, while keeping predictive performance as good as non-sparse models, as opposed to if very small effects were simply discarded \cite[]{bolli2019software}.
-This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs better or equally well as the non-sparse model.
-As for LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, it performs equally well as other LDpred2 models in simulations, and also performs very well in real data applications, provided that some quality control is performed on summary statistics (see Supplementary Note).
-This quality control aims at ensuring that effects are transferable from the summary statistics to the data where PGS are computed.
+LDpred2 also extends the original LDpred model in two ways.
+First, we provide a sparse option in LDpred2-grid which provides models that truly encourage sparsity, in contrast to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
+In practice, the sparse version od LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as shown in \cite{bolli2019software}).
+%This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs better or equally well as the non-sparse model.
+The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, therefore it does not require a validation step to choose hyper-parameters, making it an attractive option in many applications.
+LDpred2-auto performs equally well as other LDpred2 models in simulations.
+It also performs very well in real data applications, provided that some quality control is performed on summary statistics (see Supplementary Note).
+This quality control aims at ensuring that effects are transferable from the external GWAS summary statistics to the data where PGS are computed.
 
 In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
-When comparing to recently derived methods that showed higher predictive performance compared to the previous version of LDpred, LDpred2 performs favorably. 
-Indeed, LDpred2 (especially LDpred2-auto) provides the highest predictive performance for almost all the real traits tested, and very good predictive performance for the others.
-Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
+LDpred2 also performs favorably when comparing it to recently derived methods that showed higher predictive performance compared to the previous version of LDpred. 
+Indeed, LDpred2 (especially LDpred2-auto) provides the highest predictive performance for almost all the real traits tested here, and very good predictive performance for the others.
+Note that the comparison may not be totally fair for C+T and SCT, as we use the same set of variants for all methods. 
 Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.
 Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, or detrimental as it may break some of the assumptions behind their respective models.
-For now, we recommend using the same variants used in PRS-CS and used here.
-For example, we do not think using several million variants would improve predictive performance of LDpred2.
-Investigating selection of variants is a direction of future research for us.
+For now, we recommend using the same HapMap3 variants used in PRS-CS and used here.
+Investigating variant selection for LDpred2 and other methods is a direction of future research for us.
 
@@
 We use the UK Biobank data for both real data analyses and simulations \cite[]{bycroft2018uk}. 
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have passed some quality control.
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have passed some quality control \cite[]{bycroft2018uk}.
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
@@
 This results in 362,320 individuals and 1,117,493 variants.
-We use 10,000 individuals as a validation set for choosing optimal hyper-parameters and for computing correlations between variants (LD matrix $\boldsymbol{R}$). 
+We use 10,000 individuals as validation set for choosing optimal hyper-parameters and for computing correlations between variants (LD matrix $\boldsymbol{R}$). 
 We use 300,000 other individuals for running logistic GWAS to create summary statistics.
@@
 
-We simulate binary phenotypes with a heritability of $h^2 = 0.4$ (or 0.3)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}. 
+We simulate binary phenotypes with a heritability of $h^2=0.4$ (or 0.3)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}. 
 We vary the number of causal variants (300, 3000, 30,000, or 300,000) to match a range of genetic architectures from low to high polygenicity.
@@
 In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
-Then, we compare the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
+Then, we compare the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) and LDpred2-auto to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
 For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}.
@@
 Then, the effect sizes that LDpred outputs need to be scaled back by multiplying by $\left(\text{se}(\hat{\gamma}_j) ~ \sqrt{n}\right)$ (last line of algorithm \ref{algo:ldpred}).
+In LDpred2, we allow for having different $n_j$ for different variants. 
 Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{var}(\boldsymbol{y}) = 1$ only.
@@
 where $p$ is the proportion of causal variants, $M$ the number of variants and $h^2$ the (SNP) heritability.
-\cite{vilhjalmsson2015modeling} estimates $h^2$ using constrained LD score regression (intercept fixed to 1) and recommend testing a grid of hyper-parameter values for $p$ (1, 0.3, 0.1, 0.03, 0.01, 0.003 and 0.001).
+\cite{vilhjalmsson2015modeling} estimate $h^2$ using constrained LD score regression (intercept fixed to 1) and recommend testing a grid of hyper-parameter values for $p$ (1, 0.3, 0.1, 0.03, 0.01, 0.003 and 0.001).
 
@@
 
-When running LDpred2-grid, we test a grid of hyper-parameters with $p$ from a sequence of 21 values from $10^{-5}$ to $1$ on a log-scale; $h^2$ within $\{0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$, where $h^2_\text{LDSC}$ is the heritability estimate for a constrained LD score regression \cite{bulik2015ld}; and whether sparsity is enabled or not.
+When running LDpred2-grid, we test a grid of hyper-parameters with $p$ from a sequence of 21 values from $10^{-5}$ to $1$ on a log-scale; $h^2$ within $\{0.7, 1, 1.4\} \cdot h^2_\text{LDSC}$, where $h^2_\text{LDSC}$ is the heritability estimate from the constrained LD score regression \cite{bulik2015ld}; and whether sparsity is enabled or not.
 In total, this grid is of size $21 \times 3 \times 2 = 126$.
@@
 Running many Gibbs chains aims at checking that models did not diverge (and hopefully converged).
-As a criterion for non-divergence, we compute the standard deviations of the resulting predictors from the 30 models, and keep only those within 3 median absolute deviations from their median.
+As a criterion for non-divergence, we compute the standard deviations of the resulting predictors from the 30 models, keep only those within 3 median absolute deviations from their median, and average the remaining vectors of effects as final effect sizes for the ``auto'' version.
 
@@
 There is a window size parameter that needs to be set in LDpred; for a given variant, correlations with other variants outside of this window are assumed to be 0.
-The recommended value for this window (in number of variants) is to use the total number of variants divided by 3000, which corresponds to a window radius of around 2 Mb \cite[]{vilhjalmsson2015modeling}.
+The recommended value for this window (in number of variants) has been to use the total number of variants divided by 3000, which corresponds to a window radius of around 2 Mb \cite[]{vilhjalmsson2015modeling}.
 We have come to the conclusion that this window size is not large enough. Indeed, the human leukocyte antigen (HLA) region of chromosome 6 is 8 Mb long \cite[]{price2008long}. 
@@
 We run LDpred2 per chromosome.
-Even if it is possible to run LDpred2 genome-wide, this approach has two limitations.
+Even if it is possible to run LDpred2 genome-wide, this approach has two drawbacks.
 First, it can be memory and computationally demanding to do so.

### 2020-06-25 code/plot-approx-sd.R
add code and figures

@@ -1,0 +1,32 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+
+sd <- readRDS("data/sd.rds")
+
+#### Simulations ####
+
+sumstats_simu <- do.call("rbind", lapply(1:22, function(chr) {
+  readRDS(paste0("data/sumstats-simu/all_40_300_15_1_chr", chr, ".rds"))
+}))
+y <- readRDS("data/pheno-simu/all_40_300_15_1.rds")
+Neff <- 4 / (1 / sum(y[ind.gwas] == 0) + 1 / sum(y[ind.gwas] == 1))
+
+sd_val <- sd
+sd_ss <- 2 / sqrt(Neff * sumstats_simu$std.err^2)
+
+qplot(sd_val, sd_ss, alpha = I(0.5)) +
+  theme_bigstatsr() +
+  coord_equal() +
+  geom_abline(linetype = 2, color = "red") +
+  labs(x = "Standard deviations in the validation set",
+       y = "Standard deviations derived from the summary statistics")
+
+# ggsave("figures/sd-approx-simu.png", width = 7, height = 7)
+
+## See prepare-sumstats.R for BRCA

### 2020-06-25 code/plot-res-all.R

@@ -1,0 +1,121 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+library(tidyverse)
+library(future)
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+plan(multisession, workers = NCORES)
+
+pred_ldpred2 <-
+  tibble(res_file = list.files("results/ldpred2", "\\.rds$", full.names = TRUE)) %>%
+  mutate(Trait = sub("(.*/)([[:alnum:]]+)(_chr[0-9]+\\.rds)$", "\\2", res_file),
+         chr = as.integer(sub("(.*_chr)([0-9]+)(\\.rds)$", "\\2", res_file)),
+         pred = furrr::future_map(res_file, readRDS), res_file = NULL) %>%
+  unnest_longer(col = "pred", indices_to = "Method") %>%
+  filter(Method %in% c("best_beta", "beta_auto")) %>%
+  mutate(Method = case_when(Method == "beta_auto" ~ "LDpred2_auto",
+                            Method == "best_beta" ~ "LDpred2_best")) %>%
+  group_by_at(-(2:3)) %>%
+  summarise(pred = list(do.call(plus, map(pred, replace_na, replace = 0)))) %>%
+  ungroup() %>%
+  print()
+
+pred_sct <-
+  tibble(res_file = list.files("results/sct", "\\.rds$", full.names = TRUE)) %>%
+  mutate(Trait = sub("\\.rds$", "", basename(res_file)),
+         pred = furrr::future_map(res_file, readRDS), res_file = NULL) %>%
+  unnest_longer(col = "pred", indices_to = "Method") %>%
+  print()
+
+pred_lassosum <-
+  tibble(Method = "lassosum",
[…]

### 2020-06-25 code/plot-res-ldpred.R

@@ -1,0 +1,115 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+library(tidyverse)
+library(future)
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+plan(multisession, workers = NCORES)
+
+pred_ldpred2 <-
+  tibble(res_file = list.files("results/ldpred2", "\\.rds$", full.names = TRUE)) %>%
+  mutate(Trait = sub("(.*/)([[:alnum:]]+)(_chr[0-9]+\\.rds)$", "\\2", res_file),
+         chr = as.integer(sub("(.*_chr)([0-9]+)(\\.rds)$", "\\2", res_file)),
+         pred = furrr::future_map(res_file, readRDS), res_file = NULL) %>%
+  unnest_longer(col = "pred", indices_to = "Method") %>%
+  print()
+
+AUCBoot_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUCBoot(pred[is_not_na], target[is_not_na])
+}
+
+res_ldpred2 <- pred_ldpred2 %>%
+  group_by_at(-(2:3)) %>%
+  summarise(pred = list(do.call(plus, map(pred, replace_na, replace = 0)))) %>%
+  ungroup() %>%
+  mutate(Method = sub("beta", "ldpred2", Method),
+         AUCBoot = furrr::future_map2(pred, Trait, ~ {
+           pheno <- readRDS(paste0("data/pheno/", .y, ".rds"))
+           AUCBoot_no_NA(.x, pheno[ind.test])
+         }),
+         pred = NULL) %>%
+  print()
+
+auc_real <- res_ldpred2 %>%
+  filter(Method != "best_ldpred2") %>%
+  bind_rows(readRDS("results/all-ldpred1.rds")) %>%
+  mutate(Method = factor(Method, levels = c(
[…]

### 2020-06-25 code/plot-res-simus-ldpred.R

@@ -1,0 +1,113 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+
+library(tidyverse)
+library(foreach)
+pred_ldpred2 <- tibble(basename = list.files("data/pheno-simu")) %>%
+  mutate(
+    pheno_file = file.path("data/pheno-simu", basename),
+    res_file = map(file.path("results/simu-ldpred2", basename), ~ {
+      foreach(chr = 1:22, .combine = 'c') %do% {
+        sub("\\.rds$", paste0("_chr", chr, ".rds"), .x)
+      }
+    }))
+
+res_ldpred2 <- pred_ldpred2 %>%
+  mutate(basename = gsubfn::gsubfn("(.+)_([0-9]+)\\.rds$", "\\1 \\2", basename)) %>%
+  separate("basename", c("params", "iter"), sep = " ", convert = TRUE) %>%
+  unnest_longer(col = "res_file", indices_to = "chr") %>%
+  filter(file.exists(res_file)) %>%
+  mutate(pred = map(res_file, readRDS), res_file = NULL) %>%
+  unnest_longer(col = "pred", indices_to = "Method") %>%
+  group_by_at(-(4:5)) %>%
+  summarise(pred = list(do.call(plus, pred))) %>%
+  ungroup() %>%
+  mutate(Method = sub("beta", "ldpred2", Method),
+         pheno = map(pheno_file, readRDS), pheno_file = NULL,
+         AUC = map2_dbl(pred, pheno, ~ AUC(.x, .y[ind.test])),
+         pred = NULL, pheno = NULL) %>%
+  print()
+
+auc_simu <- res_ldpred2 %>%
+  bind_rows(readRDS("results/all-simu-ldpred1.rds")) %>%
+  group_by(params, Method) %>%
+  summarise(aucs = {
[…]

### 2020-06-25 code/prepare-corr-spmat.R

@@ -1,0 +1,51 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+POS <- ukb$map$physical.pos
+
+set.seed(1); ind.val <- sort(sample(nrow(G), 10e3))
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+POS2 <- snp_asGeneticPos(CHR, POS, dir = "tmp-data", ncores = NCORES)
+
+bigassertr::assert_dir("data/corr")
+
+for (chr in 1:22) {
+
+  ind.chr <- which(CHR == chr)
+
+  runonce::save_run(
+    snp_cor(
+      G, ind.row = ind.val, ind.col = ind.chr,
+      alpha = 1, infos.pos = POS2[ind.chr], size = 3 / 1000,
+      ncores = NCORES
+    ),
+    file = paste0("data/corr/chr", chr, ".rds")
+  )
+}
+## On a node with 16 cores and 128 GB RAM:
+#     user   system  elapsed
+# 6737.315   31.980  650.724
+# 6929.821   40.563  665.379
+# 7321.282   39.027  644.371
+# 5373.981   32.860  486.469
+# 5638.935 1368.178  632.483
+# 7840.772   52.543  693.595
+# 4477.675   34.834  424.483
+# 4875.977   37.007  448.571
+# 3611.675   26.642  325.730
+# 4434.387  542.683  445.706
+# 4250.445   32.365  383.651
+# 4057.731  801.579  424.581
+# 3038.082   19.844  265.982
+# 2445.123   17.932  220.284
+# 2182.027   15.725  193.493
+# 1982.660   16.003  182.821
+# 1632.852   15.099  147.933
+# 2177.030  372.233  216.424
+#  857.160    7.769   80.343
+# 1803.857   13.320  154.808
+#  788.966    6.565   70.550
+#  861.304    6.443   72.679
+# -> 2.2H in total

### 2020-06-25 code/prepare-genotypes.R

@@ -1,0 +1,71 @@
+# file.symlink("../../UKBB/", "UKBB")
+
+map_hapmap3 <- bigreadr::fread2("ldblk_1kg_eur/snpinfo_1kg_hm3")
+
+library(doParallel)
+cl <- makeCluster(22)
+parallel::clusterExport(cl, "map_hapmap3")
+list_snp_id <- parLapply(cl, 1:22, function(chr) {
+  mfi <- paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt")
+  infos_chr <- bigreadr::fread2(mfi, showProgress = FALSE)
+  joined <- dplyr::inner_join(cbind(chr = chr, infos_chr), map_hapmap3[1:2],
+                              by = c("chr" = "CHR", "V2" = "SNP"))
+  with(joined[!vctrs::vec_duplicate_detect(joined$V2), ],
+       paste(chr, V3, V4, V5, sep = "_"))
+})
+stopCluster(cl)
+
+sum(lengths(list_snp_id))  # 1,117,493
+
+
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")
+str(sample)
+sample <- sample[-1, ]
+
+
+csv <- "UKBB/ukb41181.csv"
+df0 <- bigreadr::fread2(
+  csv,
+  select = c("eid", "22020-0.0", paste0("22009-0.", 1:16)),
+  col.names = c("eid", "used_in_pca", paste0("PC", 1:16))
+)
+
+# not removed & quality controled / unrelated & white-British
+ind.indiv <- match(df0$eid, sample$ID_2)
+sub <- which(!is.na(ind.indiv) & df0$used_in_pca)
+
+# Genetically homogeneous
+dist <- bigutilsr::dist_ogk(as.matrix(df0[sub, -(1:2)]))
+sub2 <- sub[log(dist) < 5]
+length(sub2) # 362,320
+saveRDS(sub2, "data/ind_sub_csv.rds")
+
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+
+system.time(
+  rds <- bigsnpr::snp_readBGEN(
+    bgenfiles   = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = 1:22),
+    list_snp_id = list_snp_id,
+    backingfile = "data/UKBB_imp_HM3",
[…]

### 2020-06-25 code/prepare-phenotypes.R

@@ -1,0 +1,196 @@
+library(tidyverse)
+library(bigreadr)
+
+bigassertr::assert_dir("data/pheno")
+
+csv <- "UKBB/ukb41181.csv"
+sub <- readRDS("data/ind_sub_csv.rds")
+
+# Cancers
+df_cancer0 <- fread2(csv, select = paste0("2453-", 0:2, ".0"))
+df_cancer1 <- fread2(csv, select = c(paste0("20001-0.", 0:5),
+                                     paste0("20001-1.", 0:5),
+                                     paste0("20001-2.", 0:5)))
+df_cancer2 <- fread2(csv, select = c(paste0("40001-", 0:2, ".0"),
+                                     paste0("40002-0.", 0:13),
+                                     paste0("40002-1.", 0:13),
+                                     paste0("40002-2.", 0:13),
+                                     paste0("40006-", 0:31, ".0"),
+                                     paste0("41202-0.", 0:379),
+                                     paste0("41204-0.", 0:434)))
+
+# Other illnesses
+df_illness <- fread2(csv, select = c(paste0("20002-0.", 0:28),
+                                     paste0("20002-1.", 0:28),
+                                     paste0("20002-2.", 0:28)))
+df_ICD10 <- fread2(csv, select = c(paste0("40001-", 0:2, ".0"),
+                                   paste0("40002-0.", 0:13),
+                                   paste0("40002-1.", 0:13),
+                                   paste0("40002-2.", 0:13),
+                                   paste0("41202-0.", 0:379),
+                                   paste0("41204-0.", 0:434)))
+
+#### Breast cancer (BRCA) ####
+
+ind_BRCA <- sort(unique(unlist(c(
+  lapply(df_cancer1, function(x) which(x == 1002)),
+  lapply(df_cancer2, function(x) which(substr(x, 1, 3) %in% c("C50", "D05")))
+))))
+
+y <- rep(NA, nrow(df_cancer0))
+y[rowMeans(df_cancer0 == 0, na.rm = TRUE) == 1] <- 0
[…]

### 2020-06-25 code/prepare-simu-phenotypes.R

@@ -1,0 +1,53 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+POS <- ukb$map$physical.pos
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+
+bigassertr::assert_dir("data/pheno-simu")
+
+#### Anywhere on the genome ####
+
+params <- expand.grid(
+  h2   = 0.4,
+  M    = c(300, 3e3, 30e3, 300e3),
+  K    = 0.15,
+  iter = 1:10
+)
+
+purrr::pwalk(params, function(h2, M, K, iter) {
+
+  pheno_file <- sprintf("data/pheno-simu/all_%d_%d_%d_%d.rds",
+                        100 * h2, M, 100 * K, iter)
+
+  if (!file.exists(pheno_file)) {
+    print(pheno_file)
+    simu <- snp_simuPheno(G, h2, M, K, ncores = NCORES)
+    saveRDS(simu$pheno, pheno_file)
+  }
+})
+
+
+#### In HLA ####
+
+ind.HLA <- snp_indLRLDR(CHR, POS, LD.wiki34[12, ])
+
+params2 <- expand.grid(
+  h2   = 0.3,
+  M    = c(300, 3000),
+  K    = 0.15,
+  iter = 1:10
+)
+
+purrr::pwalk(params2, function(h2, M, K, iter) {
+
+  pheno_file <- sprintf("data/pheno-simu/HLA_%d_%d_%d_%d.rds",
+                        100 * h2, M, 100 * K, iter)
+
+  if (!file.exists(pheno_file)) {
+    print(pheno_file)
+    simu <- snp_simuPheno(G, h2, M, K, ind.possible = ind.HLA, ncores = NCORES)
+    saveRDS(simu$pheno, pheno_file)
+  }
+})

### 2020-06-25 code/prepare-simu-sumstats.R

@@ -1,0 +1,39 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+POS <- ukb$map$physical.pos
+
+bigassertr::assert_dir("data/sumstats-simu")
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+
+pheno_files <- list.files("data/pheno-simu")
+all_pheno <- lapply(pheno_files, function(file) {
+  readRDS(file.path("data/pheno-simu", file))
+})
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = "16", mem = "128g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+future.apply::future_lapply(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+
+  purrr::walk2(pheno_files, all_pheno, ~ {
+
+    gwas_file <- file.path("data/sumstats-simu",
+                           sub("\\.rds$", paste0("_chr", chr, ".rds"), .x))
+
+    if (!file.exists(gwas_file)) {
+      print(gwas_file)
+      gwas <- big_univLogReg(G, .y[ind.gwas], ind.train = ind.gwas,
+                             ind.col = ind.chr, ncores = 16)
+      saveRDS(gwas, gwas_file)
+    }
+  })
+})

### 2020-06-25 code/prepare-sumstats.R

@@ -1,0 +1,359 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+map <- transmute(ukb$map,
+                 chr = as.integer(chromosome), pos = physical.pos,
+                 a0 = allele1, a1 = allele2)  # reversed somehow..
+
+bigassertr::assert_dir("data/sumstats")
+
+set.seed(1); ind.val <- sort(sample(nrow(G), 10e3))
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+sd <- runonce::save_run(
+  big_parallelize(G, function(X, ind, ind.val) {
+    bigstatsr::big_scale()(X, ind.val, ind)$scale
+  }, p.combine = "c", ncores = NCORES, ind.val = ind.val),
+  file = "data/sd.rds"
+)
+
+#### Breast cancer (BRCA) ####
+
+# download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/MichailidouK_29059683_GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
+#               destfile = "tmp-data/sumstats_BRCA.txt.gz")
+# R.utils::gunzip("tmp-data/sumstats_BRCA.txt.gz")
+sumstats <- fread2("tmp-data/sumstats_BRCA.txt", na.strings = "NULL",
+                   select = c("chr", "position_b37", "a0", "a1",
+                              "bcac_onco_icogs_gwas_eaf_controls",
+                              "bcac_onco_icogs_gwas_beta",
+                              "bcac_onco_icogs_gwas_se",
+                              "bcac_onco_icogs_gwas_P1df"),
+                   col.names = c("chr", "pos", "a0", "a1", "freq",
+                                 "beta", "beta_se", "p"))
+sumstats$n_eff <- 4 / (1 / 137045 + 1 / 119078)
+
+info_snp <- snp_match(sumstats, map)
+# 11,792,542 variants to be matched.
+# 1,633,946 ambiguous SNPs have been removed.
[…]

### 2020-06-25 code/process-betas-ldpred1.R

@@ -1,0 +1,54 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+library(tidyverse)
+pred_ldpred1 <-
+  tibble(res_file = list.files("results/ldpred1", full.names = TRUE),
+         betas = map(res_file, ~ as.list(setNames(as_tibble(readRDS(.x)),
+                                                  c(rep("grid", 7), "inf"))))) %>%
+  unnest_longer("betas", indices_to = "method") %>%
+  print()
+
+str(all_betas <- do.call("cbind", pred_ldpred1$betas))
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+bigparallelr::set_blas_ncores(NCORES)
+pred <- big_prodMat(G, all_betas)
+
+pred_ldpred1$pred <- as.list(as.data.frame(pred))
+pred_ldpred1$pheno <- map(pred_ldpred1$res_file, ~ {
+  readRDS(sub("results/ldpred1", "data/pheno", .x, fixed = TRUE))
+})
+
+AUC_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUC(pred[is_not_na], target[is_not_na])
+}
+AUCBoot_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUCBoot(pred[is_not_na], target[is_not_na])
+}
+
+library(future)
+plan(multisession, workers = NCORES)
+res_ldpred1 <- pred_ldpred1 %>%
+  mutate(auc_val = map2_dbl(pred, pheno, ~ AUC_no_NA(-.x[ind.val], .y[ind.val]))) %>%
+  group_by(res_file, method) %>%
+  arrange(desc(auc_val)) %>%
+  slice(1) %>%
+  ungroup() %>%
+  mutate(Trait = gsubfn::strapply(
+    res_file, "results/ldpred1/(.+)\\.rds$", simplify = 'c')) %>%
+  transmute(Trait,
+            Method = ifelse(method == "grid", "best_ldpred1_grid", "ldpred1_inf"),
+            AUCBoot = furrr::future_map2(pred, pheno, ~ {
[…]

### 2020-06-25 code/process-betas-prscs.R

@@ -1,0 +1,59 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+library(tidyverse)
+beta_prscs <-
+  tibble(res_file = list.files("results/prscs", "\\.txt$", full.names = TRUE)) %>%
+  mutate(Trait = sub("(.+/)([[:alnum:]]+)(.+_chr[0-9]+\\.txt)$", "\\2", res_file),
+         chr = as.integer(sub("(.+_chr)([0-9]+)(\\.txt)$", "\\2", res_file)),
+         phi = as.double(sub("(.+_phi)(.+)(_chr[0-9]+\\.txt)$", "\\2", res_file))) %>%
+  group_by(Trait, phi) %>%
+  arrange(chr) %>%
+  summarize(beta = list(bigreadr::fread2(res_file))) %>%
+  print()
+
+betas <- matrix(0, ncol(G), nrow(beta_prscs))
+for (k in cols_along(betas)) {
+  tmp <- beta_prscs$beta[[k]]
+  betas[match(tmp$V2, ukb$map$rsid), k] <- tmp$V6
+}
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+bigparallelr::set_blas_ncores(NCORES)
+pred <- big_prodMat(G, betas)
+
+beta_prscs$pred <- as.list(as.data.frame(pred))
+beta_prscs$pheno <- map(beta_prscs$Trait, ~ {
+  readRDS(paste0("data/pheno/", ., ".rds"))
+})
+
+AUC_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUC(pred[is_not_na], target[is_not_na])
+}
+AUCBoot_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUCBoot(pred[is_not_na], target[is_not_na])
+}
+
+library(future)
+plan(multisession, workers = NCORES)
+res_prscs <- beta_prscs %>%
+  mutate(auc_val = map2_dbl(pred, pheno, ~ AUC_no_NA(-.x[ind.val], .y[ind.val]))) %>%
+  group_by(Trait) %>%
+  arrange(desc(auc_val)) %>%
[…]

### 2020-06-25 code/process-betas-simu-ldpred1.R

@@ -1,0 +1,41 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+
+library(tidyverse)
+pred_ldpred1 <-
+  tibble(res_file = list.files("results/simu-ldpred1", full.names = TRUE),
+       betas = map(res_file, ~ as.list(setNames(as_tibble(readRDS(.x)),
+                                                c(rep("grid", 7), "inf"))))) %>%
+  unnest_longer("betas", indices_to = "method") %>%
+  print()
+
+str(all_betas <- do.call("cbind", pred_ldpred1$betas))
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+bigparallelr::set_blas_ncores(NCORES)
+pred <- big_prodMat(G, all_betas)
+
+pred_ldpred1$pred <- as.list(as.data.frame(pred))
+pred_ldpred1$pheno <- map(pred_ldpred1$res_file, ~ {
+  readRDS(sub("results/simu-ldpred1", "data/pheno-simu", .x, fixed = TRUE))
+})
+res_ldpred1 <- pred_ldpred1 %>%
+  mutate(auc_val = map2_dbl(pred, pheno, ~ AUC(-.x[ind.val], .y[ind.val]))) %>%
+  group_by(res_file, method) %>%
+  arrange(desc(auc_val)) %>%
+  slice(1) %>%
+  ungroup() %>%
+  mutate(basename = gsubfn::gsubfn("(.+)_([0-9]+)\\.rds$", "\\1 \\2", basename(res_file))) %>%
+  separate("basename", c("params", "iter"), sep = " ", convert = TRUE) %>%
+  transmute(params, iter,
+            Method = ifelse(method == "grid", "best_ldpred1_grid", "ldpred1_inf"),
+            AUC = map2_dbl(pred, pheno, ~ AUC(-.x[ind.test], .y[ind.test]))) %>%
+  print()
+
+saveRDS(res_ldpred1, "results/all-simu-ldpred1.rds")

### 2020-06-25 code/run-lassosum.R

@@ -1,0 +1,62 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+
+library(dplyr)
+files <- tibble(basename = list.files("data/sumstats"),
+                res_file = file.path("results/lassosum", basename))
+bigassertr::assert_dir("results/lassosum")
+
+files_sub <- files %>%
+  filter(!file.exists(res_file)) %>%
+  print()
+
+AUC_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUC(pred[is_not_na], target[is_not_na])
+}
+
+rsid <- bigreadr::fread2("data/UKBB_imp_HM3_val.bim")[[2]]
+
+library(future.batchtools)
+NCORES <- 16
+plan(batchtools_slurm(resources = list(
+  t = "4:00:00", c = NCORES, mem = "128g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub, function(basename, res_file) {
+
+  y <- readRDS(file.path("data/pheno", basename))
+  sumstats <- readRDS(file.path("data/sumstats", basename))
+
+  library(lassosum)
+  doParallel::registerDoParallel(cl <- parallel::makeCluster(NCORES))
+  system.time(
+    out <- lassosum.pipeline(
+      cor = with(sumstats, beta / beta_se / sqrt(n_eff)),
+      snp = rsid[sumstats$`_NUM_ID_`],
+      A1 = sumstats$a1,
+      A2 = sumstats$a0,
+      exclude.ambiguous = FALSE,
+      test.bfile = "data/UKBB_imp_HM3_val",
+      LDblocks = "EUR.hg19",
+      cluster = cl,
+      destandardize = TRUE
+    )
+  ) # 25 min
+  parallel::stopCluster(cl)
+
+  auc.val <- apply(-do.call("cbind", out$pgs), 2, AUC_no_NA, target = y[ind.val])
+  beta_lassosum <- do.call("cbind", out$beta)[, which.max(auc.val)]
+
+  ind <- which(beta_lassosum != 0)
+  pred <- big_prodVec(G, beta_lassosum[ind], ind.row = ind.test,
[…]

### 2020-06-25 code/run-ldpred1.R

@@ -1,0 +1,73 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+ukb$map$chromosome <- as.integer(ukb$map$chromosome)
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+
+library(dplyr)
+files <- tibble(basename = list.files("data/sumstats"),
+                res_file = file.path("results/ldpred1", basename))
+bigassertr::assert_dir("results/ldpred1")
+
+files_sub <- files %>%
+  filter(!file.exists(res_file)) %>%
+  print()
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "60:00:00", c = "4", mem = "32g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub, function(basename, res_file) {
+
+  y <- readRDS(file.path("data/pheno", basename))
+  sumstats <- readRDS(file.path("data/sumstats", basename))
+
+  tmp <- tempfile(tmpdir = "tmp-data")
+  file_sumstats <- paste0(tmp, ".txt")
+  cbind(ukb$map[sumstats$`_NUM_ID_`, ], beta = sumstats$beta, pval = sumstats$p) %>%
+    bigreadr::fwrite2(file_sumstats, sep = "\t")
+  # readLines(file_sumstats, n = 5)
+  (Neff <- round(median(sumstats$n_eff)))
+
+  file_hdf5 <- paste0(tmp, ".hdf5")
+  ldpred <- "ldpred/LDpred.py"
+  # system(glue::glue("python3 {ldpred} coord --help"))
+  system(glue::glue(
+    "python3 {ldpred} coord",
+    " --gf data/UKBB_imp_HM3_val",
+    " --ssf {file_sumstats}",
+    " --skip-coordination",
+    " --rs marker.ID --A1 allele1 --A2 allele2 --pos physical.pos --chr chromosome",
+    " --pval pval --eff beta --beta",
+    " --N {Neff}",
+    " --out {file_hdf5}"
+  ))
+
+  # system(glue::glue("python3 {ldpred} gibbs --help"))
+  system(glue::glue(
[…]

### 2020-06-25 code/run-ldpred2.R

@@ -1,0 +1,127 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+
+library(dplyr)
+files <- tidyr::expand_grid(
+  basename = list.files("data/sumstats"),
+  chr = 1:22
+) %>%
+  mutate(
+    res_file = file.path("results/ldpred2", stringr::str_replace(
+      basename, "\\.rds$", paste0("_chr", chr, ".rds"))),
+    pheno_file = file.path("data/pheno",  basename),
+    gwas_file = file.path("data/sumstats", basename)) %>%
+  print()
+table(files$chr)
+all(file.exists(files$pheno_file)) & all(file.exists(files$gwas_file))
+bigassertr::assert_dir("results/ldpred2")
+
+files_sub <- files %>%
+  arrange(chr) %>%
+  filter(
+    # grepl("BRCA", basename), #trait
+    !file.exists(res_file)) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 8
+plan(batchtools_slurm(resources = list(
+  t = "6:00:00", c = NCORES, mem = "64g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub[-1], function(res_file, chr, pheno_file, gwas_file) {
+
+  y <- readRDS(pheno_file)
+  sumstats <- readRDS(gwas_file)
+
+  ind.val2 <- ind.val[!is.na(y[ind.val])]
+
+  ## indices in 'sumstats'
+  ind.chr <- which(sumstats$chr == chr)
+  (df_beta <- sumstats[ind.chr, c("beta", "beta_se", "n_eff")])
+  ## indices in 'G'
+  ind.chr2 <- sumstats$`_NUM_ID_`[ind.chr]
+  ## indices in 'corr'
+  ind.chr3 <- match(ind.chr2, which(CHR == chr))
+  corr <- readRDS(paste0("data/corr/chr", chr, ".rds"))[ind.chr3, ind.chr3]
+
+  (ldsc <- snp_ldsc2(corr, df_beta))
+  h2_est <- ldsc[["h2"]]
+
[…]

### 2020-06-25 code/run-prscs.R

@@ -1,0 +1,70 @@
+## reference data
+# download.file("https://www.dropbox.com/s/p9aqanhxvxaqv8k/ldblk_1kg_eur.tar.gz?dl=1",
+#               destfile = "data/ldblk_1kg_eur.tar.gz")
+# untar("data/ldblk_1kg_eur.tar.gz")
+
+## software
+# system("git clone https://github.com/getian107/PRScs.git")
+# system("python --version")  # 3.7.4
+
+library(dplyr)
+files <- tidyr::expand_grid(
+  basename = list.files("data/sumstats"),
+  chr = 1:22,
+  phi = 10^-(6:0)
+) %>%
+  mutate(
+    res_file = sprintf("results/prscs/%s_pst_eff_a1_b0.5_phi%.e_chr%d.txt",
+                       sub("\\.rds$", "", basename), phi, chr),
+    gwas_file = file.path("data/sumstats", basename)) %>%
+  print()
+
+bigassertr::assert_dir("results/prscs")
+
+files_sub <- files %>%
+  arrange(basename, chr) %>%
+  filter(
+    # grepl("Asthma", basename), #trait
+    !file.exists(res_file)) %>%
+  select(-basename) %>%
+  print()
+
+
+rsid <- bigreadr::fread2("data/UKBB_imp_HM3_val.bim")[[2]]
+
+library(future.batchtools)
+NCORES <- 8
+plan(batchtools_slurm(resources = list(
+  t = "4:00:00", c = NCORES, mem = "64g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+
+furrr::future_pmap(files_sub, function(chr, phi, res_file, gwas_file) {
+
+  tmp <- tempfile(tmpdir = "tmp-data", fileext = ".txt")
+  library(dplyr)
+  sumstats <- readRDS(gwas_file)
+  sumstats %>%
+    mutate(SNP = rsid[`_NUM_ID_`]) %>%
+    select(SNP, A1 = a1, A2 = a0, BETA = beta, P = p) %>%
+    bigreadr::fwrite2(tmp, sep = "\t") %>%
+    readLines(n = 5)
+
+  prscs <- "PRScs/PRScs.py"
+  library(glue)
+  # system(glue("{prscs} --help"))
+
+  system(glue(
+    "OMP_NUM_THREADS=", NCORES,
+    " python {prscs}",
[…]

### 2020-06-25 code/run-sct.R

@@ -1,0 +1,101 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+POS <- ukb$map$physical.pos
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.test <- setdiff(rows_along(G), ind.val)
+
+
+library(dplyr)
+files <- tibble(basename = list.files("data/sumstats"),
+                res_file = file.path("results/sct", basename))
+bigassertr::assert_dir("results/sct")
+
+files_sub <- files %>%
+  filter(!file.exists(res_file)) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 16
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES, mem = "128g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub, function(basename, res_file) {
+
+  # basename <- "CAD.rds"
+  y <- readRDS(file.path("data/pheno", basename))
+  sumstats <- readRDS(file.path("data/sumstats", basename))
+
+  ind.val2 <- ind.val[!is.na(y[ind.val])]
+
+  ### Clumping
+
+  chi2 <- with(sumstats, (beta / beta_se)^2)
+  lpval <- rep(NA, ncol(G))
+  lpval[sumstats$`_NUM_ID_`] <-
+    -pchisq(chi2, df = 1, lower.tail = FALSE, log.p = TRUE) / log(10)
+
+  system.time(
+    all_keep <- snp_grid_clumping(G, CHR, POS, lpS = lpval, ind.row = ind.val,
+                                  exclude = which(is.na(lpval)), ncores = NCORES)
+  ) # 3420 sec
+
+  beta <- rep(0, ncol(G))
+  beta[sumstats$`_NUM_ID_`] <- sumstats$beta
+
+  ### Thresholding
+
+  system.time(
+    multi_PRS <- snp_grid_PRS(
+      G, all_keep, betas = beta, lpS = lpval, ind.row = ind.val2,
+      grid.lpS.thr = seq_log(0.1, 0.9999 * max(lpval, na.rm = TRUE), 50),
+      backingfile = sprintf("tmp-data/%s_scores", sub("\\.rds$", "", basename)),
+      ncores = NCORES
[…]

### 2020-06-25 code/run-simu-ldpred1.R

@@ -1,0 +1,81 @@
+# system("git clone  https://github.com/bvilhjal/ldpred.git --branch 1.0.0")
+
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+
+library(dplyr)
+files <- tibble(basename = list.files("data/pheno-simu"),
+                res_file = file.path("results/simu-ldpred1", basename))
+head(files)
+bigassertr::assert_dir("results/simu-ldpred1")
+
+files_sub <- files %>%
+  filter(
+    # sub("(.*_)([0-9]+)(\\.rds)$", "\\2", basename) %in% 1:3, #iter
+    !file.exists(res_file)) %>%
+  print()
+
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "60:00:00", c = "4", mem = "32g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub, function(basename, res_file) {
+
+  y <- readRDS(file.path("data/pheno-simu", basename))
+  gwas <- do.call("rbind", lapply(1:22, function(chr) {
+    readRDS(file.path("data/sumstats-simu",
+                      sub("\\.rds$", paste0("_chr", chr, ".rds"), basename)))
+  }))
+
+  tmp <- tempfile(tmpdir = "tmp-data")
+  file_sumstats <- paste0(tmp, ".txt")
+  cbind(ukb$map, beta = gwas$estim, pval = predict(gwas, log10 = FALSE)) %>%
+    bigreadr::fwrite2(file_sumstats, sep = "\t")
+  # readLines(file_sumstats, n = 5)
+  (Neff <- round(4 / (1 / sum(y[ind.gwas] == 0) + 1 / sum(y[ind.gwas] == 1))))
+
+  file_hdf5 <- paste0(tmp, ".hdf5")
+  ldpred <- "ldpred/LDpred.py"
+  # system(glue::glue("python3 {ldpred} coord --help"))
+  system(glue::glue(
+    "python3 {ldpred} coord",
[…]

### 2020-06-25 code/run-simu-ldpred2.R

@@ -1,0 +1,110 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+
+set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+
+
+library(dplyr)
+files <- tibble(
+  basename = list.files("data/sumstats-simu"),
+  res_file = file.path("results/simu-ldpred2", basename),
+  chr = as.integer(sub("(.*_chr)([0-9]+)(\\.rds)$", "\\2", basename)),
+  pheno_file = file.path("data/pheno-simu", sub("_chr[0-9]+(\\.rds)$", "\\1", basename)),
+  gwas_file = file.path("data/sumstats-simu", basename)
+)
+head(files)
+table(files$chr)
+all(file.exists(files$pheno_file)) & all(file.exists(files$gwas_file))
+bigassertr::assert_dir("results/simu-ldpred2")
+
+files_sub <- files %>%
+  arrange(chr) %>%
+  filter(
+    # sub("(.*_)([0-9]+)(_chr[0-9]+\\.rds)$", "\\2", basename) %in% 4:10, #iter
+    !file.exists(res_file)) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 8
+plan(batchtools_slurm(resources = list(
+  t = "6:00:00", c = NCORES, mem = "64g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub, function(basename, res_file, chr, pheno_file, gwas_file) {
+
+  y <- readRDS(pheno_file)
+  gwas <- readRDS(gwas_file)
+  df_beta <- data.frame(
+    beta = gwas$estim, beta_se = gwas$std.err,
+    n_eff = 4 / (1 / sum(y[ind.gwas] == 0) + 1 / sum(y[ind.gwas] == 1)))
+
+  corr <- readRDS(paste0("data/corr/chr", chr, ".rds"))
+  (ldsc <- snp_ldsc2(corr, df_beta))
+  h2_est <- ldsc[["h2"]]
+
+  if (h2_est < 1e-4) {
+
+    rep(list(rep(0, length(ind.test))), 4) %>%
[…]

### 2020-06-25 paper/paper-ldpred2.tex
save

@@
 We then show that LDpred2 compares favorably to other polygenic score methods recently developed.
-LDpred2-auto, which automatically estimates the two LDpred parameters, performs particularly well, provided some quality control is performed on the summary statistics.
+LDpred2-auto performs particularly well, provided some quality control is performed on the summary statistics.
 LDpred2 is implemented in R package bigsnpr.
@@
 Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
-PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association,barth2020genetic}. 
+PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association,barth2020genetic,harden2020using}. 
 For example, a recently derived PGS for education attainment has been one of the most predictive variables in behavioural sciences so far \cite[]{allegrini2019genomic}.
@@
 The previous version of LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
-Yet, it has instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic}.
+Yet, it has instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic,chun2020non}.
 We have therefore implemented a new version of LDpred, LDpred2, to address these issues and improve its computational efficiency.
@@
 Indeed, LDpred2 (especially LDpred2-auto) provides the highest predictive performance for almost all the real traits tested here, and very good predictive performance for the others.
-Note that the comparison may not be totally fair for C+T and SCT, as we use the same set of variants for all methods. 
+Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
 Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.

### 2020-06-25 paper/paper-ldpred2.tex
save

@@
 Polygenic scores have become a central tool in human genetics research.
-LDpred is a popular and powerful method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
+LDpred is a popular method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
 However, LDpred has limitations that may reduce its predictive performance.
@@
 
-Authors thank Yixuan Qiu for pointing to matrix-free solvers.
+Authors thank Naomi Wray and Alkes Price for pointing to issues due to long-range LD regions in LDpred1, Yixuan Qiu for pointing to matrix-free solvers, Doug Speed and others for early testing of the software and for providing useful feedback, GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
 This research has been conducted using the UK Biobank Resource under Application Number 41181.
-Authors would also like to thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
 

### 2020-09-11 paper/paper-ldpred2.tex
start to move and add things around

@@
 
-%\usepackage{xr}
-%\externaldocument{paper-ldpred2-supp}
+\usepackage{xr}
+\externaldocument{paper-ldpred2-supp}
 
@@
 Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
-LDpred2 has 4 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimate $p$ and $h^2$ from data and therefore is free of hyper-parameters to tune.
+LDpred2 has 4 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimates $p$ and $h^2$ from data and therefore is free of hyper-parameters to tune.
 As a recall, LDpred v1 has two options: LDpred1-inf, and LDpred1-grid where only $p$ is optimized and $h^2$ is estimated from constrained LD score regression \cite[]{bulik2015ld}.
 
-We compare the two versions of LDpred using six simulation scenarios to understand the expected impact of using the new version of LDpred.
+We compare the two versions of LDpred using seven simulation scenarios to understand the expected impact of using the new version of LDpred.
 We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
+We also compare running LDpred2 (including choosing optimal hyper-paramaters) per chromosome (``perchr'') or genome-wide (``gwide'').
+Finally, we compare LDpred2 to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum, PRS-CS and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
 Area Under the ROC Curve (AUC) values are reported.
-Finally, we compare LDpred2 to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
 
@@
 Figure \ref{fig:AUC-simu} presents the simulation results comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr). 
-Six simulation scenarios are used, each repeated 10 times.
+Seven simulation scenarios are used, each repeated 10 times.
 In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome.
@@
 \begin{figure}[htb]
-\centerline{\includegraphics[width=0.85\textwidth]{AUC-simu}}
+\centerline{\includegraphics[width=0.9\textwidth]{AUC-simu}}
+\caption{Six variants of LDpred are compared (two variants of LDpred1 and four variants of LDpred2 run per chromosome) in the seven simulation scenarios (summarized in table \ref{tab:simus}).
+Briefly, the first part of the scenario name corresponds to the location of causal variants, the second part is the heritability (in \%), the third part is the number of causal variants, and the prevalence is always 15\%.
+Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC from 10 simulations for each scenario. 
+%See corresponding values in table S1.
+}
+\label{fig:AUC-simu}
+\end{figure}
+
+%%%%
+
+Figure \ref{fig:AUC-simu2-grid} presents the simulation results comparing LDpred1-grid and LDpred2-grid when varying the GWAS sample size from 10,000 to 300,000 in the simulation scenario with 3000 causal variants anywhere on the genome (``all\_40\_3000'').
+Each simulation is repeated 10 times.
+We can see that LDpred1-grid performs equally well as LDpred2-grid run genome-wide for the smallest GWAS sample sizes. 
+Yet, LDpred1-grid starts to fail when sample sizes is larger. Indeed for the scenario with sample size of 120,000, LDpred1 provides a mean AUC of [TODO] while LDpred2 provides a mean AUC of [TODO].
+Moreover, LDpred1 even performs slightly worse in the scenarios with a sample size of 300,000, while the performance of LDpred2 continues to increase steadily towards the maximum achievable of AUC of 82.5\% for a heritability of 40\% and a prevalence of 15\% \cite[]{wray2010genetic}.
+
+%% FIGURE 2 HERE
+\begin{figure}[htb]
+	\centerline{\includegraphics[width=0.8\textwidth]{AUC-simu2-grid}}
 	\caption{Six variants of LDpred are compared (two variants of LDpred1 in magenta and orange, and four variants of LDpred2 in blue and green colors) in the six simulation scenarios. 
@@
 	}
-\label{fig:AUC-simu}
+	\label{fig:AUC-simu2-grid}
 \end{figure}
@@
 
-Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr).
+[TODO: CHANGE SOME NUMBERS]
+
+Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr) run genome-wide.
 Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
@@
 For example, for BRCA, AUC improves from 58.9\% with LDpred1 to 64.6\% with LDpred2, and from 57.4\% to 77.0\% for T1D.
-These improvements are also significant for MDD and CAD, albeit to a lesser extend. 
+These improvements are also significant for MDD and CAD, albeit to a lesser extent. 
 For Asthma and RA, predictive performance of LDpred1 and LDpred2 are similar.
@@
 Moreover, LDpred2-auto also performs well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
-When comparing to other PGS methods, we only report the best LDpred2 model (chosen using the validation set) and LDpred2-auto. 
-LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA (Figure \ref{fig:AUC-all}).
-We note that LDpred2-auto usually performs even better than LDpred-best (Figure \ref{fig:AUC-all}).
-This means that choosing the best LDpred model per chromosome leads to overfitting in the validation set, which does not transfer favorably when evaluating prediction performance in the test set.
 
-%% FIGURE 2 HERE
+%% FIGURE 3 HERE
 \begin{figure}[htbp]
-\centerline{\includegraphics[width=0.85\textwidth]{AUC-real}}
+	\centerline{\includegraphics[width=0.9\textwidth]{AUC-real}}
 	\caption{Six variants of LDpred are compared (two variants of LDpred1 in magenta and orange, and four variants of LDpred2 in blue and green colors) in the real data applications using published external summary statistics.
@@
 
-%% FIGURE 3 HERE
+%%%%
+
+Figure \ref{fig:AUC-all} presents the results of real data applications comparing LDpred2 to several other PGS methods.
+LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA.
+%We note that LDpred2-auto usually performs even better than LDpred-best (Figure \ref{fig:AUC-all}).
+This means that choosing the best LDpred model per chromosome leads to overfitting in the validation set, which does not transfer favorably when evaluating prediction performance in the test set.
+
+[TODO AGAIN]
+
+%% FIGURE 4 HERE
 \begin{figure}[htbp]
-\centerline{\includegraphics[width=0.85\textwidth]{AUC-all}}
-\caption{LDpred2 is compared with C+T, SCT, lassosum and PRS-CS in the real data applications using published external summary statistics.
+	\centerline{\includegraphics[width=0.95\textwidth]{AUC-all}}
+	\caption{LDpred2 is compared with C+T, SCT, lassosum, PRS-CS in the real data applications using published external summary statistics.
 		Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
@@
 \end{figure}
+
+\subsection*{Running LDpred2 per chromosome or genome-wide?}
+
+Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, as it was recommended in the first versions of this paper.
+[DESCRIBE ALL FIGURES SHOWING THAT]
+
+Figure \ref{fig:AUC-simu2-ldpred2} shows
+
+Figure \ref{fig:AUC-real-ldpred2} shows
 
@@
 We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved}. 
-Indeed, these regions, especially the HLA region, contain lots of variants associated with many traits, and are therefore very useful for prediction.
+Indeed, these regions, especially the HLA region, contain many variants associated with many traits, and are therefore very useful for prediction.
 
@@
 First, we provide a sparse option in LDpred2-grid which provides models that truly encourage sparsity, in contrast to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
-In practice, the sparse version od LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as shown in \cite{bolli2019software}).
+In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as shown in \cite{bolli2019software}).
 %This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs better or equally well as the non-sparse model.
@@
 Each simulation scenario is repeated 10 times and averages of the Area Under the ROC Curve (AUC) are reported.
+
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|c|}
+		\hline
+		Simulation & Number and location of causal variants & Heritability & Prevalence \\
+		\hline
+		all\_40\_300 & 300 anywhere on the genome & 40\% & 15\% \\
+		all\_40\_3000 & 3000 anywhere on the genome & 40\% & 15\% \\
+		all\_40\_30000 & 30,000 anywhere on the genome & 40\% & 15\% \\
+		all\_40\_300000 & 300,000 anywhere on the genome & 40\% & 15\% \\
+		HLA\_30\_300 & 300 in the HLA region & 30\% & 15\% \\
+		HLA\_30\_3000 & 3000 in the HLA region & 30\% & 15\% \\
+		both\_40 & 10,000 anywhere on the genome + 300 in the HLA region & 20\% + 20\% & 15\% \\
+		\hline
+	\end{tabular}
+\caption{The seven simulations scenarios. \label{tab:simus}}
+\end{table}
 
@@
 
+\subsection*{Quality control of summary statistics}
+
+For summary statistics of binary traits derived from a logistic regression, 
+\begin{equation}\label{eq:approx-sd}
+\text{sd}(\boldsymbol{G_j}) \approx \dfrac{2}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n_\text{eff}}} ~,
+\end{equation}
+where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. Anyway, we recommend to verify this assumption and to perform some quality control.
+Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
+However, in real data applications, where summary statistics come from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
+Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics (right-hand side of equation \eqref{eq:approx-sd}) and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set (left-hand side).
+We removed variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-09-12 paper/paper-ldpred2.tex
continue moving and adding things around

@@
 
+[NEED TO INCLUDE FIGURE \ref{fig:AUC-auto} ABOUT AUTO MODELS]
+
 %% FIGURE 4 HERE
@@
 
-Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, as it was recommended in the first versions of this paper.
-[DESCRIBE ALL FIGURES SHOWING THAT]
-
-Figure \ref{fig:AUC-simu2-ldpred2} shows
-
-Figure \ref{fig:AUC-real-ldpred2} shows
+Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, contrary to was recommended in the first versions of this paper.
+Figure \ref{fig:AUC-simu2-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS shample size is small.
+For example, in the scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from [TODO] to, LDpred2-grid from to, and LDpred2-auto from to [TODO].
+We note that however that the simulation scenario used to produce results in figure \ref{fig:AUC-simu2-ldpred2} is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
+Figure \ref{fig:AUC-real-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
+Indeed, except for LDpred2-inf applied to Type 1 Diabetes (T1D) for which running it per chromosome provides an AUC of [TODO] versus of [TODO] when run genome-wide, all other cases show similar or better performance for running all variants of LDpred2 genome-wide.
 
@@
 
-We simulate binary phenotypes with a heritability of $h^2=0.4$ (or 0.3)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}. 
+We simulate binary phenotypes with a heritability of $h^2=0.4$ (or $h^2=0.3$ when causal effects are all located in the HLA region of chromosome 6)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}.
 We vary the number of causal variants (300, 3000, 30,000, or 300,000) to match a range of genetic architectures from low to high polygenicity.
@@
 Causal variants are chosen randomly anywhere on the genome. 
+Such simulation of phenotypes based on real genotypes is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
+We also vary the sample size to compute GWAS summary statistics in the scenario with 3000 causal variants; in addition to a sample size of 300,000, we also use 10,000, 20,000, 50,000 and 120,000. 
 We design two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25.5-33.5 Mb).
 Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
-Such simulation of phenotypes based on real genotypes is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
+Finally, we design a seventh simulation scenario as a mixture of previous scenarios; we simulate 300 causal variants in the HLA region explaining 20\% of the variance in liability, and 10,000 causal variants anywhere on the genome explaining another 20\% (thus $h^2 = 0.4$ in total).
+
+All simulation scenarios are summarized in table \ref{tab:simus}.
 Each simulation scenario is repeated 10 times and averages of the Area Under the ROC Curve (AUC) are reported.
+The 95\% confidence interval (CI) from 10,000 non-parametric bootstrap replicates of the mean AUC of the 10 simulations for each scenario is also reported.
 
@@
 	\centering
-	\begin{tabular}{|c|c|c|c|c|c|c|}
+	\begin{tabular}{|c|c|c|c|c|c|c|c|}
 		\hline
-		Simulation & Number and location of causal variants & Heritability & Prevalence \\
+		Simulation & Number and location of causal variants & Heritability & N (x1000) &  Prevalence \\
 		\hline
-		all\_40\_300 & 300 anywhere on the genome & 40\% & 15\% \\
-		all\_40\_3000 & 3000 anywhere on the genome & 40\% & 15\% \\
-		all\_40\_30000 & 30,000 anywhere on the genome & 40\% & 15\% \\
-		all\_40\_300000 & 300,000 anywhere on the genome & 40\% & 15\% \\
-		HLA\_30\_300 & 300 in the HLA region & 30\% & 15\% \\
-		HLA\_30\_3000 & 3000 in the HLA region & 30\% & 15\% \\
-		both\_40 & 10,000 anywhere on the genome + 300 in the HLA region & 20\% + 20\% & 15\% \\
+		all\_40\_300 & 300 anywhere on the genome & 40\% & 300 & 15\% \\
+		all\_40\_3000 & 3000 anywhere on the genome & 40\% & 10, 20, 50, 120, 300 & 15\% \\
+		all\_40\_30000 & 30,000 anywhere on the genome & 40\% & 300 & 15\% \\
+		all\_40\_300000 & 300,000 anywhere on the genome & 40\% & 300 & 15\% \\
+		HLA\_30\_300 & 300 in the HLA region & 30\% & 300 & 15\% \\
+		HLA\_30\_3000 & 3000 in the HLA region & 30\% & 300 & 15\% \\
+		both\_40 & 10,000 anywhere on the genome + 300 in the HLA region & 20\% + 20\% & 300 & 15\% \\
 		\hline
 	\end{tabular}
-\caption{The seven simulations scenarios. \label{tab:simus}}
+\caption{The seven simulations scenarios. N: GWAS sample size. \label{tab:simus}}
 \end{table}
@@
 We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We defined phenotypes as in \cite{prive2019making}. Some quality control is applied to summary statistics (see Supplementary Note).
+We defined phenotypes as in \cite{prive2019making}. Some quality control is applied to summary statistics (see Method section ``Quality control of summary statistics'' below).
 For more details, please refer to our R code (Software and code availability section).
@@
 
+[TODO: CHANGE]
 In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
-Then, we compare the best of the four LDpred2 models (chosen using the validation set, and for each chromosome) and LDpred2-auto to four other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum and PRS-CS \cite[]{prive2019making,mak2017polygenic,ge2019polygenic}.
-For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making}.
+Then, we compare LDpred2-inf, LDpred2-grid and LDpred2-auto to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum(-auto), PRS-CS(-auto) and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
+For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making} [TODO: DESCRIBE PARAMETERS].
 For lassosum, we use the default grid of hyper-parameters $s$ and $\lambda$. 
-Finally, for PRS-CS, we use the grid $\{10^{-6}, 10^{-5}, \dots, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters $a$ and $b$.
+We also alternatively choose the optimal hyper-parameters based on pseudo-validation (instead of choosing them based on performance in the validation set) and refer to this method as lassosum-auto.
+Finally, for PRS-CS, we use the grid $\{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}, 0.01, 0.1, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters $a$ and $b$.
+PRS-CS-auto automatically estimates $\Phi$.
+
+[TODO: DESCRIBE SBayesR]
 
@@
 The first approximation is possible because $\hat{\gamma}_j$ is expected to be small, while the second approximation assumes that the effects from covariates are small. 
-Thus, we can derive $\text{sd}(\boldsymbol{G_j}) \approx \dfrac{\text{sd}(\boldsymbol{y})}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n}}$ and then $\left(\text{sd}(\boldsymbol{G_j}) ~ \hat{\gamma}_j\right) \approx \dfrac{\hat{\gamma}_j}{\text{se}(\hat{\gamma}_j)} \dfrac{\text{sd}(\boldsymbol{y})}{\sqrt{n}}$.
+Thus we can derive 
+\begin{equation}\label{eq:approx-sd-lin}
+\text{sd}(\boldsymbol{G_j}) \approx \dfrac{\text{sd}(\boldsymbol{y})}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n}}
+\end{equation}
+and then $\left(\text{sd}(\boldsymbol{G_j}) ~ \hat{\gamma}_j\right) \approx \dfrac{\hat{\gamma}_j}{\text{se}(\hat{\gamma}_j)} \dfrac{\text{sd}(\boldsymbol{y})}{\sqrt{n}}$.
 Let us go back to equation \ref{eq:joint}. 
@@
 In the case of binary phenotypes, the approximation of the standard deviations is slightly different (see Supplementary Note).
+
+\subsection*{Quality control of summary statistics}
+
+For summary statistics of binary traits derived from a logistic regression, instead of equation \eqref{eq:approx-sd-lin}, we have
+\begin{equation}\label{eq:approx-sd}
+\text{sd}(\boldsymbol{G_j}) \approx \dfrac{2}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n_\text{eff}}} ~,
+\end{equation}
+where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. We recommend to verify this assumption and to perform some quality control.
+Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
+However, in real data applications, where summary statistics come from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
+Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics (right-hand side of equation \eqref{eq:approx-sd-lin} or \eqref{eq:approx-sd}) and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set (left-hand side).
+Note that, in order to compute $\text{SD}_\text{ss}$ in the case of summary statistics from a linear regression, $\text{sd}(\boldsymbol{y})$ from equation \eqref{eq:approx-sd-lin} can be estimated using e.g.\ the median value of $\text{sd}(\boldsymbol{G_j}) \cdot \text{se}(\hat{\gamma}_j) \cdot \sqrt{n}$.
+We recommend removing variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
 
@@
 
-\subsection*{Quality control of summary statistics}
-
-For summary statistics of binary traits derived from a logistic regression, 
-\begin{equation}\label{eq:approx-sd}
-\text{sd}(\boldsymbol{G_j}) \approx \dfrac{2}{\text{se}(\hat{\gamma}_j) ~ \sqrt{n_\text{eff}}} ~,
-\end{equation}
-where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. Anyway, we recommend to verify this assumption and to perform some quality control.
-Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
-However, in real data applications, where summary statistics come from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
-Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics (right-hand side of equation \eqref{eq:approx-sd}) and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set (left-hand side).
-We removed variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Authors thank Naomi Wray and Alkes Price for pointing to issues due to long-range LD regions in LDpred1, Yixuan Qiu for pointing to matrix-free solvers, Doug Speed and others for early testing of the software and for providing useful feedback, GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
+Authors would like to thank the four reviewers of this paper for their relevant comments and valuable suggestions, Naomi Wray and Alkes Price for pointing to issues due to long-range LD regions in LDpred1, Yixuan Qiu for pointing to matrix-free solvers, Doug Speed and others for early testing of the software and for providing useful feedback, GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
 This research has been conducted using the UK Biobank Resource under Application Number 41181.

### 2020-09-15 paper/paper-ldpred2.tex
continue moving and adding things around

@@
 \providecommand{\tabularnewline}{\\}
+\renewcommand*{\arraystretch}{1.2}
 
@@
 
-We compare the two versions of LDpred using seven simulation scenarios to understand the expected impact of using the new version of LDpred.
+We compare the two versions of LDpred using multiple simulation scenarios to understand the expected impact of using the new version of LDpred.
 We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
@@
 
+[TODO: SHOULD REMOVE THE "best\_" BEFORE ldpred2\_grid?]
+
 %% FIGURE 1 HERE
@@
 \centerline{\includegraphics[width=0.9\textwidth]{AUC-simu}}
-\caption{Six variants of LDpred are compared (two variants of LDpred1 and four variants of LDpred2 run per chromosome) in the seven simulation scenarios (summarized in table \ref{tab:simus}).
+\caption{Six variants of LDpred are compared (two variants of LDpred1 and four variants of LDpred2 run per chromosome) in the seven simulation scenarios summarized in table \ref{tab:simus}.
 Briefly, the first part of the scenario name corresponds to the location of causal variants, the second part is the heritability (in \%), the third part is the number of causal variants, and the prevalence is always 15\%.
-Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC from 10 simulations for each scenario. 
+Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
 %See corresponding values in table S1.
@@
 
+[TODO: ADD CORRESPONDING TABLES WHEN CODE OUT]
+
 %%%%
@@
 Figure \ref{fig:AUC-simu2-grid} presents the simulation results comparing LDpred1-grid and LDpred2-grid when varying the GWAS sample size from 10,000 to 300,000 in the simulation scenario with 3000 causal variants anywhere on the genome (``all\_40\_3000'').
-Each simulation is repeated 10 times.
-We can see that LDpred1-grid performs equally well as LDpred2-grid run genome-wide for the smallest GWAS sample sizes. 
-Yet, LDpred1-grid starts to fail when sample sizes is larger. Indeed for the scenario with sample size of 120,000, LDpred1 provides a mean AUC of [TODO] while LDpred2 provides a mean AUC of [TODO].
-Moreover, LDpred1 even performs slightly worse in the scenarios with a sample size of 300,000, while the performance of LDpred2 continues to increase steadily towards the maximum achievable of AUC of 82.5\% for a heritability of 40\% and a prevalence of 15\% \cite[]{wray2010genetic}.
+We can see that LDpred1-grid performs equally well as LDpred2-grid (when LDpred2 is run genome-wide) for the smallest GWAS sample sizes. 
+Yet, LDpred1-grid starts failing when sample size becomes larger.
+Indeed for the scenario with a sample size of 120,000, LDpred1 provides a mean AUC of 70.7\% while LDpred2 provides a mean AUC of 74.6\%.
+Moreover, LDpred1 even performs slightly worse with a mean AUC of 70.4\% when the sample size is 300,000, while the performance of LDpred2 continues to increase steadily to 79.1\% towards the maximum achievable AUC of 82.5\% (for a heritability of 40\% and a prevalence of 15\%, see \cite{wray2010genetic}).
 
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{AUC-simu2-grid}}
-	\caption{Six variants of LDpred are compared (two variants of LDpred1 in magenta and orange, and four variants of LDpred2 in blue and green colors) in the six simulation scenarios. 
-		Scenarios are named in 4 parts separated by underscores: 1) ``all'' means that causal variants are randomly sampled anywhere on the genome while ``HLA'' means that they are sampled in the HLA region of chromosome 6 (25.5-33.5 Mb); 2) the heritability (in \%); 3) the number of causal variants; 4) the prevalence (in \%).
+	\caption{LDpred1-grid is compared to LDpred2-grid when varying GWAS sample size in scenario ``all\_40\_3000''.  
 	Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
@@
 Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
-For T1D, T2D, BRCA and PRCA, all main LDpred2 models perform much better than LDpred1.
-For example, for BRCA, AUC improves from 58.9\% with LDpred1 to 64.6\% with LDpred2, and from 57.4\% to 77.0\% for T1D.
-These improvements are also significant for MDD and CAD, albeit to a lesser extent. 
+For BRCA, CAD, MDD, PRCA, T1D and T2D, LDpred2-inf and LDpred2-grid perform much better than LDpred1-inf and LDpred1-grid respectively.
+For example, for BRCA, AUC improves from 58.9\% with LDpred1-grid to 65.5\% with LDpred2-grid, and from 57.4\% to 78.4\% for T1D.
 For Asthma and RA, predictive performance of LDpred1 and LDpred2 are similar.
-As in simulations, the sparse version of LDpred2 performs similarly to the non-sparse version.
-Moreover, LDpred2-auto also performs well, and even better than LDpred2-grid for PRCA with an improvement from an AUC of 68.2\% to 70.2\%.
+As in simulations, the sparse version of LDpred2-grid performs as well as the non-sparse version.
+Sparsity of resulting effects ranges from 19.3\% for RA to 54.4\% for Asthma [TODO: ADD REF TO TABLE THAT SHOWS THE SPARSITY OF THE CORRESPONDING MODEL?]
 
@@
 Figure \ref{fig:AUC-all} presents the results of real data applications comparing LDpred2 to several other PGS methods.
-LDpred2 performs best for Asthma, RA, T1D and TD2, while lassosum performs best for BRCA, CAD, MDD and PRCA.
-%We note that LDpred2-auto usually performs even better than LDpred-best (Figure \ref{fig:AUC-all}).
-This means that choosing the best LDpred model per chromosome leads to overfitting in the validation set, which does not transfer favorably when evaluating prediction performance in the test set.
-
-[TODO AGAIN]
-
-[NEED TO INCLUDE FIGURE \ref{fig:AUC-auto} ABOUT AUTO MODELS]
+LDpred2 performs best for all phenotypes, lassosum performs relatively well for all phenotypes, PRS-CS always performs slightly worse than lassosum, and SBayesR performs as well as LDpred2 for BRCA, MDD, PRCA and T2D, but severely underperforms for the autoimmune diseases (T1D and RA) with an AUC of 58.1\% for T1D compared to 78.4\% with LDpred2-grid.
 
@@
 
+Figure \ref{fig:AUC-auto} presents the results of real data applications comparing ``grid'' models with their ``auto'' counterparts, i.e.\ models that directly estimate parameters from the data and do not require tuning hyper-parameters using a validation set.
+We remind readers that SBayesR is such an ``auto'' model, but does not have any ``grid'' counterpart.
+All ``grid'' models usually perform better than their ``auto'' counterpart. 
+Yet, LDpred2-auto performs similarly as LDpred2-grid, as discussed before, except for CAD and to a lesser extent for Asthma. 
+Similarly, lassosum-auto performs similarly as lassosum, expect for MDD (AUC of 54.9\% vs 59.4\%). 
+Similarly, PRS-CS-auto also underperforms for MDD (AUC of 57.5\% for PRS-CS-auto vs 53.9\% for PRS-CS).
+In contrast, SBayesR does not underperforms for MDD but severely underperforms for T1D and RA, as discussed previously.
+
+LDpred2-auto performs well except for CAD for which it provides an AUC of [TODO]\% versus of [TODO]\% with LDpred2-grid.
+[TODO: TALK ABOUT THE WRONG H2 ESTIMATION FOR CAD?
++ MOVE THIS TO THE DISCUSSION]
+
 \subsection*{Running LDpred2 per chromosome or genome-wide?}
@@
 Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, contrary to was recommended in the first versions of this paper.
-Figure \ref{fig:AUC-simu2-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS shample size is small.
-For example, in the scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from [TODO] to, LDpred2-grid from to, and LDpred2-auto from to [TODO].
+Figure \ref{fig:AUC-simu2-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS sample size is small.
+For example, in the scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from 57.2\% to 57.7\%, LDpred2-grid from 58.6\% to 59.6\%, and LDpred2-auto from 57.3\% to 59.3\%.
 We note that however that the simulation scenario used to produce results in figure \ref{fig:AUC-simu2-ldpred2} is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
 Figure \ref{fig:AUC-real-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
-Indeed, except for LDpred2-inf applied to Type 1 Diabetes (T1D) for which running it per chromosome provides an AUC of [TODO] versus of [TODO] when run genome-wide, all other cases show similar or better performance for running all variants of LDpred2 genome-wide.
+Indeed, except for LDpred2-inf applied to T1D for which running it per chromosome provides an AUC of [TODO] versus of [TODO] when run genome-wide, all other cases show similar or better performance for running all variants of LDpred2 genome-wide.
+For example,
 
@@
 \section{Discussion}
+
+[TODO: ALSO TALK ABOUT LDPRED2 PARAMS AND POSSIBLY FUTURE WORK OF IMPROVING THEM VIA E.G. FUNCTIONAL ANNOTATIONS.]
 
@@
 
-We use the UK Biobank data for both real data analyses and simulations \cite[]{bycroft2018uk}. 
+We use the UK Biobank imputed data for both real data analyses and simulations \cite[]{bycroft2018uk}. 
+We use dosage data for LDpred2, which we read from BGEN files using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}.
+For the other software that require a PLINK bed file as input, we use function \texttt{snp\_writeBed} that round dosages to write genotype data in bed format.
 We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have passed some quality control \cite[]{bycroft2018uk}.
@@
 
-We simulate binary phenotypes with a heritability of $h^2=0.4$ (or $h^2=0.3$ when causal effects are all located in the HLA region of chromosome 6)  using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}.
+We simulate binary phenotypes with a heritability of $h^2=0.4$ using a Liability Threshold Model (LTM) with a prevalence of 15\% \cite[]{falconer1965inheritance}. 
 We vary the number of causal variants (300, 3000, 30,000, or 300,000) to match a range of genetic architectures from low to high polygenicity.
@@
 We design two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25.5-33.5 Mb).
+In these two scenarios, we use $h^2=0.3$ instead of $h^2=0.4$ because the total heritability is gathered in one chromosome only.
 Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
@@
 	\centering
-	\begin{tabular}{|l|c|c|c|c|}
+	\begin{tabular}{|l||c|c|c||c|c|c|c|}
 		\hline
-		Trait & GWAS citation & GWAS sample size & GWAS \#variants \\
+		Trait & GWAS citation & GWAS sample size & GWAS \#variants & matched \#variants & Mean $\chi^2$ & \#hits & \#indep hits \\
 		\hline
-		Breast cancer (BRCA) & \cite{michailidou2017association} & 137,045 / 119,078 & 11,792,542 \\
-		Rheumatoid arthritis (RA) & \cite{okada2014genetics} & ~~29,880 / ~~73,758 & ~~9,739,303 \\
-		Type 1 diabetes (T1D) & \cite{censin2017childhood} & ~~~~~5913 / ~~~~~8828  & ~~8,996,866 \\
-		Type 2 diabetes (T2D) & \cite{scott2017expanded} & ~~26,676 / 132,532 & 12,056,346 \\
-		Prostate cancer (PRCA) & \cite{schumacher2018association} & ~~79,148 / ~~61,106 & 20,370,946 \\
-		Depression (MDD) & \cite{wray2018genome} & ~~59,851 / 113,154 & 13,554,550 \\
-		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & ~~60,801 / 123,504 & ~~9,455,778 \\
-		Asthma & \cite{demenais2018multiancestry} & ~~19,954 / 107,715 & ~~2,001,280 \\
+		Breast cancer (BRCA) & \cite{michailidou2017association} & 137,045 / 119,078 & 11,792,542 & 1,114,424 & 1.70 & 2941 & 157 \\
+		Rheumatoid arthritis (RA) & \cite{okada2014genetics} & ~~29,880 / ~~73,758 & ~~9,739,303 & ~~~656,087 & 2.10 & 3663 & 75 \\
+		Type 1 diabetes (T1D) & \cite{censin2017childhood} & ~~~~~5913 / ~~~~~8828  & ~~8,996,866 & ~~~514,420 & 1.93 & 1902 & 39 \\
+		Type 2 diabetes (T2D) & \cite{scott2017expanded} & ~~26,676 / 132,532 & 12,056,346 & 1,108,760 & 1.24 & 388 & 39 \\
+		Prostate cancer (PRCA) & \cite{schumacher2018association} & ~~79,148 / ~~61,106 & 20,370,946 & 1,115,688 & 1.55 & 2762 & 137 \\
+		Depression (MDD) & \cite{wray2018genome} & ~~59,851 / 113,154 & 13,554,550 & 1,103,440 & 1.27 & 166 & 5 \\
+		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & ~~60,801 / 123,504 & ~~9,455,778 & 1,108,313 & 1.13 & 352 & 37 \\
+		Asthma & \cite{demenais2018multiancestry} & ~~19,954 / 107,715 & ~~2,001,280 & ~~~980,430 & 1.10 & 365 & 15 \\
 		\hline
 	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used. The GWAS sample size is the number of cases / controls in the GWAS. \label{tab:sumstats}}
+	\caption{Summary of external GWAS summary statistics used.
+The GWAS sample size is the number of cases / controls in the GWAS.
+The mean chi-squared statistics $\chi^2$ and the number of hits ($p < 5 \cdot 10^{-8}$) are reported after restricting to HapMap3 variants and matching with the UKBB data.
+Independent hits are the hits remaining after clumping at $r^2 > 0.01$ within 10 Mbp windows.
+\label{tab:sumstats}}
 \end{table}

### 2020-09-16 paper/paper-ldpred2.tex
same

@@
 These scores have proven useful for studying the genetic architecture and relationships between diseases and traits \cite[]{purcell2009common,kong2018nature}.
-Moreover, there are high hopes for using these scores in clinical practice to improve disease risk estimates and prognostic accuracies.
-The heritability, i.e.\ the proportion of phenotypic variance that is attributable to genetics, determines an upper limit on the predictive performance of PGS and thus their value as a diagnostic tool.
+Moreover, there are high hopes for using these scores in clinical practice to improve disease risk estimates and predictive accuracy.
+The heritability, i.e.\ the proportion of phenotypic variance that is attributable to genetics, determines an upper limit on the predictive performance of PGS and thus their value as a predictive tool.
 Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
@@
 
-We compare the two versions of LDpred using multiple simulation scenarios to understand the expected impact of using the new version of LDpred.
+We make use of the UK Biobank data to compare the two versions of LDpred using multiple simulation scenarios to understand the expected impact of using the new version of LDpred.
 We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
 We also compare running LDpred2 (including choosing optimal hyper-paramaters) per chromosome (``perchr'') or genome-wide (``gwide'').
-Finally, we compare LDpred2 to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum, PRS-CS and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
+Finally, we use external  compare LDpred2 to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum, PRS-CS and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
 Area Under the ROC Curve (AUC) values are reported.
@@
 In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as shown in \cite{bolli2019software}).
-%This sparse model performs equally well as the non-sparse model, therefore we encourage users to test it in the grid and to choose it if it performs better or equally well as the non-sparse model.
 The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, therefore it does not require a validation step to choose hyper-parameters, making it an attractive option in many applications.
@@
 PRS-CS-auto automatically estimates $\Phi$.
+We use the 10,000 individuals of the validation set as LD reference, except for PRS-CS which provides its own LD reference based on 503 individuals from the 1000 Genomes data.
+Note that \cite{ge2019polygenic} argued that the sample size of the LD reference has little impact on the performance of PRS-CS.
 
@@
 
+\subsection*{Providing an LD reference}\label{sec:ld-ref}
+
+We provide an LD reference for European ancestry to be used by researchers who do not have access to a large dataset to be used as LD reference.
+We use the 362,320 UKBB individuals as used here, with some further quality control based on allele frequencies.
+Indeed, some large allele frequency mismatches have been reported recently in the UKBB when comparing to other datasets, which are apparently due to mismappings \cite[]{kunert2020allele}.
+Using 503 European individuals from the 1000 Genomes (1000G) data, we remove variants with allele frequency differences between UKBB and 1000G at $p<10^{-5}$ (Figure \ref{fig:af-outliers}).
+We also remove variants with minor allele frequencies less than 1\% in the 1000G or less than 0.5\% in the UKBB.
+1,054,330 variants remain, which we use to compute the LD reference provided.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-09-17 paper/paper-ldpred2.tex
revise abstract/intro/results/discu

@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
 
@@
 We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns the two LDpred parameters from data.
-We benchmark LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and efficiency, as well as providing much more accurate polygenic scores.
-We then show that LDpred2 compares favorably to other polygenic score methods recently developed.
-LDpred2-auto performs particularly well, provided some quality control is performed on the summary statistics.
+We benchmark predictive performance of LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness, and now providing much more accurate polygenic scores compared to LDpred1.
+We then show that LDpred2 also outperforms other polygenic score methods recently developed, such as lassosum, PRS-CS and SBayesR.
+Note that, in contrast to what was recommended in the first version of this paper, we now recommend to run LDpred2 genome-wide instead of per chromosome.
 LDpred2 is implemented in R package bigsnpr.
@@
 This issue has led to the removal of such regions from analyses  \cite[]{marquez2018modeling,lloyd2019improved}, which is unfortunate since this region of the genome contains many known disease-associated variants, particularly with autoimmune diseases and psychiatric disorders \cite[]{mokhtari2016major,matzaraki2017mhc}.
-%LDpred is also considered not particularly fast, nor memory-efficient, nor it provides truly sparse models as expected by its model specification.
 
-Here, we present LDpred2, a new version of LDpred that addresses these issues while markedly improving its computational efficiency.
+Here, we present LDpred2, a new version of LDpred that addresses these issues while markedly improving its computational efficiency, allowing exploring a larger grid of parameters in the same computational time as LDpred1.
 We provide this faster and more robust implementation of LDpred in R package bigsnpr \cite[]{prive2017efficient}.
@@
 Second, we also provide an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, and therefore does not require validation data to tune hyper-parameters.
-We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions, or when the proportion of causal variants is small.
+We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions, when the proportion of causal variants is small, and when sample size is large.
 We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing prediction accuracy.
-Moreover, LDpred2-auto, which does not require any validation set, performs very well, provided some quality control is performed on the summary statistics.
+Moreover, LDpred2-auto, which does not require any validation set, performs almost as well as LDpred2-grid (the main LDpred2 model which tunes hyper-parameters from a grid in a validation set), provided some quality control is performed on the summary statistics.
 
@@
 Here we present LDpred2, a new version of LDpred \cite[]{vilhjalmsson2015modeling}.
-LDpred2 has 4 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants) and $h^2$ (the SNP heritability) are tuned using a validation set; 3) LDpred2-sparse, which is similar to LDpred2-grid but where effects can be exactly 0, offering a version of LDpred that can provide sparse effects; 4) LDpred2-auto, which automatically estimates $p$ and $h^2$ from data and therefore is free of hyper-parameters to tune.
-As a recall, LDpred v1 has two options: LDpred1-inf, and LDpred1-grid where only $p$ is optimized and $h^2$ is estimated from constrained LD score regression \cite[]{bulik2015ld}.
+LDpred2 has 3 options: 1) LDpred2-inf, which provides an analytical solution under the infinitesimal model of LDpred1; 2) LDpred2-grid (or simply LDpred2) that is the main LDpred model, where a grid of values for hyper-parameters $p$ (the proportion of causal variants), $h^2$ (the SNP heritability), and possibly the sparsity option (as a third hyper-parameter) are tuned using a validation set; 3) LDpred2-auto, which automatically estimates $p$ and $h^2$ from data and therefore is free of hyper-parameters to tune.
+Note that the sparse option in LDpred2-grid slightly modifies the Gibbs sampler in LDpred2 to be able to fit effects that are exactly 0.
+For users who do not particularly care about sparsity, they can still tune this parameter in LDpred2-grid to explore a larger parameter space. 
+For this paper, to show the effect of this sparse option on predictive performance, we deliberately separate the LDpred2-grid method in two ``different'' methods that we call LDpred2-grid-nosp (sparse option always disabled) and LDpred2-grid-sp (sparse option always enabled) in the results.
+As a recall, LDpred v1 has two options: LDpred1-inf, and LDpred1-grid where only $p$ is tuned while $h^2$ is estimated from constrained LD score regression \cite[]{bulik2015ld}.
 
-We make use of the UK Biobank data to compare the two versions of LDpred using multiple simulation scenarios to understand the expected impact of using the new version of LDpred.
+We make use of the UK Biobank data to compare the two versions of LDpred using several simulation scenarios to understand the expected impact of using the new version of LDpred.
 We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
 We also compare running LDpred2 (including choosing optimal hyper-paramaters) per chromosome (``perchr'') or genome-wide (``gwide'').
-Finally, we use external  compare LDpred2 to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum, PRS-CS and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
+Finally, we compare LDpred2 to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum, PRS-CS and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
 Area Under the ROC Curve (AUC) values are reported.
@@
 For example, in simulations with 300 causal variants, a mean AUC of 73.6\% is obtained with LDpred1 while a value of 81.7\% is obtained with LDpred2.
-In these scenarios, all 3 non-infinitesimal models of LDpred2 perform equally well.
+In these scenarios, all non-infinitesimal models of LDpred2 perform equally well.
 As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are located in a long-range LD region. 
 In contrast, all LDpred2 models perform well in these two HLA scenarios, but LDpred2-auto performs slightly worse than other LDpred2 models in these extreme scenarios.
-
-[TODO: SHOULD REMOVE THE "best\_" BEFORE ldpred2\_grid?]
 
@@
 \centerline{\includegraphics[width=0.9\textwidth]{AUC-simu}}
-\caption{Six variants of LDpred are compared (two variants of LDpred1 and four variants of LDpred2 run per chromosome) in the seven simulation scenarios summarized in table \ref{tab:simus}.
+\caption{Two variants of LDpred1 are compared with four variants of LDpred2 (run per chromosome) in the seven simulation scenarios summarized in table \ref{tab:simus}.
 Briefly, the first part of the scenario name corresponds to the location of causal variants, the second part is the heritability (in \%), the third part is the number of causal variants, and the prevalence is always 15\%.
 Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
+[TODO: MAKE SURE ABOUT NAMING OF BOOTSTRAP TECHNIQUE + ADD CORRESPONDING TABLES IN SUPPMAT WHEN CODE OUT]
 %See corresponding values in table S1.
@@
 
-[TODO: ADD CORRESPONDING TABLES WHEN CODE OUT]
-
 %%%%
@@
 We can see that LDpred1-grid performs equally well as LDpred2-grid (when LDpred2 is run genome-wide) for the smallest GWAS sample sizes. 
-Yet, LDpred1-grid starts failing when sample size becomes larger.
-Indeed for the scenario with a sample size of 120,000, LDpred1 provides a mean AUC of 70.7\% while LDpred2 provides a mean AUC of 74.6\%.
-Moreover, LDpred1 even performs slightly worse with a mean AUC of 70.4\% when the sample size is 300,000, while the performance of LDpred2 continues to increase steadily to 79.1\% towards the maximum achievable AUC of 82.5\% (for a heritability of 40\% and a prevalence of 15\%, see \cite{wray2010genetic}).
+Yet, LDpred1-grid starts providing lower predictive performance than LDpred2-grid when sample size becomes larger. 
+For example, in the scenario with a sample size of 120,000, LDpred1 provides a mean AUC of 70.7\% while LDpred2 provides a mean AUC of 74.6\%.
+Moreover, LDpred1 even performs slightly worse with a mean AUC of 70.4\% when the sample size is increased to 300,000.
+In contrast, the performance of LDpred2 continues to increase steadily to 79.1\% towards the maximum achievable AUC of 82.5\% for this scenario with a heritability of 40\% and a prevalence of 15\% \cite[]{wray2010genetic}.
 
@@
 
-[TODO: CHANGE SOME NUMBERS]
-
-Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr) run genome-wide.
+Figure \ref{fig:AUC-real} presents the results of real data applications comparing LDpred1 (v1.0.0 as implemented by \cite{vilhjalmsson2015modeling}) with the new LDpred2 (as implemented in R package bigsnpr) when run genome-wide.
 Eight case-control phenotypes are used, summarized in table \ref{tab:sumstats}.
@@
 As in simulations, the sparse version of LDpred2-grid performs as well as the non-sparse version.
-Sparsity of resulting effects ranges from 19.3\% for RA to 54.4\% for Asthma [TODO: ADD REF TO TABLE THAT SHOWS THE SPARSITY OF THE CORRESPONDING MODEL?]
+Sparsity of resulting effects ranges from 19.3\% for RA to 54.4\% for Asthma [TODO: ADD REF TO TABLE THAT SHOWS THE SPARSITY OF THE CORRESPONDING MODEL WHEN OUT -> SEE SCREENSHOT IN SUPPMAT FOR NOW]
 
@@
 	\centerline{\includegraphics[width=0.9\textwidth]{AUC-real}}
-	\caption{Six variants of LDpred are compared (two variants of LDpred1 in magenta and orange, and four variants of LDpred2 in blue and green colors) in the real data applications using published external summary statistics.
+	\caption{Two variants of LDpred1 are compared with four variants of LDpred2 (run genome-wide) in the real data applications using published external summary statistics.
 	Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
+	[TODO: MAKE SURE ABOUT NAMING OF BOOTSTRAP TECHNIQUE + ADD CORRESPONDING TABLES IN SUPPMAT WHEN CODE OUT]
 	%See corresponding values in table S2.
@@
 Figure \ref{fig:AUC-all} presents the results of real data applications comparing LDpred2 to several other PGS methods.
-LDpred2 performs best for all phenotypes, lassosum performs relatively well for all phenotypes, PRS-CS always performs slightly worse than lassosum, and SBayesR performs as well as LDpred2 for BRCA, MDD, PRCA and T2D, but severely underperforms for the autoimmune diseases (T1D and RA) with an AUC of 58.1\% for T1D compared to 78.4\% with LDpred2-grid.
+LDpred2 performs best for all phenotypes, lassosum performs relatively well for all phenotypes, PRS-CS always performs slightly worse than lassosum, and SBayesR performs as well as LDpred2 for BRCA, MDD, PRCA and T2D, but severely underperforms for the autoimmune diseases (T1D and RA).
+For example, SBayesR provides an AUC of 58.1\% for T1D compared to 78.4\% with LDpred2-grid.
 
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{AUC-all}}
-	\caption{LDpred2 is compared with C+T, SCT, lassosum, PRS-CS in the real data applications using published external summary statistics.
+	\caption{LDpred2 is compared with C+T, SCT, lassosum, PRS-CS and SBayesR in the real data applications using published external summary statistics.
 	Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
@@
 
-Figure \ref{fig:AUC-auto} presents the results of real data applications comparing ``grid'' models with their ``auto'' counterparts, i.e.\ models that directly estimate parameters from the data and do not require tuning hyper-parameters using a validation set.
+Figure \ref{fig:AUC-auto} presents the results of real data applications comparing ``grid'' models with their ``auto'' counterparts, i.e.\ models that directly estimate parameters from the data and do not require tuning hyper-parameters from a grid using a validation set.
 We remind readers that SBayesR is such an ``auto'' model, but does not have any ``grid'' counterpart.
 All ``grid'' models usually perform better than their ``auto'' counterpart. 
-Yet, LDpred2-auto performs similarly as LDpred2-grid, as discussed before, except for CAD and to a lesser extent for Asthma. 
+LDpred2-auto performs similarly as LDpred2-grid, except for CAD (AUC of 61.8\% vs 63.6\%) and to a lesser extent for Asthma (AUC of 58.4\% vs 59.3\%). 
 Similarly, lassosum-auto performs similarly as lassosum, expect for MDD (AUC of 54.9\% vs 59.4\%). 
-Similarly, PRS-CS-auto also underperforms for MDD (AUC of 57.5\% for PRS-CS-auto vs 53.9\% for PRS-CS).
-In contrast, SBayesR does not underperforms for MDD but severely underperforms for T1D and RA, as discussed previously.
-
-LDpred2-auto performs well except for CAD for which it provides an AUC of [TODO]\% versus of [TODO]\% with LDpred2-grid.
-[TODO: TALK ABOUT THE WRONG H2 ESTIMATION FOR CAD?
-+ MOVE THIS TO THE DISCUSSION]
+Similarly, PRS-CS-auto also underperforms for MDD (53.9\% for PRS-CS-auto vs AUC of 57.5\% for PRS-CS).
+In contrast, SBayesR does not underperforms for MDD but severely underperforms for T1D and RA, as discussed in the previous paragraph.
 
@@
 
-Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, contrary to was recommended in the first versions of this paper.
+Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, contrary to was recommended in the first version of this paper.
 Figure \ref{fig:AUC-simu2-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS sample size is small.
-For example, in the scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from 57.2\% to 57.7\%, LDpred2-grid from 58.6\% to 59.6\%, and LDpred2-auto from 57.3\% to 59.3\%.
+For example, in the simulation scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from 57.2\% to 57.7\%, LDpred2-grid from 58.6\% to 59.6\%, and LDpred2-auto from 57.3\% to 59.3\%.
 We note that however that the simulation scenario used to produce results in figure \ref{fig:AUC-simu2-ldpred2} is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
-Figure \ref{fig:AUC-real-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
-Indeed, except for LDpred2-inf applied to T1D for which running it per chromosome provides an AUC of [TODO] versus of [TODO] when run genome-wide, all other cases show similar or better performance for running all variants of LDpred2 genome-wide.
-For example,
+However, figure \ref{fig:AUC-real-ldpred2} also shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
+Indeed, except for LDpred2-inf applied to T1D, all other cases show similar or better performance for running all variants of LDpred2 genome-wide. 
+For example, running LDpred2-grid genome-wide rather than per chromosome is particularly beneficial for CAD (AUC of 63.6\% vs 61.5\%) and PRCA (AUC of 70.2\% vs 68.2\%).
 
@@
 
+[TODO: TALK ABOUT THE WRONG H2 ESTIMATION FOR CAD IN DISCU? -> CF. TABLE + SOME FIGURES IN SUPPMAT]
+
 The previous version of LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
-Yet, it has instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported prediction accuracies \cite[]{choi2019prsice,ge2019polygenic,chun2020non}.
+Yet, it has instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported predictive accuracy of LDpred \cite[]{choi2019prsice,ge2019polygenic,chun2020non}.
 We have therefore implemented a new version of LDpred, LDpred2, to address these issues and improve its computational efficiency.
-We show that LDpred2 is more stable and provides higher predictive performance than LDpred1, particularly when handling long-range LD regions or less polygenic traits.
+We show that LDpred2 is more stable and provides higher predictive performance than LDpred1, particularly when handling long-range LD regions, less polygenic traits, and large GWAS sample sizes.
 We hypothesize that LDpred1 does not use a LD window size that is large enough to account for long-range LD such as in the HLA region.
@@
 This enables LDpred2 to work well even when causal variants are located in long-range LD regions.
-%We hypothesize that using a larger LD window explains the 5 real traits where LDpred2 performs much better than LDpred1, especially because LDpred2-inf performs better than LDpred1-inf.
 We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved}. 
 Indeed, these regions, especially the HLA region, contain many variants associated with many traits, and are therefore very useful for prediction.
+Other modifications, such as a better handling of numerical errors when working with exponentials (equation \eqref{eq:postp}), or a larger hyper-parameter search space, may also contribute to the improved robustness in LDpred2.
 
-In LDpred2, we also expand the grid of hyper-parameters examined with now more values for $p$ (21 instead of 7 by default in LDpred1) and for $h^2$ (3 instead of 1).
+In LDpred2, we expand the grid of hyper-parameters examined with now more values for $p$ (21 instead of 7 by default in LDpred1) and for $h^2$ (3 instead of 1).
 When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see below).
@@
 The efficiency of LDpred2 is achieved through an efficient parallel implementation in C++.
-The parallelization is performed on the grid of hyper-parameters, but can also be performed chromosome-wise, as each chromosome is processed independently in LDpred2 (see Methods).
-Efficient parallelization is possible because we developed an on-disk sparse matrix format accessed using memory mapping in R package bigsparser. 
-As an example, it takes around less than three hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
+Efficient parallelization over the grid of hyper-parameters is possible because we use an on-disk sparse matrix format accessed using memory mapping. 
+This special data format is available in R package bigsparser, which has been developed for this paper. 
+As an example, it takes less than three hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
+It takes less than 36 hours to run all these LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide).
 It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
@@
 First, we provide a sparse option in LDpred2-grid which provides models that truly encourage sparsity, in contrast to LDpred1 which outputs very small non-zero effect sizes \cite[]{cecile2019polygenic}.
-In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as shown in \cite{bolli2019software}).
-The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, therefore it does not require a validation step to choose hyper-parameters, making it an attractive option in many applications.
-LDpred2-auto performs equally well as other LDpred2 models in simulations.
-It also performs very well in real data applications, provided that some quality control is performed on summary statistics (see Supplementary Note).
+In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as tested in \cite{bolli2019software}).
+The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, therefore which does not require a validation set to choose hyper-parameters, making it an attractive option in many applications.
+LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications, provided that some quality control is performed on summary statistics (see Methods section ``Quality control of summary statistics'').
 This quality control aims at ensuring that effects are transferable from the external GWAS summary statistics to the data where PGS are computed.
@@
 In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
-LDpred2 also performs favorably when comparing it to recently derived methods that showed higher predictive performance compared to the previous version of LDpred. 
-Indeed, LDpred2 (especially LDpred2-auto) provides the highest predictive performance for almost all the real traits tested here, and very good predictive performance for the others.
-Note that the comparison may not be fair for C+T and SCT, as we use the same set of variants for all methods. 
-Indeed, one should probably use a set enriched for statistically significant variants in C+T and SCT \cite[]{prive2019making}.
-Further investigation is needed to determine whether doing so would be beneficial for other methods such as LDpred2, lassosum and PRS-CS, or detrimental as it may break some of the assumptions behind their respective models.
-For now, we recommend using the same HapMap3 variants used in PRS-CS and used here.
-Investigating variant selection for LDpred2 and other methods is a direction of future research for us.
+When compared to recently derived methods that showed higher predictive performance than LDpred1, LDpred2 provides the highest predictive performance for all the real traits tested here.
+For now, we recommend using the same HapMap3 variants used in PRS-CS and used here when running LDpred2. 
+Indeed, HapMap3 variants have passed a number of quality controls, are generally well imputed and offer a good coverage of the whole genome.
+However, investigating alternatives in variant selection for LDpred2 and other PGS methods is a direction of future research for us.
+Would it be beneficial e.g.\ to use a set enriched for statistically significant variants?
 
@@
 
+[TODO: TALK ABOUT AUC + CI]
+
 %% TABLE 1 HERE
@@
 		\hline
-		Trait & GWAS citation & GWAS sample size & GWAS \#variants & matched \#variants & Mean $\chi^2$ & \#hits & \#indep hits \\
+		Trait & GWAS citation & GWAS sample size & \# GWAS variants & \# matched variants & Mean $\chi^2$ & \# hits & \# indep hits \\
 		\hline
@@
 The mean chi-squared statistics $\chi^2$ and the number of hits ($p < 5 \cdot 10^{-8}$) are reported after restricting to HapMap3 variants and matching with the UKBB data.
-Independent hits are the hits remaining after clumping at $r^2 > 0.01$ within 10 Mbp windows.
+Independent hits are the hits remaining after clumping at $r^2 > 0.01$ within 10 Mbp.
 \label{tab:sumstats}}
@@
 
-The second extension, LDpred2-sparse, aims at providing sparse effect size estimates, i.e.\ some resulting effects are exactly 0.
+The second extension, which can be enabled using a third hyper-parameter in LDpred2-grid, aims at providing sparse effect size estimates, i.e.\ some resulting effects are exactly 0.
 When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j$ and $\omega_j$ to 0 (lines 6-8 of algorithm \ref{algo:ldpred}).
-Note that LDpred2-auto does not have a sparse option, but it is possible to run LDpred2-sparse with the estimates of $p$ and $h^2$ from LDpred2-auto.
+We also provide a sparse option for LDpred2-auto by running LDpred2-grid with only one model with the sparsity enabled and using the estimates of $p$ and $h^2$ from LDpred2-auto.
 

### 2020-09-18 paper/paper-ldpred2.tex
revise methods + captions

@@
 
-[TODO: ALSO TALK ABOUT LDPRED2 PARAMS AND POSSIBLY FUTURE WORK OF IMPROVING THEM VIA E.G. FUNCTIONAL ANNOTATIONS.]
+[TODO: ALSO TALK ABOUT LDPRED2 PARAMS AND POSSIBLY FUTURE WORK OF IMPROVING THEM VIA E.G. FUNCTIONAL ANNOTATIONS?]
 
@@
 We vary the number of causal variants (300, 3000, 30,000, or 300,000) to match a range of genetic architectures from low to high polygenicity.
+Causal variants are chosen randomly anywhere on the genome. 
 Liability scores are computed from a model with additive effects only: we compute the liability score of the $i$-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G}_{i,j} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G}_{i,j}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon_i$ follows a Gaussian distribution $N(0, 1 - h^2)$.
-Causal variants are chosen randomly anywhere on the genome. 
+Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
 Such simulation of phenotypes based on real genotypes is implemented in function \texttt{snp\_simuPheno} of R package bigsnpr.
-We also vary the sample size to compute GWAS summary statistics in the scenario with 3000 causal variants; in addition to a sample size of 300,000, we also use 10,000, 20,000, 50,000 and 120,000. 
+We also vary the sample size to compute GWAS summary statistics in the scenario with 3000 causal variants; in addition to a GWAS sample size of 300,000, we also use 10,000, 20,000, 50,000 and 120,000.
 We design two other simulation scenarios with 300 or 3000 causal variants randomly chosen in the HLA region (chromosome 6, 25.5-33.5 Mb).
 In these two scenarios, we use $h^2=0.3$ instead of $h^2=0.4$ because the total heritability is gathered in one chromosome only.
-Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
-Finally, we design a seventh simulation scenario as a mixture of previous scenarios; we simulate 300 causal variants in the HLA region explaining 20\% of the variance in liability, and 10,000 causal variants anywhere on the genome explaining another 20\% (thus $h^2 = 0.4$ in total).
+Finally, we design a seventh simulation scenario as a mixture of previous scenarios; we simulate 300 causal variants in the HLA region explaining 20\% of the variance in liability, and 10,000 causal variants anywhere on the genome explaining another 20\% ($h^2 = 0.4$ in total).
 
@@
 
-[TODO: CHANGE]
 In real data applications, we first compare all four LDpred2 models to the two LDpred1 models.
 Then, we compare LDpred2-inf, LDpred2-grid and LDpred2-auto to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum(-auto), PRS-CS(-auto) and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
-For C+T and SCT, we use a large grid of hyper-parameters as in \cite{prive2019making} [TODO: DESCRIBE PARAMETERS].
-For lassosum, we use the default grid of hyper-parameters $s$ and $\lambda$. 
+For C+T and SCT \cite[]{prive2019making}, we use the default large grid of hyper-parameters testing a threshold of clumping $r_c^2$ within \{0.01, 0.05, 0.1, 0.2, 0.5, 0.8, 0.95\}, a base size of clumping window within \{50, 100, 200, 500\} in Kbp where the actual window size is then computed as the base size divided by $r_c^2$, and a sequence of 50 thresholds on -log10(p-values) between 0.1 and the most significant p-value, equally spaced on a log scale.
+For lassosum \cite[]{mak2017polygenic}, we use the default grid of hyper-parameters $s$ and $\lambda$. 
 We also alternatively choose the optimal hyper-parameters based on pseudo-validation (instead of choosing them based on performance in the validation set) and refer to this method as lassosum-auto.
-Finally, for PRS-CS, we use the grid $\{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}, 0.01, 0.1, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters $a$ and $b$.
+For PRS-CS \cite[]{ge2019polygenic}, we use the grid $\{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}, 0.01, 0.1, 1\}$ for the global scaling hyper-parameter $\Phi$, and the default values for hyper-parameters a (1) and b (0.5).
 PRS-CS-auto automatically estimates $\Phi$.
-We use the 10,000 individuals of the validation set as LD reference, except for PRS-CS which provides its own LD reference based on 503 individuals from the 1000 Genomes data.
-Note that \cite{ge2019polygenic} argued that the sample size of the LD reference has little impact on the performance of PRS-CS.
-
-[TODO: DESCRIBE SBayesR]
+Note that we use the 10,000 individuals of the validation set as LD reference for all methods, except for PRS-CS which provides its own LD reference based on 503 individuals from the 1000 Genomes data; \cite{ge2019polygenic} argued that the sample size of the LD reference has little impact on the performance of PRS-CS.
+Finally, for SBayesR \cite[]{lloyd2019improved}, shrunk LD matrices are built using option `\texttt{-{}-make-shrunk-ldm}' and SBayesR is run with parameters `\texttt{-{}-pi 0.95,0.02,0.02,0.01}' and `\texttt{-{}-gamma 0.0,0.01,0.1,1}', and a chain of length 10,000 with 2000 burn-in iterations.
 
@@
 Note that LDpred1 and other similar methods scale the output dividing by the standard deviation of genotypes. This is correct when $\text{var}(\boldsymbol{y}) = 1$ only.
-In the case of binary phenotypes, the approximation of the standard deviations is slightly different (see Supplementary Note).
 
@@
 Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics (right-hand side of equation \eqref{eq:approx-sd-lin} or \eqref{eq:approx-sd}) and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set (left-hand side).
-Note that, in order to compute $\text{SD}_\text{ss}$ in the case of summary statistics from a linear regression, $\text{sd}(\boldsymbol{y})$ from equation \eqref{eq:approx-sd-lin} can be estimated using e.g.\ the median value of $\text{sd}(\boldsymbol{G_j}) \cdot \text{se}(\hat{\gamma}_j) \cdot \sqrt{n}$.
+Note that, in order to compute $\text{SD}_\text{ss}$ in the case of summary statistics from a linear regression, $\text{sd}(\boldsymbol{y})$ from equation \eqref{eq:approx-sd-lin} can be estimated using e.g.\ the median value of $\text{sd}(\boldsymbol{G_j}) \cdot \text{se}(\hat{\gamma}_j) \cdot \sqrt{n}$ or by using the fact that the maximum value of $\text{SD}_\text{ss}$ should be $\sqrt{0.5}$.
 We recommend removing variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
@@
 
-To estimate effect sizes $\beta_j$, \cite{vilhjalmsson2015modeling} use a Gibbs sampler, as described in algorithm \ref{algo:ldpred}.
+To estimate effect sizes $\beta_j$, we use a Gibbs sampler as in \cite{vilhjalmsson2015modeling}, which is described in algorithm \ref{algo:ldpred}.
 First, the residualized marginal effect for variant $j$ is computed as
@@
 This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly to data without the need of a validation dataset to choose best-performing hyper-parameters.
-To estimate $p$, we count the number of causal variants (i.e.\ $M_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
+To estimate $p$ in the Gibbs sampler, we count the number of non-zero variants (i.e.\ $M_c = \sum_j(\beta_j \neq 0)$ in equation \ref{eg:random_beta}). 
 We can assume that $M_c \sim \text{Binom}(M, p)$, so if we place a prior $p \sim \text{Beta}(1, 1) \equiv \mathcal{U}(0, 1)$, we can sample $p$ from the posterior $p \sim \text{Beta}(1 + M_c, 1 + M - M_c)$.
@@
 When the sparse solution is sought and when $\bar{p}_j < p$, we set $\beta_j$ and $\omega_j$ to 0 (lines 6-8 of algorithm \ref{algo:ldpred}).
-We also provide a sparse option for LDpred2-auto by running LDpred2-grid with only one model with the sparsity enabled and using the estimates of $p$ and $h^2$ from LDpred2-auto.
+We also provide a sparse option for LDpred2-auto by running LDpred2-grid with one set of parameters only: with the sparsity enabled and using the estimates of $p$ and $h^2$ from LDpred2-auto.
 
@@
 We recommend to use genetic positions and to use a size parameter of 3 cM when computing the correlation between variants for LDpred2. 
-Note that, in the code, we use \texttt{size = 3 / 1000} since parameter \texttt{size} is internally multiplied by 1000 in the bigsnpr functions.
+Note that, in the code, we use `\texttt{size = 3 / 1000}' since parameter \texttt{size} is internally multiplied by 1000 in the bigsnpr functions.
 
-\subsection*{New strategy for running LDpred2}\label{sec:bychr}
+\subsection*{Running LDpred2 per chromosome or genome-wide?}\label{sec:bychr}
 
-We run LDpred2 per chromosome.
-Even if it is possible to run LDpred2 genome-wide, this approach has two drawbacks.
+In this paper, we investigate if it would be beneficial to run LDpred2 per chromosome, instead of genome-wide.
+Indeed, running LDpred2 genome-wide has two drawbacks.
 First, it can be memory and computationally demanding to do so.
@@
 Assuming the same model for genetic effects genome-wide could result in model misspecification, which would lead to suboptimal predictive performance.
-Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results. 
+Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results (at least for LDpred-inf).
 To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
-Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC.
+Here, we use the Z-Score from the regression of the phenotype by the PRS since we have found it more robust than using the AUC when running LDpred2 per chromosome.
 
@@
 1,054,330 variants remain, which we use to compute the LD reference provided.
+We also provide an example R script on how to use this LD reference provided.
 

### 2020-09-30 code/example-with-provided-ldref.R
reexport code

@@ -1,0 +1,122 @@
+library(bigsnpr)
+library(bigreadr)
+
+NCORES <- 30  # TO MODIFY
+
+## Information for the variants provided in the LD reference
+map_ldref <- readRDS("ld-ref/map.rds")
+
+## Breast cancer summary statistics
+
+# download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/MichailidouK_29059683_GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
+#               destfile = "tmp-data/sumstats_BRCA.txt.gz")
+# R.utils::gunzip("tmp-data/sumstats_BRCA.txt.gz")
+sumstats <- fread2("tmp-data/sumstats_BRCA.txt", na.strings = "NULL",
+                   select = c("chr", "position_b37", "a0", "a1",
+                              "bcac_onco_icogs_gwas_beta",
+                              "bcac_onco_icogs_gwas_se"),
+                   col.names = c("chr", "pos", "a0", "a1", "beta", "beta_se"))
+sumstats$n_eff <- 4 / (1 / 137045 + 1 / 119078)
+
+info_snp <- snp_match(sumstats, map_ldref)
+# 11,792,542 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 1,054,287 variants have been matched; 0 were flipped and 0 were reversed.
+(info_snp <- tidyr::drop_na(tibble::as_tibble(info_snp)))
+
+sd_ldref <- with(info_snp, sqrt(2 * af_UKBB * (1 - af_UKBB)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2))
+
+is_bad <-
+  sd_ss < (0.5 * sd_ldref) | sd_ss > (sd_ldref + 0.1) | sd_ss < 0.1 | sd_ldref < 0.05
+
+library(ggplot2)
+qplot(sd_ldref, sd_ss, color = is_bad, alpha = I(0.5)) +
+  theme_bigstatsr() +
+  coord_equal() +
+  scale_color_viridis_d(direction = -1) +
+  geom_abline(linetype = 2, color = "red") +
+  labs(x = "Standard deviations derived from allele frequencies of the LD reference",
+       y = "Standard deviations derived from the summary statistics",
+       color = "Removed?")
+
+df_beta <- info_snp[!is_bad, ]
+
+# Here, you also want to restrict to the variants present
[…]

### 2020-09-30 code/plot-approx-sd.R

@@ -2,10 +2,7 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
-ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+load("data/ind_gwas_val_test.RData")
 
 sd <- readRDS("data/sd.rds")
 

### 2020-09-30 code/plot-params.R

@@ -1,0 +1,94 @@
+library(dplyr)
+library(purrr)
+
+## LDpred2-gwide params
+files <- list.files("results/ldpred2-gwide", full.names = TRUE)
+res_ldpred2_gwide <- tibble(res = lapply(files, readRDS)) %>%
+  tidyr::unnest_wider("res")
+
+res_ldpred2_gwide %>%
+  transmute(
+    Trait = sub("\\.rds$", "", basename(files)),
+    h2_ldsc_obs = map_dbl(params, ~ median(.x$h2)),
+    sparsity_grid = map_dbl(params, ~ {
+      .x %>%
+        filter(sparse) %>%
+        arrange(desc(score)) %>%
+        slice(1) %>%
+        pull("sparsity")
+    }),
+    p_auto  = map_dbl(auto, ~ median(map_dbl(.x, "p_est"))),
+    h2_auto_obs = map_dbl(auto, ~ median(map_dbl(.x, "h2_est"))),
+    prev_pop = map_dbl(Trait, ~ {
+      mean(readRDS(paste0("data/pheno/", .x, ".rds")), na.rm = TRUE)
+    }),
+    h2_auto_liab = map2_dbl(h2_auto_obs, prev_pop, ~ .x * bigsnpr::coef_to_liab(.y))
+  ) %>%
+  print() %>%
+  xtable::xtable(align = "|l|c|c|c|c|c|c|c|") %>%
+  print(include.rownames = FALSE)
+#   Trait  h2_ldsc_obs sparsity_grid   p_auto h2_auto_obs prev_pop h2_auto_liab
+#   <chr>        <dbl>         <dbl>    <dbl>       <dbl>    <dbl>        <dbl>
+# 1 Asthma      0.104          0.544 0.000738      0.0534  0.150         0.0640
+# 2 BRCA        0.127          0.414 0.00474       0.141   0.0758        0.136
+# 3 CAD         0.0819         0.441 0.000978      0.0428  0.0619        0.0387
+# 4 MDD         0.0923         0.377 0.0675        0.0877  0.0893        0.0890
+# 5 PRCA        0.188          0.516 0.00181       0.189   0.0546        0.164
+# 6 RA          0.416          0.193 0.0774        0.562   0.0286        0.406
+# 7 T1D         0.955          0.498 0.00482       1.40    0.00244       0.575
[…]

### 2020-09-30 code/plot-res-all.R

@@ -2,27 +2,21 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
+load("data/ind_val_test.RData")
 
-library(tidyverse)
 library(future)
 (NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
 plan(multisession, workers = NCORES)
 
-pred_ldpred2 <-
-  tibble(res_file = list.files("results/ldpred2", "\\.rds$", full.names = TRUE)) %>%
-  mutate(Trait = sub("(.*/)([[:alnum:]]+)(_chr[0-9]+\\.rds)$", "\\2", res_file),
-         chr = as.integer(sub("(.*_chr)([0-9]+)(\\.rds)$", "\\2", res_file)),
-         pred = furrr::future_map(res_file, readRDS), res_file = NULL) %>%
+library(dplyr)
+library(tidyr)
+pred_ldpred2_gwide <-
+  tibble(res_file = list.files("results/ldpred2-gwide", "\\.rds$", full.names = TRUE)) %>%
+  mutate(Trait = sub("\\.rds$", "", basename(res_file)),
+         pred = furrr::future_map(res_file, ~ as.list(readRDS(.)$pred)),
+         res_file = NULL) %>%
   unnest_longer(col = "pred", indices_to = "Method") %>%
-  filter(Method %in% c("best_beta", "beta_auto")) %>%
-  mutate(Method = case_when(Method == "beta_auto" ~ "LDpred2_auto",
-                            Method == "best_beta" ~ "LDpred2_best")) %>%
-  group_by_at(-(2:3)) %>%
-  summarise(pred = list(do.call(plus, map(pred, replace_na, replace = 0)))) %>%
-  ungroup() %>%
+  mutate(Method = paste0(sub("beta", "ldpred2", Method), "_gwide")) %>%
   print()
 
 pred_sct <-
[…]

### 2020-09-30 code/plot-res-ldpred.R

@@ -2,11 +2,10 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
+load("data/ind_val_test.RData")
 
-library(tidyverse)
+library(dplyr)
+library(tidyr)
 library(future)
 (NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
 plan(multisession, workers = NCORES)
@@ -17,99 +16,127 @@ pred_ldpred2 <-
          chr = as.integer(sub("(.*_chr)([0-9]+)(\\.rds)$", "\\2", res_file)),
          pred = furrr::future_map(res_file, readRDS), res_file = NULL) %>%
   unnest_longer(col = "pred", indices_to = "Method") %>%
+  filter(Method != "best_beta") %>%
+  group_by_at(-(2:3)) %>%
+  summarise(pred = list(do.call(plus, lapply(pred, replace_na, replace = 0)))) %>%
+  ungroup() %>%
+  mutate(Method = paste0(sub("beta", "ldpred2", Method), "_perchr")) %>%
   print()
+
+pred_ldpred2_gwide <-
+  tibble(res_file = list.files("results/ldpred2-gwide", "\\.rds$", full.names = TRUE)) %>%
+  mutate(Trait = sub("\\.rds$", "", basename(res_file)),
+         pred = furrr::future_map(res_file, ~ as.list(readRDS(.)$pred)),
+         res_file = NULL) %>%
+  unnest_longer(col = "pred", indices_to = "Method") %>%
+  mutate(Method = paste0(sub("beta", "ldpred2", Method), "_gwide")) %>%
+  print()
+
 
 AUCBoot_no_NA <- function(pred, target) {
   is_not_na <- which(!is.na(target))
   AUCBoot(pred[is_not_na], target[is_not_na])
 }
 
-res_ldpred2 <- pred_ldpred2 %>%
-  group_by_at(-(2:3)) %>%
[…]

### 2020-09-30 code/plot-res-simus-ldpred.R

@@ -2,40 +2,42 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
-ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+load("data/ind_gwas_val_test.RData")
 
-library(tidyverse)
-library(foreach)
+library(dplyr)
+library(tidyr)
+
 pred_ldpred2 <- tibble(basename = list.files("data/pheno-simu")) %>%
   mutate(
     pheno_file = file.path("data/pheno-simu", basename),
-    res_file = map(file.path("results/simu-ldpred2", basename), ~ {
-      foreach(chr = 1:22, .combine = 'c') %do% {
-        sub("\\.rds$", paste0("_chr", chr, ".rds"), .x)
-      }
-    }))
+    res_file = lapply(file.path("results/simu-ldpred2", basename), function(.x) {
+      sapply(1:22, function(chr) sub("\\.rds$", paste0("_chr", chr, ".rds"), .x))
+    })) %>%
+  print()
 
 res_ldpred2 <- pred_ldpred2 %>%
   mutate(basename = gsubfn::gsubfn("(.+)_([0-9]+)\\.rds$", "\\1 \\2", basename)) %>%
   separate("basename", c("params", "iter"), sep = " ", convert = TRUE) %>%
   unnest_longer(col = "res_file", indices_to = "chr") %>%
-  filter(file.exists(res_file)) %>%
-  mutate(pred = map(res_file, readRDS), res_file = NULL) %>%
+  mutate(pred = lapply(res_file, readRDS), res_file = NULL) %>%
   unnest_longer(col = "pred", indices_to = "Method") %>%
   group_by_at(-(4:5)) %>%
   summarise(pred = list(do.call(plus, pred))) %>%
   ungroup() %>%
-  mutate(Method = sub("beta", "ldpred2", Method),
[…]

### 2020-09-30 code/plot-res-simus2-ldpred.R

@@ -1,0 +1,158 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+load("data/ind_gwas_val_test.RData")
+
+library(dplyr)
+library(tidyr)
+
+pred_ldpred2 <-
+  tibble(basename = list.files("results/simu-ldpred2", "all_40_3000_15")) %>%
+  mutate(
+    pheno_file = file.path("data/pheno-simu", sub("_chr[0-9]+.*", ".rds", basename)),
+    res_file = file.path("results/simu-ldpred2", basename)) %>%
+  print()
+
+res_ldpred2 <- pred_ldpred2 %>%
+  mutate(N = sub(".*_chr[0-9]+(.*)\\.rds$", "\\1", basename),
+         N = ifelse(N == "", 300000L, as.integer(sub("_", "", N))),
+         chr = as.integer(sub(".*_chr([0-9]+).*\\.rds$", "\\1", basename)),
+         basename = NULL) %>%
+  mutate(pred = lapply(res_file, readRDS), res_file = NULL) %>%
+  unnest_longer(col = "pred", indices_to = "Method") %>%
+  group_by(pheno_file, N, Method) %>%
+  summarise(pred = list(do.call(plus, pred))) %>%
+  ungroup() %>%
+  mutate(Method = paste0(sub("beta", "ldpred2", Method), "_perchr"),
+         pheno = lapply(pheno_file, readRDS), pheno_file = NULL,
+         AUC = purrr::map2_dbl(pred, pheno, ~ AUC(.x, .y[ind.test])),
+         pred = NULL, pheno = NULL) %>%
+  print()
+
+res_ldpred2_gwide <-
+  tibble(res_file = list.files("results/simu-ldpred2-gwide/", full.names = TRUE),
+         pheno_file = file.path("data/pheno-simu",
+                                sub("_[0-9]+\\.rds", ".rds", basename(res_file))),
+         N = as.integer(sub(".+_([0-9]+)\\.rds", "\\1", res_file))) %>%
+  mutate(pred = lapply(res_file, function(file) as.list(readRDS(file)$pred)),
+         res_file = NULL) %>%
[…]

### 2020-09-30 code/prepare-corr-spmat.R

@@ -4,7 +4,7 @@ G <- ukb$genotypes
 CHR <- as.integer(ukb$map$chromosome)
 POS <- ukb$map$physical.pos
 
-set.seed(1); ind.val <- sort(sample(nrow(G), 10e3))
+load("data/ind_val_test.RData")
 
 (NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
 POS2 <- snp_asGeneticPos(CHR, POS, dir = "tmp-data", ncores = NCORES)
@@ -49,3 +49,6 @@ for (chr in 1:22) {
 #  788.966    6.565   70.550
 #  861.304    6.443   72.679
 # -> 2.2H in total
+
+sum(file.size(paste0("data/corr/chr", 1:22, ".rds"))) / 1024^3
+# -> 9.5 GB total

### 2020-09-30 code/prepare-genotypes.R

@@ -67,5 +67,14 @@ ukb$map <- dplyr::mutate(ukb$map, chromosome = as.integer(chromosome),
 ukb$fam <- snp_fake(n = nrow(G), m = 1)$fam
 ukb$fam$sample.ID <- sample$ID_2[ind.indiv[sub2]]
 
-set.seed(1); ind.val <- sort(sample(nrow(G), 10e3))
+RNGversion("3.5.1"); set.seed(1)
+ind.val <- sort(sample(nrow(G), 10e3))
+# for simulations
+ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+save(ind.val, ind.gwas, ind.test, file = "data/ind_gwas_val_test.RData")
+# for real applications
+ind.test <- setdiff(rows_along(G), ind.val)
+save(ind.val, ind.test, file = "data/ind_val_test.RData")
+
 snp_writeBed(ukb, bedfile = "data/UKBB_imp_HM3_val.bed", ind.row = ind.val)

### 2020-09-30 code/prepare-simu-phenotypes.R

@@ -51,3 +51,26 @@ purrr::pwalk(params2, function(h2, M, K, iter) {
     saveRDS(simu$pheno, pheno_file)
   }
 })
+
+
+#### In both ####
+
+h2 <- 0.4
+M_HLA <- 300
+M_all <- 10e3
+K <- 0.15
+
+for (iter in 1:10) {
+
+  pheno_file <- sprintf("data/pheno-simu/both_%d_%d_%d.rds",
+                        100 * h2, 100 * K, iter)
+
+  if (!file.exists(pheno_file)) {
+    print(pheno_file)
+    simu1 <- snp_simuPheno(G, h2, M_HLA, ind.possible = ind.HLA)
+    simu2 <- snp_simuPheno(G, h2, M_all, ncores = NCORES)
+    simu_liab <- (simu1$pheno + simu2$pheno) / sqrt(2)
+    simu_pheno <- (simu_liab > qnorm(K, lower.tail = FALSE)) + 0L
+    saveRDS(simu_pheno, pheno_file)
+  }
+}

### 2020-09-30 code/prepare-simu-sumstats.R

@@ -4,36 +4,65 @@ G <- ukb$genotypes
 CHR <- as.integer(ukb$map$chromosome)
 POS <- ukb$map$physical.pos
 
-bigassertr::assert_dir("data/sumstats-simu")
+load("data/ind_gwas_val_test.RData")
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
+#### Simus varying disease architecture ####
 
 pheno_files <- list.files("data/pheno-simu")
-all_pheno <- lapply(pheno_files, function(file) {
-  readRDS(file.path("data/pheno-simu", file))
-})
 
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
   t = "12:00:00", c = "16", mem = "128g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
+bigassertr::assert_dir("data/sumstats-simu")
+
 future.apply::future_lapply(1:22, function(chr) {
 
   ind.chr <- which(CHR == chr)
 
-  purrr::walk2(pheno_files, all_pheno, ~ {
+  purrr::walk(pheno_files, ~ {
 
     gwas_file <- file.path("data/sumstats-simu",
                            sub("\\.rds$", paste0("_chr", chr, ".rds"), .x))
 
     if (!file.exists(gwas_file)) {
       print(gwas_file)
-      gwas <- big_univLogReg(G, .y[ind.gwas], ind.train = ind.gwas,
+      y <- readRDS(file.path("data/pheno-simu", .x))
+      gwas <- big_univLogReg(G, y[ind.gwas], ind.train = ind.gwas,
                              ind.col = ind.chr, ncores = 16)
       saveRDS(gwas, gwas_file)
     }
   })
+})
+
+#### Simus varying sample size ####
+
+grid <- expand.grid(
+  pheno_file = pheno_files[grepl("all_40_3000_15", pheno_files)],
+  N = c(10e3, 20e3, 50e3, 120e3),
+  chr = 1:22
+)
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = "16", mem = "128g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(grid, function(pheno_file, N, chr) {
+
[…]

### 2020-09-30 code/prepare-sumstats.R

@@ -11,14 +11,25 @@ map <- transmute(ukb$map,
 
 bigassertr::assert_dir("data/sumstats")
 
-set.seed(1); ind.val <- sort(sample(nrow(G), 10e3))
+load("data/ind_val_test.RData")
 (NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
 sd <- runonce::save_run(
-  big_parallelize(G, function(X, ind, ind.val) {
-    bigstatsr::big_scale()(X, ind.val, ind)$scale
-  }, p.combine = "c", ncores = NCORES, ind.val = ind.val),
+  sqrt(big_colstats(G, ind.val, ncores = NCORES)$var),
   file = "data/sd.rds"
 )
+
+SUMMARY <- function(info_snp) {
+  chi2 <- with(info_snp, (beta / beta_se)^2)
+  print(round(mean(chi2, na.rm = TRUE), 2))
+  S <- rep(NA, ncol(G)); S[info_snp$`_NUM_ID_`] <- chi2
+  signif <- pchisq(S, df = 1, lower.tail = FALSE) < 5e-8
+  print(sum(signif, na.rm = TRUE))
+  ind.keep <- snp_clumping(
+    G, infos.chr = map$chr, infos.pos = map$pos, S = S,
+    ind.row = ind.val, thr.r2 = 0.01, size = 10e3, ncores = NCORES,
+    exclude = which(is.na(S) | !signif))
+  print(length(ind.keep))
+}
 
 #### Breast cancer (BRCA) ####
 
@@ -43,9 +54,10 @@ info_snp <- bigsnpr::snp_match(sumstats, map, strand_flip = FALSE)
 # 11,792,542 variants to be matched.
 # 1,188,767 variants have been matched; 0 were flipped and 0 were reversed.
 (info_snp <- tidyr::drop_na(as_tibble(info_snp)))
+SUMMARY(info_snp)
 
 sd_val <- sd[info_snp$`_NUM_ID_`]
-sd_ss <- with(info_snp, 1 / sqrt(n_eff / 4 * beta_se^2))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2))
 
 is_bad <-
   sd_ss < (0.5 * sd_val) | sd_ss > (sd_val + 0.1) | sd_ss < 0.1 | sd_val < 0.05
[…]

### 2020-09-30 code/process-betas-ldpred1.R

@@ -2,9 +2,7 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
+load("data/ind_val_test.RData")
 
 library(tidyverse)
 pred_ldpred1 <-

### 2020-09-30 code/process-betas-prscs.R

@@ -2,16 +2,14 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
+load("data/ind_val_test.RData")
 
-library(tidyverse)
+library(dplyr)
 beta_prscs <-
   tibble(res_file = list.files("results/prscs", "\\.txt$", full.names = TRUE)) %>%
   mutate(Trait = sub("(.+/)([[:alnum:]]+)(.+_chr[0-9]+\\.txt)$", "\\2", res_file),
          chr = as.integer(sub("(.+_chr)([0-9]+)(\\.txt)$", "\\2", res_file)),
-         phi = as.double(sub("(.+_phi)(.+)(_chr[0-9]+\\.txt)$", "\\2", res_file))) %>%
+         phi = sub("(.+_phi)(.+)(_chr[0-9]+\\.txt)$", "\\2", res_file)) %>%
   group_by(Trait, phi) %>%
   arrange(chr) %>%
   summarize(beta = list(bigreadr::fread2(res_file))) %>%
@@ -24,13 +22,15 @@ for (k in cols_along(betas)) {
 }
 
 (NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
-bigparallelr::set_blas_ncores(NCORES)
-pred <- big_prodMat(G, betas)
+# bigparallelr::set_blas_ncores(NCORES)
+# pred <- big_prodMat(G, betas)
+pred <- big_parallelize(G, function(X, ind, betas) {
+  bigstatsr::big_prodMat(X, betas[ind, ], ind.col = ind)
+}, p.combine = plus, ncores = NCORES, betas = betas)
 
 beta_prscs$pred <- as.list(as.data.frame(pred))
-beta_prscs$pheno <- map(beta_prscs$Trait, ~ {
-  readRDS(paste0("data/pheno/", ., ".rds"))
-})
+beta_prscs$pheno <-
+  lapply(paste0("data/pheno/", beta_prscs$Trait, ".rds"), readRDS)
 
 AUC_no_NA <- function(pred, target) {
   is_not_na <- which(!is.na(target))
@@ -43,17 +43,23 @@ AUCBoot_no_NA <- function(pred, target) {
 
 library(future)
[…]

### 2020-09-30 code/process-betas-sbayesr.R

@@ -1,0 +1,47 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+
+load("data/ind_val_test.RData")
+
+library(dplyr)
+res_sbayesr <-
+  tibble(res_file = list.files("results/sbayesr", "\\.snpRes$", full.names = TRUE)) %>%
+  mutate(Trait = sub(".+/([[:alnum:]]+)_chr[0-9]+\\.snpRes$", "\\1", res_file),
+         chr = as.integer(sub(".+/[[:alnum:]]+_chr([0-9]+)\\.snpRes", "\\1", res_file))) %>%
+  group_by(Trait) %>%
+  arrange(chr) %>%
+  summarize(beta = list(bigreadr::fread2(res_file))) %>%
+  print()
+
+betas <- matrix(0, ncol(G), nrow(res_sbayesr))
+for (k in cols_along(betas)) {
+  tmp <- res_sbayesr$beta[[k]]
+  ind <- match(tmp$Name, ukb$map$marker.ID)
+  stopifnot(!anyNA(ind))
+  stopifnot(all.equal(ukb$map$allele1[ind], tmp$A2))
+  stopifnot(all.equal(ukb$map$allele2[ind], tmp$A1))
+  betas[ind, k] <- tmp$A1Effect
+}
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+# bigparallelr::set_blas_ncores(NCORES)
+pred <- big_parallelize(G, function(X, ind, betas, ind.test) {
+  bigstatsr::big_prodMat(X, betas[ind, ], ind.row = ind.test, ind.col = ind)
+}, p.combine = plus, ncores = NCORES, betas = betas, ind.test = ind.test)
+res_sbayesr$pred <- as.list(as.data.frame(pred))
+res_sbayesr$beta <- NULL
+
+AUCBoot_no_NA <- function(pred, target) {
+  is_not_na <- which(!is.na(target))
+  AUCBoot(pred[is_not_na], target[is_not_na])
+}
+
+res_sbayesr$AUCBoot <- purrr::pmap(res_sbayesr, function(Trait, pred) {
+  y <- readRDS(paste0("data/pheno/", Trait, ".rds"))
+  AUCBoot_no_NA(pred, y[ind.test])
+})
[…]

### 2020-09-30 code/process-betas-simu-ldpred1.R

@@ -2,40 +2,45 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
-ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+load("data/ind_gwas_val_test.RData")
 
-library(tidyverse)
+library(dplyr)
 pred_ldpred1 <-
-  tibble(res_file = list.files("results/simu-ldpred1", full.names = TRUE),
-       betas = map(res_file, ~ as.list(setNames(as_tibble(readRDS(.x)),
-                                                c(rep("grid", 7), "inf"))))) %>%
-  unnest_longer("betas", indices_to = "method") %>%
+  tibble(
+    res_file = list.files("results/simu-ldpred1", full.names = TRUE),
+    betas = lapply(res_file, function(.x) {
+      as.list(setNames(as_tibble(readRDS(.x)), c(rep("grid", 7), "inf")))
+    })
+  ) %>%
+  tidyr::unnest_longer("betas", indices_to = "method") %>%
   print()
 
 str(all_betas <- do.call("cbind", pred_ldpred1$betas))
 
 (NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
 bigparallelr::set_blas_ncores(NCORES)
-pred <- big_prodMat(G, all_betas)
+pred <- runonce::save_run(big_prodMat(G, all_betas),
+                          file = "tmp-data/all_pred_simu_ldpred1.rds")
 
 pred_ldpred1$pred <- as.list(as.data.frame(pred))
-pred_ldpred1$pheno <- map(pred_ldpred1$res_file, ~ {
-  readRDS(sub("results/simu-ldpred1", "data/pheno-simu", .x, fixed = TRUE))
-})
+library(future.apply)
+plan(multisession, workers = NCORES)
+pheno_files <- sub("results/simu-ldpred1/(.*_15_[0-9]+).*",
+                   "data/pheno-simu/\\1.rds", pred_ldpred1$res_file)
+# all(file.exists(pheno_files))
[…]

### 2020-09-30 code/provide-ld-ref.R

@@ -1,0 +1,100 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+#### Filter data from allele frequency errors ####
+
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+map <- transmute(ukb$map,
+                 chr = as.integer(chromosome), pos = physical.pos,
+                 a0 = allele1, a1 = allele2, rsid)  # reversed somehow..
+
+(NCORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE")) - 1L)
+af <- runonce::save_run(
+  big_colstats(G, ncores = NCORES)$sum / (2 * nrow(G)),
+  file = "data/af.rds"
+)
+
+obj.bed <- bed(download_1000G("tmp-data"))
+map2 <- dplyr::transmute(obj.bed$map, chr = chromosome, pos = physical.pos,
+                         a1 = allele1, a0 = allele2, beta = 1)
+info_snp <- snp_match(map2, map)
+# 1,664,852 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 1,058,123 variants have been matched; 0 were flipped and 0 were reversed.
+
+fam2 <- bigreadr::fread2(sub_bed(obj.bed$bedfile, ".fam2"))
+ind_eur <- which(fam2$`Super Population` == "EUR")
+af2 <- bed_MAF(obj.bed, ind.row = ind_eur, ind.col = info_snp$`_NUM_ID_.ss`,
+               ncores = NCORES)
+
+af_UKBB <- af[info_snp$`_NUM_ID_`]
+af_1000G <- af2$af
+
+N <- nrow(G)
+af_avg <- (af_UKBB * N + af_1000G * 503) / (N + 503)
+chi2 <- (af_UKBB - af_1000G)^2 / (af_avg * (1 - af_avg) * (1 / N + 1 / 503))
+hist(pval <- pchisq(chi2, df = 1, lower.tail = FALSE), breaks = 0:50 / 50)
+
+is_bad <- pval < 1e-5 | af_1000G < 0.01 | af_1000G > 0.99 | af_UKBB < 0.005 | af_UKBB > 0.995
+qplot(af_UKBB, af_1000G, color = is_bad) +
+  theme_bigstatsr() +
+  geom_abline(color = "red") +
+  coord_equal() +
[…]

### 2020-09-30 code/run-lassosum.R

@@ -2,10 +2,7 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
-
+load("data/ind_val_test.RData")
 
 library(dplyr)
 files <- tibble(basename = list.files("data/sumstats"),
@@ -26,19 +23,22 @@ rsid <- bigreadr::fread2("data/UKBB_imp_HM3_val.bim")[[2]]
 library(future.batchtools)
 NCORES <- 16
 plan(batchtools_slurm(resources = list(
-  t = "4:00:00", c = NCORES, mem = "128g",
+  t = "12:00:00", c = NCORES, mem = "128g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
 furrr::future_pmap(files_sub, function(basename, res_file) {
 
+  # basename <- "T1D.rds"
   y <- readRDS(file.path("data/pheno", basename))
   sumstats <- readRDS(file.path("data/sumstats", basename))
+  t <- with(sumstats, beta / beta_se)
+  n <- sumstats$n_eff
 
   library(lassosum)
   doParallel::registerDoParallel(cl <- parallel::makeCluster(NCORES))
   system.time(
     out <- lassosum.pipeline(
-      cor = with(sumstats, beta / beta_se / sqrt(n_eff)),
+      cor = t / sqrt(n - 2 + t^2),
       snp = rsid[sumstats$`_NUM_ID_`],
       A1 = sumstats$a1,
       A2 = sumstats$a0,
@@ -49,6 +49,7 @@ furrr::future_pmap(files_sub, function(basename, res_file) {
       destandardize = TRUE
     )
   ) # 25 min
+  pseudo_out <- pseudovalidate(out, cluster = cl)
   parallel::stopCluster(cl)
 
   auc.val <- apply(-do.call("cbind", out$pgs), 2, AUC_no_NA, target = y[ind.val])
@@ -57,6 +58,9 @@ furrr::future_pmap(files_sub, function(basename, res_file) {
   ind <- which(beta_lassosum != 0)
   pred <- big_prodVec(G, beta_lassosum[ind], ind.row = ind.test,
                       ind.col = sumstats$`_NUM_ID_`[ind])
+  ind <- which(pseudo_out$best.beta != 0)
[…]

### 2020-09-30 code/run-ldpred1.R

@@ -3,10 +3,7 @@ ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 ukb$map$chromosome <- as.integer(ukb$map$chromosome)
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
-
+load("data/ind_val_test.RData")
 
 library(dplyr)
 files <- tibble(basename = list.files("data/sumstats"),

### 2020-09-30 code/run-ldpred2-gwide.R

@@ -1,0 +1,124 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+
+load("data/ind_val_test.RData")
+
+library(dplyr)
+files <- tibble::tibble(basename = list.files("data/sumstats")) %>%
+  mutate(
+    res_file = file.path("results/ldpred2-gwide", basename),
+    pheno_file = file.path("data/pheno",  basename),
+    gwas_file = file.path("data/sumstats", basename)) %>%
+  print()
+all(file.exists(files$pheno_file)) & all(file.exists(files$gwas_file))
+bigassertr::assert_dir("results/ldpred2-gwide")
+
+files_sub <- files %>%
+  filter(!file.exists(res_file)) %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 30
+plan(batchtools_slurm(resources = list(
+  t = "2-00:00", c = NCORES + 2, mem = "250g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub[-1], function(res_file, pheno_file, gwas_file) {
+
+  # res_file <- files_sub$res_file[7]
+  # pheno_file <- files_sub$pheno_file[7]
+  # gwas_file <- files_sub$gwas_file[7]
+
+  y <- readRDS(pheno_file)
+  sumstats <- readRDS(gwas_file)
+
+  ind.val2 <- ind.val[!is.na(y[ind.val])]
+
+  tmp <- tempfile(tmpdir = "tmp-data")
+  on.exit(file.remove(paste0(tmp, ".sbk")), add = TRUE)
+
+  for (chr in 1:22) {
+
+    # print(chr)
+
+    ## indices in 'sumstats'
+    ind.chr <- which(sumstats$chr == chr)
+    ## indices in 'G'
+    ind.chr2 <- sumstats$`_NUM_ID_`[ind.chr]
+    ## indices in 'corr'
+    ind.chr3 <- match(ind.chr2, which(CHR == chr))
+
+    corr0 <- readRDS(paste0("data/corr/chr", chr, ".rds"))[ind.chr3, ind.chr3]
+
+    if (chr == 1) {
+      df_beta <- sumstats[ind.chr, c("beta", "beta_se", "n_eff", "_NUM_ID_")]
+      ld <- Matrix::colSums(corr0^2)
[…]

### 2020-09-30 code/run-ldpred2.R

@@ -3,10 +3,7 @@ ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 CHR <- as.integer(ukb$map$chromosome)
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
-
+load("data/ind_val_test.RData")
 
 library(dplyr)
 files <- tidyr::expand_grid(

### 2020-09-30 code/run-prscs.R

@@ -11,7 +11,7 @@ library(dplyr)
 files <- tidyr::expand_grid(
   basename = list.files("data/sumstats"),
   chr = 1:22,
-  phi = 10^-(6:0)
+  phi = c(10^-(6:0), NA)
 ) %>%
   mutate(
     res_file = sprintf("results/prscs/%s_pst_eff_a1_b0.5_phi%.e_chr%d.txt",
@@ -61,7 +61,7 @@ furrr::future_pmap(files_sub, function(chr, phi, res_file, gwas_file) {
     " --bim_prefix=data/UKBB_imp_HM3_val",
     " --sst_file={tmp}",
     " --n_gwas={round(median(sumstats$n_eff))}",
-    " --phi={phi}",
+    if (is.na(phi)) "" else " --phi={phi}",
     " --chrom={chr}",
     " --out_dir={sub('_pst.+$', '', res_file)}"
   ))

### 2020-09-30 code/run-sbayesr.R

@@ -1,0 +1,89 @@
+# zip <- runonce::download_file(
+#   "https://cnsgenomics.com/software/gctb/download/gctb_2.02_Linux.zip",
+#   dir = "tmp-data")
+# unzip(zip, exdir = "tmp-data")
+gctb <- "tmp-data/gctb_2.02_Linux/gctb"
+# bigsnpr:::make_executable(gctb)
+plink <- bigsnpr::download_plink("tmp-data")
+
+## Compute LD
+
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "2-00:00", c = 2, mem = "125g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+future.apply::future_lapply(1:22, function(chr) {
+  if (!file.exists(paste0("tmp-data/ldm_", chr, ".ldm.shrunk.bin"))) {
+    system(glue::glue(
+      "{plink} --bfile data/UKBB_imp_HM3_val",
+      " --chr {chr} --maf 0.01 --make-bed",
+      " --out tmp-data/myplink_{chr}"
+    ))
+    system.time(
+      system(glue::glue(
+        "{gctb} --bfile tmp-data/myplink_{chr}",
+        " --make-shrunk-ldm",
+        # " --gen-map tmp-data/chr{chr}.OMNI.interpolated_genetic_map", # too few variants matched
+        " --out tmp-data/ldm_{chr}"
+      ))
+    )
+    # 19 min for chr 22 (16410 SNPs)
+    # 1h30 for chr 14 (37031 SNPs)
+    unlink(paste0("tmp-data/myplink_", chr, "*"))
+  }
+}, future.globals = c("gctb", "plink"))
+
+
+## Compute SBayesR
+
+library(dplyr)
+files <- tidyr::expand_grid(
+  basename = list.files("data/sumstats"),
+  chr = 1:22
+) %>%
+  mutate(
+    res_file = paste0("results/sbayesr/", sub("\\.rds$", "", basename), "_chr", chr),
+    gwas_file = file.path("data/sumstats", basename)) %>%
+  print()
+
+bigassertr::assert_dir("results/sbayesr")
+
+files_sub <- files %>%
+  arrange(basename, chr) %>%
+  filter(!file.exists(paste0(res_file, ".snpRes"))) %>%
+  select(-basename) %>%
[…]

### 2020-09-30 code/run-sct.R

@@ -4,10 +4,7 @@ G <- ukb$genotypes
 CHR <- as.integer(ukb$map$chromosome)
 POS <- ukb$map$physical.pos
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.test <- setdiff(rows_along(G), ind.val)
-
+load("data/ind_val_test.RData")
 
 library(dplyr)
 files <- tibble(basename = list.files("data/sumstats"),

### 2020-09-30 code/run-simu-ldpred1.R

@@ -4,15 +4,26 @@ library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
-ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
+load("data/ind_gwas_val_test.RData")
 
 library(dplyr)
-files <- tibble(basename = list.files("data/pheno-simu"),
-                res_file = file.path("results/simu-ldpred1", basename))
-head(files)
+files <- tibble(
+  basename = list.files("data/sumstats-simu"),
+  chr = as.integer(sub(".*_chr([0-9]+).*\\.rds$", "\\1", basename)),
+  N = sub(".*_chr[0-9]+(.*)\\.rds$", "\\1", basename),
+  pheno_file = file.path("data/pheno-simu", sub("_chr[0-9]+.*(\\.rds)$", "\\1", basename)),
+  gwas_file = file.path("data/sumstats-simu", basename),
+  res_file = file.path("results/simu-ldpred1", sub("_chr[0-9]+", "", basename))
+) %>%
+  mutate(N = ifelse(N == "", 300000L, as.integer(sub("_", "", N)))) %>%
+  group_by(pheno_file, N, res_file) %>%
+  arrange(chr) %>%
+  summarize(gwas_files = list(c(gwas_file))) %>%
+  print()
+# files <- tibble(basename = list.files("data/pheno-simu"),
+#                 res_file = file.path("results/simu-ldpred1", basename))
+dim(files)
+table(files$N)
 bigassertr::assert_dir("results/simu-ldpred1")
 
 files_sub <- files %>%
@@ -24,23 +35,27 @@ files_sub <- files %>%
 
 library(future.batchtools)
 plan(batchtools_slurm(resources = list(
-  t = "60:00:00", c = "4", mem = "32g",
+  t = "3-00:00", c = "4", mem = "32g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pmap(files_sub, function(basename, res_file) {
+furrr::future_pmap(files_sub, function(pheno_file, N, res_file, gwas_files) {
 
[…]

### 2020-09-30 code/run-simu-ldpred2-gwide.R

@@ -1,0 +1,137 @@
+library(bigsnpr)
+ukb <- snp_attach("data/UKBB_imp_HM3.rds")
+G <- ukb$genotypes
+CHR <- as.integer(ukb$map$chromosome)
+
+load("data/ind_gwas_val_test.RData")
+
+library(dplyr)
+files <- tibble(
+  basename = list.files("data/sumstats-simu", pattern = "all_40_3000_15"),
+  chr = as.integer(sub(".*_chr([0-9]+).*\\.rds$", "\\1", basename)),
+  N = sub(".*_chr[0-9]+(.*)\\.rds$", "\\1", basename),
+  pheno_file = file.path("data/pheno-simu", sub("_chr[0-9]+.*(\\.rds)$", "\\1", basename)),
+  gwas_file = file.path("data/sumstats-simu", basename)
+) %>%
+  mutate(N = ifelse(N == "", 300000L, as.integer(sub("_", "", N))),
+         res_file = file.path("results/simu-ldpred2-gwide",
+                              paste0(sub("_chr[0-9]+.*", "", basename), "_", N, ".rds"))) %>%
+  group_by(pheno_file, N, res_file) %>%
+  arrange(chr) %>%
+  summarize(gwas_files = list(c(gwas_file))) %>%
+  print()
+table(files$N)
+bigassertr::assert_dir("results/simu-ldpred2-gwide")
+
+files_sub <- files %>%
+  filter(!file.exists(res_file)) %>%
+  # group_by(N) %>% slice(1:3) %>% ungroup() %>%
+  print()
+
+library(future.batchtools)
+NCORES <- 30
+plan(batchtools_slurm(resources = list(
+  t = "2-00:00", c = NCORES + 2, mem = "250g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pmap(files_sub, function(pheno_file, gwas_files, N, res_file) {
+
+  # (pheno_file <- files_sub$pheno_file[7])
+  # (gwas_files <- files_sub$gwas_files[[7]])
+  # (N <- files_sub$N[7])
+  # (res_file <- files_sub$res_file[7])
+
+  y <- readRDS(pheno_file)
+  (Neff <- 4 / (1 / sum(y[ind.gwas] == 0) + 1 / sum(y[ind.gwas] == 1)) * N / 300e3)
+  sumstats <- do.call("rbind", lapply(gwas_files, readRDS)) %>%
[…]

### 2020-09-30 code/run-simu-ldpred2.R

@@ -3,22 +3,21 @@ ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
 CHR <- as.integer(ukb$map$chromosome)
 
-set.seed(1)
-ind.val <- sort(sample(nrow(G), 10e3))
-ind.gwas <- sort(sample(setdiff(rows_along(G), ind.val), 300e3))
-ind.test <- setdiff(rows_along(G), c(ind.gwas, ind.val))
-
+load("data/ind_gwas_val_test.RData")
 
 library(dplyr)
 files <- tibble(
   basename = list.files("data/sumstats-simu"),
   res_file = file.path("results/simu-ldpred2", basename),
-  chr = as.integer(sub("(.*_chr)([0-9]+)(\\.rds)$", "\\2", basename)),
-  pheno_file = file.path("data/pheno-simu", sub("_chr[0-9]+(\\.rds)$", "\\1", basename)),
+  chr = as.integer(sub(".*_chr([0-9]+).*\\.rds$", "\\1", basename)),
+  N = sub(".*_chr[0-9]+(.*)\\.rds$", "\\1", basename),
+  pheno_file = file.path("data/pheno-simu", sub("_chr[0-9]+.*(\\.rds)$", "\\1", basename)),
   gwas_file = file.path("data/sumstats-simu", basename)
-)
+) %>%
+  mutate(N = ifelse(N == "", 300000L, as.integer(sub("_", "", N))))
 head(files)
 table(files$chr)
+table(files$N)
 all(file.exists(files$pheno_file)) & all(file.exists(files$gwas_file))
 bigassertr::assert_dir("results/simu-ldpred2")
 
@@ -30,20 +29,21 @@ files_sub <- files %>%
   print()
 
 library(future.batchtools)
-NCORES <- 8
+NCORES <- 15
 plan(batchtools_slurm(resources = list(
-  t = "6:00:00", c = NCORES, mem = "64g",
+  t = "12:00:00", c = NCORES, mem = "125g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
 
-furrr::future_pmap(files_sub, function(basename, res_file, chr, pheno_file, gwas_file) {
+furrr::future_pmap(files_sub[-1], function(res_file, chr, N, pheno_file, gwas_file) {
 
   y <- readRDS(pheno_file)
   gwas <- readRDS(gwas_file)
-  df_beta <- data.frame(
-    beta = gwas$estim, beta_se = gwas$std.err,
[…]

### 2020-10-01 paper/paper-ldpred2.tex
some text changes

@@
 We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns the two LDpred parameters from data.
-We benchmark predictive performance of LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness, and now providing much more accurate polygenic scores compared to LDpred1.
-We then show that LDpred2 also outperforms other polygenic score methods recently developed, such as lassosum, PRS-CS and SBayesR.
+We benchmark predictive performance of LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and predictive accuracy compared to LDpred1.
+We then show that LDpred2 also outperforms other polygenic score methods recently developed, with a mean AUC over the 8 real traits analyzed here of 65.1\%, compared to 63.8\% for lassosum, 62.9\% for PRS-CS and 61.5\% for SBayesR.
 Note that, in contrast to what was recommended in the first version of this paper, we now recommend to run LDpred2 genome-wide instead of per chromosome.
@@
 Second, we also provide an ``auto'' option, where LDpred2 automatically estimates the sparsity $p$ and the SNP heritability $h^2$, and therefore does not require validation data to tune hyper-parameters.
-We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions, when the proportion of causal variants is small, and when sample size is large.
+We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions, when the proportion of causal variants is small, and when GWAS sample size is large.
 We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing prediction accuracy.
-Moreover, LDpred2-auto, which does not require any validation set, performs almost as well as LDpred2-grid (the main LDpred2 model which tunes hyper-parameters from a grid in a validation set), provided some quality control is performed on the summary statistics.
+Moreover, LDpred2-auto, which does not require any validation set, performs almost as well as the main LDpred2 model that tunes hyper-parameters from a grid in a validation set, provided some quality control is performed on the summary statistics.
 
@@
 For users who do not particularly care about sparsity, they can still tune this parameter in LDpred2-grid to explore a larger parameter space. 
-For this paper, to show the effect of this sparse option on predictive performance, we deliberately separate the LDpred2-grid method in two ``different'' methods that we call LDpred2-grid-nosp (sparse option always disabled) and LDpred2-grid-sp (sparse option always enabled) in the results.
+For this paper, in order to show the effect of this sparse option on predictive performance, we deliberately separate the LDpred2-grid method in two ``different'' methods that we call LDpred2-grid-nosp (sparse option always disabled) and LDpred2-grid-sp (sparse option always enabled) in the results.
 As a recall, LDpred v1 has two options: LDpred1-inf, and LDpred1-grid where only $p$ is tuned while $h^2$ is estimated from constrained LD score regression \cite[]{bulik2015ld}.
@@
 We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
-We also compare running LDpred2 (including choosing optimal hyper-paramaters) per chromosome (``perchr'') or genome-wide (``gwide'').
+We also compare running LDpred2 per chromosome (``perchr'', including choosing optimal hyper-paramaters) or genome-wide (``gwide'').
 Finally, we compare LDpred2 to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum, PRS-CS and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
@@
 As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are located in a long-range LD region. 
-In contrast, all LDpred2 models perform well in these two HLA scenarios, but LDpred2-auto performs slightly worse than other LDpred2 models in these extreme scenarios.
+In contrast, all LDpred2 models perform well in these two HLA scenarios, but LDpred2-auto performs slightly worse than the other LDpred2 models in these extreme scenarios.
 
@@
 Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
-[TODO: MAKE SURE ABOUT NAMING OF BOOTSTRAP TECHNIQUE + ADD CORRESPONDING TABLES IN SUPPMAT WHEN CODE OUT]
-%See corresponding values in table S1.
+Corresponding values are reported in table S1.
 }
@@
 Figure \ref{fig:AUC-simu2-grid} presents the simulation results comparing LDpred1-grid and LDpred2-grid when varying the GWAS sample size from 10,000 to 300,000 in the simulation scenario with 3000 causal variants anywhere on the genome (``all\_40\_3000'').
-We can see that LDpred1-grid performs equally well as LDpred2-grid (when LDpred2 is run genome-wide) for the smallest GWAS sample sizes. 
-Yet, LDpred1-grid starts providing lower predictive performance than LDpred2-grid when sample size becomes larger. 
+LDpred1-grid performs equally well as LDpred2-grid (when LDpred2 is run genome-wide) for the smallest GWAS sample sizes. 
+Yet, LDpred1-grid starts providing lower predictive performance than LDpred2-grid when GWAS sample size becomes larger. 
 For example, in the scenario with a sample size of 120,000, LDpred1 provides a mean AUC of 70.7\% while LDpred2 provides a mean AUC of 74.6\%.
@@
 	Bars present the mean and 95\% CI of 10,000 non-parametric bootstrap replicates of the mean AUC of 10 simulations for each scenario. 
-	%See corresponding values in table S1.
+	Corresponding values are reported in table S2.
 	}
@@
 For Asthma and RA, predictive performance of LDpred1 and LDpred2 are similar.
-As in simulations, the sparse version of LDpred2-grid performs as well as the non-sparse version.
-Sparsity of resulting effects ranges from 19.3\% for RA to 54.4\% for Asthma [TODO: ADD REF TO TABLE THAT SHOWS THE SPARSITY OF THE CORRESPONDING MODEL WHEN OUT -> SEE SCREENSHOT IN SUPPMAT FOR NOW]
+As in simulations, the sparse version of LDpred2-grid performs as well as the non-sparse version. Sparsity of resulting effects ranges from 19.3\% for RA to 54.4\% for Asthma (Table S5).
 
@@
 	Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
-	[TODO: MAKE SURE ABOUT NAMING OF BOOTSTRAP TECHNIQUE + ADD CORRESPONDING TABLES IN SUPPMAT WHEN CODE OUT]
-	%See corresponding values in table S2.
+	Corresponding values are reported in table S3.
 	}
@@
 	Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
-	%See corresponding values in table S2.
+	Corresponding values are reported in table S4.
 	}
@@
 Similarly, lassosum-auto performs similarly as lassosum, expect for MDD (AUC of 54.9\% vs 59.4\%). 
-Similarly, PRS-CS-auto also underperforms for MDD (53.9\% for PRS-CS-auto vs AUC of 57.5\% for PRS-CS).
+Similarly, PRS-CS-auto also underperforms for MDD (AUC of 53.9\% for PRS-CS-auto vs 57.5\% for PRS-CS).
 In contrast, SBayesR does not underperforms for MDD but severely underperforms for T1D and RA, as discussed in the previous paragraph.
@@
 For example, in the simulation scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from 57.2\% to 57.7\%, LDpred2-grid from 58.6\% to 59.6\%, and LDpred2-auto from 57.3\% to 59.3\%.
-We note that however that the simulation scenario used to produce results in figure \ref{fig:AUC-simu2-ldpred2} is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
+Note that the simulation scenario used to produce results in figure \ref{fig:AUC-simu2-ldpred2} is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
 However, figure \ref{fig:AUC-real-ldpred2} also shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
-Indeed, except for LDpred2-inf applied to T1D, all other cases show similar or better performance for running all variants of LDpred2 genome-wide. 
+Indeed, except for LDpred2-inf applied to T1D, all other cases show similar or better performance when running LDpred2 genome-wide. 
 For example, running LDpred2-grid genome-wide rather than per chromosome is particularly beneficial for CAD (AUC of 63.6\% vs 61.5\%) and PRCA (AUC of 70.2\% vs 68.2\%).
@@
 Yet, it has instability issues that have been pointed out \cite[]{marquez2018modeling,lloyd2019improved} and likely contributed to discrepancies in reported predictive accuracy of LDpred \cite[]{choi2019prsice,ge2019polygenic,chun2020non}.
-We have therefore implemented a new version of LDpred, LDpred2, to address these issues and improve its computational efficiency.
+We have therefore implemented a new version of LDpred, LDpred2, to address these issues.
 We show that LDpred2 is more stable and provides higher predictive performance than LDpred1, particularly when handling long-range LD regions, less polygenic traits, and large GWAS sample sizes.
@@
 This enables LDpred2 to work well even when causal variants are located in long-range LD regions.
-We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved}. 
-Indeed, these regions, especially the HLA region, contain many variants associated with many traits, and are therefore very useful for prediction.
+We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved} since these regions, especially the HLA region, contain many variants associated with many traits, and are therefore very useful for prediction.
 Other modifications, such as a better handling of numerical errors when working with exponentials (equation \eqref{eq:postp}), or a larger hyper-parameter search space, may also contribute to the improved robustness in LDpred2.
 
-In LDpred2, we expand the grid of hyper-parameters examined with now more values for $p$ (21 instead of 7 by default in LDpred1) and for $h^2$ (3 instead of 1).
+In LDpred2, we also expand the grid of hyper-parameters examined with now more values for $p$ (21 instead of 7 by default in LDpred1) and for $h^2$ (3 instead of 1).
 When testing the grid of hyper-parameters of $p$ and $h^2$, we also allow for testing an option to enable sparse models in LDpred2 (see below).
@@
 In practice, the sparse version of LDpred2 can drastically reduce the number of variants used in the PGS without impacting its predictive performance, as opposed to discarding the smallest effects after having run LDpred (as tested in \cite{bolli2019software}).
-The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, therefore which does not require a validation set to choose hyper-parameters, making it an attractive option in many applications.
+The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, which therefore does not require any validation set to tune hyper-parameters from. 
+LDpred2-auto is an attractive option in many applications, especially since we now also provide an LD reference that can be used directly. 
 LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications, provided that some quality control is performed on summary statistics (see Methods section ``Quality control of summary statistics'').
@@
 Causal variants are chosen randomly anywhere on the genome. 
-Liability scores are computed from a model with additive effects only: we compute the liability score of the $i$-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G}_{i,j} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele count of individual $i$ for variant $j$, $\widetilde{G}_{i,j}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon_i$ follows a Gaussian distribution $N(0, 1 - h^2)$.
+Liability scores are computed from a model with additive effects only: we compute the liability score of the $i$-th individual as \(y_i = \sum_{j\in S_\text{causal}} w_j \widetilde{G}_{i,j} + \epsilon_i,\) where $S_\text{causal}$ is the set of causal variants, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$, $G_{i,j}$ is the allele dosage of individual $i$ for variant $j$, $\widetilde{G}_{i,j}$ corresponds to its standardized version (zero mean and unit variance), and $\epsilon_i$ follows a Gaussian distribution $N(0, 1 - h^2)$.
 Both parts of the $y_i$'s are scaled such that the variance of the genetic liability is exactly $h^2$ and the variance of the total liability is exactly $1$. 
@@
 The 95\% confidence interval (CI) from 10,000 non-parametric bootstrap replicates of the mean AUC of the 10 simulations for each scenario is also reported.
+In other words, we sample 10,000 bootstrap replicates of these 10 AUC values and compute their respective mean. 
+We then report the mean of these 10,000 values, along with their quantile at 2.5\% and at 97.5\% to act as the 95\% confidence interval (CI) for the mean AUC.
 
@@
 For more details, please refer to our R code (Software and code availability section).
-
-[TODO: TALK ABOUT AUC + CI]
 
@@
 Then, we compare LDpred2-inf, LDpred2-grid and LDpred2-auto to several other methods: Clumping and Thresholding (C+T), Stacked C+T (SCT), lassosum(-auto), PRS-CS(-auto) and SBayesR \cite[]{prive2019making,mak2017polygenic,ge2019polygenic,lloyd2019improved}.
-For C+T and SCT \cite[]{prive2019making}, we use the default large grid of hyper-parameters testing a threshold of clumping $r_c^2$ within \{0.01, 0.05, 0.1, 0.2, 0.5, 0.8, 0.95\}, a base size of clumping window within \{50, 100, 200, 500\} in Kbp where the actual window size is then computed as the base size divided by $r_c^2$, and a sequence of 50 thresholds on -log10(p-values) between 0.1 and the most significant p-value, equally spaced on a log scale.
+For C+T and SCT \cite[]{prive2019making}, we use the default large grid of hyper-parameters testing a threshold of clumping $r_c^2$ within \{0.01, 0.05, 0.1, 0.2, 0.5, 0.8, 0.95\}, a base size of clumping window within \{50, 100, 200, 500\} in Kb where the actual window size is then computed as the base size divided by $r_c^2$, and a sequence of 50 thresholds on -log10(p-values) between 0.1 and the most significant p-value, equally spaced on a log scale.
 For lassosum \cite[]{mak2017polygenic}, we use the default grid of hyper-parameters $s$ and $\lambda$. 
@@
 PRS-CS-auto automatically estimates $\Phi$.
+Finally, for SBayesR \cite[]{lloyd2019improved}, shrunk LD matrices are built using option `\texttt{-{}-make-shrunk-ldm}' and SBayesR is run with parameters `\texttt{-{}-pi 0.95,0.02,0.02,0.01}' and `\texttt{-{}-gamma 0.0,0.01,0.1,1}', and a chain of length 10,000 with 2000 burn-in iterations.
 Note that we use the 10,000 individuals of the validation set as LD reference for all methods, except for PRS-CS which provides its own LD reference based on 503 individuals from the 1000 Genomes data; \cite{ge2019polygenic} argued that the sample size of the LD reference has little impact on the performance of PRS-CS.
-Finally, for SBayesR \cite[]{lloyd2019improved}, shrunk LD matrices are built using option `\texttt{-{}-make-shrunk-ldm}' and SBayesR is run with parameters `\texttt{-{}-pi 0.95,0.02,0.02,0.01}' and `\texttt{-{}-gamma 0.0,0.01,0.1,1}', and a chain of length 10,000 with 2000 burn-in iterations.
+
+We use the Area Under the ROC Curve (AUC) to compare methods.
+We sample 10,000 bootstrap replicates of the individuals in the test set and compute the AUC for each of these. 
+We then report the mean of these 10,000 values, along with their quantile at 2.5\% and at 97.5\% to act as the 95\% confidence interval (CI) for the AUC.
+This is implemented in function \texttt{AUCBoot} of R package bigstatsr.
 
@@
 In total, this grid is of size $21 \times 3 \times 2 = 126$.
-When running LDpred2-auto, we run it 30 times with $h^2_\text{LDSC}$ as initial value for $h^2$ and a sequence of 30 values from $10^{-4}$ to $0.9$ on a log-scale as initial values for $p$.
-Running many Gibbs chains aims at checking that models did not diverge (and hopefully converged).
+When running LDpred2-auto, we run it 30 times with $h^2_\text{LDSC}$ as initial value for $h^2$ and a sequence of 30 values from $10^{-4}$ to $0.9$ equally spaced on a log scale as initial values for $p$.
+Running many Gibbs chains aims at checking whether models did not diverge (and hopefully converged).
 As a criterion for non-divergence, we compute the standard deviations of the resulting predictors from the 30 models, keep only those within 3 median absolute deviations from their median, and average the remaining vectors of effects as final effect sizes for the ``auto'' version.
@@
 Doubling to 2M variants would require 128 GB of RAM to store the matrix.
-Second, as noted in \cite{prive2019making}, it may be beneficial to assume that the architecture of traits is different for different chromosomes.
+Second, it may be beneficial to assume that the architecture of traits is different for different chromosomes.
 For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
 Assuming the same model for genetic effects genome-wide could result in model misspecification, which would lead to suboptimal predictive performance.
-Moreover, since the inverse of a block-diagonal matrix is formed from the inverse of each block, it should be safe to run LDpred2 for each chromosome and then combine the results (at least for LDpred-inf).
 To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
@@
 
-%[TODO: EXPORT CODE FROM CLUSTER]
+The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
+A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
 
 All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}. 
-The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
+We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
 

### 2020-10-01 paper/paper-ldpred2.tex
prepare for bioRxiv

@@
 
-[TODO: ALSO TALK ABOUT LDPRED2 PARAMS AND POSSIBLY FUTURE WORK OF IMPROVING THEM VIA E.G. FUNCTIONAL ANNOTATIONS?]
-
-[TODO: TALK ABOUT THE WRONG H2 ESTIMATION FOR CAD IN DISCU? -> CF. TABLE + SOME FIGURES IN SUPPMAT]
-
 The previous version of LDpred has been widely used and has the potential to provide polygenic models with good predictive performance \cite[]{khera2018genome}. 
@@
 LDpred2-auto is an attractive option in many applications, especially since we now also provide an LD reference that can be used directly. 
-LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications, provided that some quality control is performed on summary statistics (see Methods section ``Quality control of summary statistics'').
+LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications.
+
+However, LDpred2-auto requires that some quality control is performed on summary statistics (see Methods section ``Quality control of summary statistics''). 
 This quality control aims at ensuring that effects are transferable from the external GWAS summary statistics to the data where PGS are computed.
+Note that this quality control is also beneficial for other methods such as lassosum (data not shown).
+Even when this quality control is performed, LDpred2-auto can slightly underperform in some situations, e.g. for CAD and Asthma here (Figure \ref{fig:AUC-all}).
+When looking more closely at the results for CAD, even though chains seemed to have converged (Figure S5), we can see that the heritability estimate from LDpred2-auto is off compared to the one from LD score regression (Table S5), and that its estimate of $p$ (\textasciitilde0.001) is off compared to the best performing $p$ in LDpred2-grid (\textasciitilde0.01 in figure S4).
+One possible explanation for this discrepancy is that the summary statistics we use for CAD come from a meta-analysis of 48 small studies, with some of them from non-European ancestries.
+Moreover, the summary statistics we use for Asthma come from a meta-analysis of 66 small multiancestry studies.
+This could break some of the assumptions used in LDpred2-auto and explain why LDpred2-auto underperforms for CAD and Asthma compared to LDpred2-grid.
+Future work is needed to study how to make best use of these multiancestry meta-analyses from many small studies in the context of genetic prediction.
+Future work is also needed to assess how relevant is the estimation of parameters from LDpred2-auto, and whether its estimates and its predictive performance could be improved by e.g.
+incorporating functional annotations.
 
@@
 
-We provide an LD reference for European ancestry to be used by researchers who do not have access to a large dataset to be used as LD reference.
+We provide an LD reference for European ancestry to be used by researchers who cannot compute their own.
 We use the 362,320 UKBB individuals as used here, with some further quality control based on allele frequencies.
 Indeed, some large allele frequency mismatches have been reported recently in the UKBB when comparing to other datasets, which are apparently due to mismappings \cite[]{kunert2020allele}.
+Note that most of these errors should be captured by the quality control we propose in this paper.
 Using 503 European individuals from the 1000 Genomes (1000G) data, we remove variants with allele frequency differences between UKBB and 1000G at $p<10^{-5}$ (Figure \ref{fig:af-outliers}).
@@
 A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
+The European LD reference we provide here is available at \url{https://doi.org/10.6084/m9.figshare.13034123}.
 

### 2020-10-03 paper/paper-ldpred2.tex
finished?

@@
 We then show that LDpred2 also outperforms other polygenic score methods recently developed, with a mean AUC over the 8 real traits analyzed here of 65.1\%, compared to 63.8\% for lassosum, 62.9\% for PRS-CS and 61.5\% for SBayesR.
-Note that, in contrast to what was recommended in the first version of this paper, we now recommend to run LDpred2 genome-wide instead of per chromosome.
+Note that LDpred2 provides more accurate polygenic scores when run genome-wide, instead of per chromosome.
 LDpred2 is implemented in R package bigsnpr.
@@
 We show that LDpred2 provides higher predictive performance than LDpred1 (LDpred v1.0.0), especially when there are causal variants in long-range LD regions, when the proportion of causal variants is small, and when GWAS sample size is large.
-We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing prediction accuracy.
+We also show that the new sparse option performs equally well as the non-sparse version, enabling LDpred2 to provide sparse effects without losing any predictive accuracy.
 Moreover, LDpred2-auto, which does not require any validation set, performs almost as well as the main LDpred2 model that tunes hyper-parameters from a grid in a validation set, provided some quality control is performed on the summary statistics.
@@
 
-We make use of the UK Biobank data to compare the two versions of LDpred using several simulation scenarios to understand the expected impact of using the new version of LDpred.
+We use the UK Biobank data to compare the two versions of LDpred using several simulation scenarios to understand the expected impact of using the new version of LDpred.
 We also compare these two versions of LDpred using eight case-control phenotypes of major interest and for which there are published external summary statistics available and a substantial number of cases in the UK Biobank data.
@@
 As expected, LDpred1 performs poorly in HLA scenarios, i.e.\ when causal variants are located in a long-range LD region. 
-In contrast, all LDpred2 models perform well in these two HLA scenarios, but LDpred2-auto performs slightly worse than the other LDpred2 models in these extreme scenarios.
+In contrast, all LDpred2 models perform well in these two HLA scenarios, but LDpred2-auto performs slightly worse.
+There are two possible explanations for this: first, these two scenarios are very extreme scenarios where causal effects may have large effects in strong LD; second, because there is no contribution from the other chromosomes in these scenarios, any prediction added to the score from other chromosomes would be noise that could reduce the performance of the overall score.
+In the ``both\_40'' simulation scenario, where effects in HLA are generated with a smaller heritability (20\%) and where there are also other effects all over the genome contributing another 20\% of heritability, predictive performance of LDpred2-auto are on par with the other LDpred2 models.
 
@@
 
-Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome, contrary to was recommended in the first version of this paper.
+Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome.
 Figure \ref{fig:AUC-simu2-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS sample size is small.
@@
 We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved} since these regions, especially the HLA region, contain many variants associated with many traits, and are therefore very useful for prediction.
+As sample size grows, the Gibbs sampler used in LDpred relies more on the LD being calculated and adjusted for appropriately, providing an intuitive explanation for why LDpred1 performs much worse than LDpred2 when GWAS sample size is large.
 Other modifications, such as a better handling of numerical errors when working with exponentials (equation \eqref{eq:postp}), or a larger hyper-parameter search space, may also contribute to the improved robustness in LDpred2.
@@
 As an example, it takes less than three hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
-It takes less than 36 hours to run all these LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide).
+It takes between 12 and 36 hours to run all these LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide); LDpred1 takes a similar time.
 It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
@@
 The European LD reference we provide here is available at \url{https://doi.org/10.6084/m9.figshare.13034123}. 
+We note that this is LD summary data and not individual-level data.
 

### 2020-11-11 code/plot-approx-sd.R
add two figures

@@ -1,3 +1,5 @@
+#### Simulations ####
+
 library(bigsnpr)
 ukb <- snp_attach("data/UKBB_imp_HM3.rds")
 G <- ukb$genotypes
@@ -5,8 +7,6 @@ G <- ukb$genotypes
 load("data/ind_gwas_val_test.RData")
 
 sd <- readRDS("data/sd.rds")
-
-#### Simulations ####
 
 sumstats_simu <- do.call("rbind", lapply(1:22, function(chr) {
   readRDS(paste0("data/sumstats-simu/all_40_300_15_1_chr", chr, ".rds"))
@@ -26,4 +26,82 @@ qplot(sd_val, sd_ss, alpha = I(0.5)) +
 
 # ggsave("figures/sd-approx-simu.png", width = 7, height = 7)
 
-## See prepare-sumstats.R for BRCA
+
+#### Real data ####
+
+library(bigsnpr)
+library(bigreadr)
+
+## Information for the variants provided in the LD reference
+info <- readRDS(url("https://github.com/privefl/bigsnpr/raw/master/data-raw/hm3_variants.rds"))
+str(info)
+
+#### Coronary artery disease (CAD) ####
+
+# download.file("ftp://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/NikpayM_26343387_GCST003116/cad.add.160614.website.txt",
+#               destfile = "tmp-data/sumstats_CAD.txt")
+sumstats <- fread2("tmp-data/sumstats_CAD.txt",
+                   select = c("chr", "bp_hg19", "noneffect_allele",
+                              "effect_allele", "beta", "se_dgc", "p_dgc"),
+                   col.names = c("chr", "pos", "a0", "a1", "beta", "beta_se", "p"))
+sumstats$n_eff <- 4 / (1 / 60801 + 1 / 123504)
+
+info_snp <- snp_match(sumstats, info)
+# 9,455,778 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 1,116,009 variants have been matched; 0 were flipped and 1,011,595 were reversed.
+(info_snp <- tidyr::drop_na(tibble::as_tibble(info_snp)))
+
+sd_ldref <- with(info_snp, sqrt(2 * maf * (1 - maf)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2))
+
+is_bad <-
[…]

### 2020-11-24 code/plot-res-simus-ldpred.R
update fig1

@@ -58,19 +58,38 @@ library(ggplot2)
 cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
                "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 
-ggplot(auc_simu, aes(params, mean, fill = Method, color = Method)) +
-  bigstatsr::theme_bigstatsr() +
-  geom_hline(yintercept = 0.5, linetype = 2) +
-  geom_col(position = position_dodge(), alpha = 0.6, color = "black", size = 1) +
-  geom_errorbar(aes(ymin = inf, ymax = sup),
-                position = position_dodge(width = 0.9), color = "black", width = 0.2, size = 1) +
-  scale_y_continuous(limits = c(0.5, 0.92), minor_breaks = 0:50 / 50,
-                     oob = scales::rescale_none) +
-  labs(x = "Simulation", y = "AUC") +
-  theme(legend.position = c(0.45, 0.8)) +
-  scale_fill_manual(values = cbPalette[c(6, 7, 4, 2, 5, 8)])
+plot_grid(
+  ggplot(filter(auc_simu, startsWith(params, "all_")),
+         aes(params, mean, fill = Method, color = Method)) +
+    bigstatsr::theme_bigstatsr() +
+    geom_hline(yintercept = 0.5, linetype = 2) +
+    geom_col(position = position_dodge(), alpha = 0.6, color = "black", size = 1) +
+    geom_errorbar(aes(ymin = inf, ymax = sup),
+                  position = position_dodge(width = 0.9), color = "black", width = 0.2, size = 1) +
+    scale_y_continuous(limits = c(0.5, 0.85), minor_breaks = 0:50 / 50,
+                       oob = scales::rescale_none) +
+    labs(x = "Simulation (with effects everywhere)", y = "AUC") +
+    theme(legend.position = c(0.76, 0.8)) +
+    scale_fill_manual(values = cbPalette[c(6, 7, 4, 2, 5, 8)]),
 
-# ggsave("figures/AUC-simu.pdf", width = 12.5, height = 6.5)
+  ggplot(filter(auc_simu, !startsWith(params, "all_")),
+         aes(params, mean, fill = Method, color = Method)) +
+    bigstatsr::theme_bigstatsr() +
+    geom_hline(yintercept = 0.5, linetype = 2) +
[…]

### 2020-11-24 paper/paper-ldpred2.tex
minor changes to text

@@
 The heritability, i.e.\ the proportion of phenotypic variance that is attributable to genetics, determines an upper limit on the predictive performance of PGS and thus their value as a predictive tool.
-Nevertheless, a number of studies have explored the use of PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
+{\color{red} Nevertheless, a number of studies have shown the potential benefits of using PGS in clinical settings} \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
 PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association,barth2020genetic,harden2020using}. 
@@
 Seven simulation scenarios are used, each repeated 10 times.
+{\color{red}300,000 individuals are used for computing the GWAS summary statistics, 10,000 for computing the LD matrix and choosing the best-performing hyper-parameters, and the remaining 52,320 individuals for testing.}
 In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome.
@@
 Similarly, PRS-CS-auto also underperforms for MDD (AUC of 53.9\% for PRS-CS-auto vs 57.5\% for PRS-CS).
-In contrast, SBayesR does not underperforms for MDD but severely underperforms for T1D and RA, as discussed in the previous paragraph.
+In contrast, SBayesR does not underperform for MDD but severely underperforms for T1D and RA, as discussed in the previous paragraph.
 
@@
 We show that LDpred2 is more stable and provides higher predictive performance than LDpred1, particularly when handling long-range LD regions, less polygenic traits, and large GWAS sample sizes.
+{\color{red} Now that the Gibbs sampler in LDpred2 is more stable, it seems that the infinitesimal model never outperforms the grid model (which includes an infinitesimal model when using $p=1$).}
 We hypothesize that LDpred1 does not use a LD window size that is large enough to account for long-range LD such as in the HLA region.
-In LDpred2, we use a window size of 3 cM, which is larger than the default value used in LDpred1.
-This enables LDpred2 to work well even when causal variants are located in long-range LD regions.
-We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved} since these regions, especially the HLA region, contain many variants associated with many traits, and are therefore very useful for prediction.
+In LDpred2, we use a window size of 3 cM, which enables LDpred2 to work well even when causal variants are located in long-range LD regions.
+We argue against removing these regions as sometimes suggested in the literature \cite[]{marquez2018modeling,lloyd2019improved} since these regions, especially the HLA region, contain many variants associated with many traits, and are therefore often useful for prediction.
 As sample size grows, the Gibbs sampler used in LDpred relies more on the LD being calculated and adjusted for appropriately, providing an intuitive explanation for why LDpred1 performs much worse than LDpred2 when GWAS sample size is large.
@@
 Efficient parallelization over the grid of hyper-parameters is possible because we use an on-disk sparse matrix format accessed using memory mapping. 
-This special data format is available in R package bigsparser, which has been developed for this paper. 
-As an example, it takes less than three hours in total to run LDpred2-inf, LDpred2-grid (126 hyper-parameter values parallelized over 8 cores) and LDpred2-auto (30 initial values for $p$ parallelized over 8 cores) for a chromosome with 100,000 variants.
-It takes between 12 and 36 hours to run all these LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide); LDpred1 takes a similar time.
-It takes 11 minutes to pre-compute the LD matrix for 10,000 individuals and 100,000 variants (using 16 cores).
+This new data format is available in R package bigsparser, which has been developed for this paper. 
+{\color{red}It takes between 12 and 36 hours to run all the LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide) using 30 cores; LDpred1 takes a similar time (note that LDpred1 cannot be easily run in parallel).
+It takes a little more than two hours to pre-compute the LD matrix for 10,000 individuals and 1.1 million HapMap3 variants (using 16 cores).
+This pre-computation is fast compared to running the Gibbs sampler in LDpred2 and needs to be performed only once for all traits.
+It can also be further parallelized on different compute nodes for different chromosomes; then it takes only 11 minutes for the largest chromosome.}
 
@@
 The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, which therefore does not require any validation set to tune hyper-parameters from. 
-LDpred2-auto is an attractive option in many applications, especially since we now also provide an LD reference that can be used directly.
+LDpred2-auto is an attractive option in many applications, especially since we also provide an LD reference that can be used directly {\color{red} (Section ``Software and code availability'')}.
 LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications.
@@
 This could break some of the assumptions used in LDpred2-auto and explain why LDpred2-auto underperforms for CAD and Asthma compared to LDpred2-grid. 
+{\color{red} For example, for these two phenotypes, the summary statistics are off compared to the expected identity line in the QC screening procedure presented here (see figures S8 and S9 compared to S7).}
 Future work is needed to study how to make best use of these multiancestry meta-analyses from many small studies in the context of genetic prediction.
-Future work is also needed to assess how relevant is the estimation of parameters from LDpred2-auto, and whether its estimates and its predictive performance could be improved by e.g.
+Future work is also needed to assess how relevant the estimation of parameters from LDpred2-auto is, and whether its estimates and its predictive performance could be improved by e.g.
 incorporating functional annotations.
@@
 For the other software that require a PLINK bed file as input, we use function \texttt{snp\_writeBed} that round dosages to write genotype data in bed format.
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank; these individuals are unrelated and have passed some quality control \cite[]{bycroft2018uk}.
+{\color{red}
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).}
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
@@
 We use external published GWAS summary statistics listed in table \ref{tab:sumstats}.
-We defined phenotypes as in \cite{prive2019making}. Some quality control is applied to summary statistics (see Method section ``Quality control of summary statistics'' below).
+We defined phenotypes as in \cite{prive2019making}. 
+{\color{red} Briefly, we use self-reported illness codes (field 20001 for cancers and 20002 otherwise) and ICD10 codes (fields 40001, 40002, 41202 and 41204 for all diseases and field 40006 specifically for cancers).}
+Some quality control is applied to summary statistics (see Method section ``Quality control of summary statistics'' below).
 For more details, please refer to our R code (Software and code availability section).
@@
 \end{equation}
-where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. We recommend to verify this assumption and to perform some quality control.
-Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure \ref{fig:sd-approx-simu}). 
-However, in real data applications, where summary statistics come from a meta-analysis of many external datasets, this approximation can be invalidated (Figure \ref{fig:sd-approx-brca}).
+where $n_\text{eff} = \dfrac{4	}{1 / n_\text{case} + 1 / n_\text{control}}$. We strongly recommend to verify this assumption and to perform some quality control.
+Indeed, in simulations, the approximation of equation \eqref{eq:approx-sd} seems valid (Figure S6). 
+However, in real data applications, where summary statistics come from a meta-analysis of many external datasets, this approximation can be invalidated (Figure S7).
 Let us denote by $\text{SD}_\text{ss}$ the standard deviations derived from the summary statistics (right-hand side of equation \eqref{eq:approx-sd-lin} or \eqref{eq:approx-sd}) and by $\text{SD}_\text{val}$ the standard deviations of genotypes of individuals in the validation set (left-hand side).
-Note that, in order to compute $\text{SD}_\text{ss}$ in the case of summary statistics from a linear regression, $\text{sd}(\boldsymbol{y})$ from equation \eqref{eq:approx-sd-lin} can be estimated using e.g.\ the median value of $\text{sd}(\boldsymbol{G_j}) \cdot \text{se}(\hat{\gamma}_j) \cdot \sqrt{n}$ or by using the fact that the maximum value of $\text{SD}_\text{ss}$ should be $\sqrt{0.5}$.
-We recommend removing variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure \ref{fig:sd-approx-brca}).
+Note that, in order to compute $\text{SD}_\text{ss}$ in the case of summary statistics from a linear regression, $\text{sd}(\boldsymbol{y})$ from equation \eqref{eq:approx-sd-lin} can be estimated e.g.\ by the median value of $\left(\text{sd}(\boldsymbol{G_j}) \cdot \text{se}(\hat{\gamma}_j) \cdot \sqrt{n}\right)$ or by using the fact that the maximum value of $\text{SD}_\text{ss}$ should be $\sqrt{0.5}$, i.e.\ estimating $\text{sd}(\boldsymbol{y})$ by $\text{min}\left(\sqrt{0.5}\cdot \text{se}(\hat{\gamma}_j) \cdot \sqrt{n}\right)$.
+We recommend removing variants with $\text{SD}_\text{ss} < 0.5 \cdot \text{SD}_\text{val}$ or $\text{SD}_\text{ss} > 0.1 + \text{SD}_\text{val}$ or $\text{SD}_\text{ss} < 0.1$ or $\text{SD}_\text{val} < 0.05$ (Figure S7).
 
@@
 In this paper, we investigate if it would be beneficial to run LDpred2 per chromosome, instead of genome-wide.
-Indeed, running LDpred2 genome-wide has two drawbacks.
-First, it can be memory and computationally demanding to do so.
-For around one million (1M) variants, storing the 1M$\times$1M sparse correlation matrix takes more than 32 GB of memory.
-Doubling to 2M variants would require 128 GB of RAM to store the matrix.
-Second, it may be beneficial to assume that the architecture of traits is different for different chromosomes.
+{\color{red} 
+It may be beneficial for some phenotypes to assume that their genetic architecture is different for different chromosomes.
 For example, chromosome 6 clearly encompasses a larger proportion of the heritability of autoimmune diseases compared to other chromosomes \cite[]{shi2016contrasting}.
-Assuming the same model for genetic effects genome-wide could result in model misspecification, which would lead to suboptimal predictive performance.
+Assuming the same model for genetic effects genome-wide could result in model misspecification, which may lead to suboptimal predictive performance.
+Moreover, running LDpred2 per chromosome allows to process chromosomes on different compute nodes in parallel.
+Note that, to run LDpred2-grid-gwide, one could estimate the heritability genome-wide, run LDpred2-grid per chromosome, then combine scores for all chromosomes for each hyper-parameter set, and finally choose the final genome-wide score corresponding to the best-performing set of hyper-parameters.
+However, for the sake of simplicity, because computations of LDpred2-auto-gwide cannot be easily split like this, we recommend to run all LDpred2-gwide models using the full LD matrix.
+When restricting to HapMap3 variants, the full sparse LD matrix is around 30 GB.}
 To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
@@
 Note that most of these errors should be captured by the quality control we propose in this paper.
-Using 503 European individuals from the 1000 Genomes (1000G) data, we remove variants with allele frequency differences between UKBB and 1000G at $p<10^{-5}$ (Figure \ref{fig:af-outliers}).
+Using 503 European individuals from the 1000 Genomes (1000G) data, we remove variants with allele frequency differences between UKBB and 1000G at $p<10^{-5}$ (Figure S10).
 We also remove variants with minor allele frequencies less than 1\% in the 1000G or less than 0.5\% in the UKBB.

### 2020-11-26 paper/cabios-template/main.tex
use bioinfo template

--- 2020-11-24 paper/paper-ldpred2.tex
@@
-%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english, 12pt]{article}
-\usepackage{times}
-%\usepackage{algorithm2e}
-\usepackage{url}
-\usepackage{bbm}
-\usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{amsmath, amsthm, amssymb}
-\usepackage{setspace}
-\usepackage{lineno}
-\usepackage{hyperref}
-\usepackage{bbm}
-\usepackage{makecell}
+\documentclass{bioinfo}
+\copyrightyear{2015} \pubyear{2015}
 
-%\renewcommand{\arraystretch}{1.8}
+%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%% ADDED BY AUTHOR %%%%
 
-\usepackage{xr}
-\externaldocument{paper-ldpred2-supp}
-
-%\linenumbers
-%\doublespacing
-\onehalfspacing
-%\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
-
-%Pour les rajouts
-\usepackage{color}
-\definecolor{trustcolor}{rgb}{0,0,1}
-
-\usepackage{dsfont}
-\usepackage[warn]{textcomp}
-\usepackage{adjustbox}
-\usepackage{multirow}
-\usepackage{graphicx}
-\graphicspath{{../figures/}}
-\DeclareMathOperator*{\argmin}{\arg\!\min}
 \usepackage{algorithm} 
@@
 
-\let\tabbeg\tabular
-\let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\makeatletter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Bold symbol macro for standard LaTeX users
-%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}
-
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
-\renewcommand*{\arraystretch}{1.2}
-
-\usepackage{babel}
-\makeatother
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Manuscript Category}
 
 \begin{document}
+\firstpage{1}
+
+\subtitle{Subject Section}
 
 \title{LDpred2: better, faster, stronger}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,3,}*}$}
+\author[Priv{\'e} \textit{et~al}.]{Florian Priv\'e,$^{\text{1,}*}$ Julyan Arbel,$^{\text{2}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,3,}*}$}
+\address{$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
+$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
+$^{\text{\sf 3}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark.}
 
-\date{~ }
+\corresp{$^\ast$To whom correspondence should be addressed.}
+
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
+\abstract{\textbf{Motivation:} Polygenic scores have become a central tool in human genetics research.
+LDpred is a popular method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
+However, LDpred has limitations that may reduce its predictive performance.\\
+\textbf{Results:} 
+Here we present LDpred2, a new version of LDpred that addresses these issues.
+We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns the two LDpred parameters from data.We benchmark predictive performance of LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and predictive accuracy compared to LDpred1.
+We then show that LDpred2 also outperforms other polygenic score methods recently developed, with a mean AUC over the 8 real traits analyzed here of 65.1\%, compared to 63.8\% for lassosum, 62.9\% for PRS-CS and 61.5\% for SBayesR.
+Note that LDpred2 provides more accurate polygenic scores when run genome-wide, instead of per chromosome.\\
+\textbf{Availability:} LDpred2 is implemented in R package bigsnpr.\\
+\textbf{Contacts:} \href{florian.prive.21@gmail.com}{florian.prive.21@gmail.com} and \href{bjv@econ.au.dk}{bjv@econ.au.dk}\\
+\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
+online.}
+
 \maketitle
-
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, 38000, France. \\
-\noindent$^{\text{\sf 3}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
-
-\noindent Contacts:
-\begin{itemize}
-\item \url{florian.prive.21@gmail.com}
-\item \url{bjv@econ.au.dk}
-\end{itemize}
-
-\vspace*{4em}
-
-\abstract{	
-Polygenic scores have become a central tool in human genetics research.
-LDpred is a popular method for deriving polygenic scores based on summary statistics and a matrix of correlation between genetic variants.
-However, LDpred has limitations that may reduce its predictive performance.
-Here we present LDpred2, a new version of LDpred that addresses these issues.
-We also provide two new options in LDpred2: a ``sparse'' option that can learn effects that are exactly 0, and an ``auto'' option that directly learns the two LDpred parameters from data.
-We benchmark predictive performance of LDpred2 against the previous version on simulated and real data, demonstrating substantial improvements in robustness and predictive accuracy compared to LDpred1.
-We then show that LDpred2 also outperforms other polygenic score methods recently developed, with a mean AUC over the 8 real traits analyzed here of 65.1\%, compared to 63.8\% for lassosum, 62.9\% for PRS-CS and 61.5\% for SBayesR.
-Note that LDpred2 provides more accurate polygenic scores when run genome-wide, instead of per chromosome.
-LDpred2 is implemented in R package bigsnpr.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
 
@@
 The heritability, i.e.\ the proportion of phenotypic variance that is attributable to genetics, determines an upper limit on the predictive performance of PGS and thus their value as a predictive tool.
-{\color{red} Nevertheless, a number of studies have shown the potential benefits of using PGS in clinical settings} \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
+Nevertheless, a number of studies have shown the potential benefits of using PGS in clinical settings \cite[]{pashayan2015implications,willoughby2019genetic,abraham2019genomic}.
 PGS are also extensively used in epidemiology and economics as predictive variables of interest \cite[]{musliner2015polygenic,horsdal2019association,barth2020genetic,harden2020using}. 
@@
 
+%\enlargethispage{12pt}
+
+\section{Approach}
+
 Here, we present LDpred2, a new version of LDpred that addresses these issues while markedly improving its computational efficiency, allowing exploring a larger grid of parameters in the same computational time as LDpred1.
@@
 Moreover, LDpred2-auto, which does not require any validation set, performs almost as well as the main LDpred2 model that tunes hyper-parameters from a grid in a validation set, provided some quality control is performed on the summary statistics.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Seven simulation scenarios are used, each repeated 10 times.
-{\color{red}300,000 individuals are used for computing the GWAS summary statistics, 10,000 for computing the LD matrix and choosing the best-performing hyper-parameters, and the remaining 52,320 individuals for testing.}
+300,000 individuals are used for computing the GWAS summary statistics, 10,000 for computing the LD matrix and choosing the best-performing hyper-parameters, and the remaining 52,320 individuals for testing.
 In the first four simulation scenarios, a heritability $h^2$ of 40\% and a prevalence of 15\% are used. We simulate 300, 3000, 30,000 or 300,000 causal variants anywhere on the genome.
@@
 %% FIGURE 1 HERE
-\begin{figure}[htb]
-\centerline{\includegraphics[width=0.9\textwidth]{AUC-simu}}
+\begin{figure}[htbp]
+	\centering
+	\includegraphics[width=0.95\linewidth]{AUC-simu}
 	\caption{Two variants of LDpred1 are compared with four variants of LDpred2 (run per chromosome) in the seven simulation scenarios summarized in table \ref{tab:simus}.
@@
 %% FIGURE 2 HERE
-\begin{figure}[htb]
-	\centerline{\includegraphics[width=0.8\textwidth]{AUC-simu2-grid}}
+\begin{figure}[htbp]
+	\centering
+	\includegraphics[width=0.95\linewidth]{AUC-simu2-grid}
 	\caption{LDpred1-grid is compared to LDpred2-grid when varying GWAS sample size in scenario ``all\_40\_3000''.  
@@
 %% FIGURE 3 HERE
-\begin{figure}[htbp]
-	\centerline{\includegraphics[width=0.9\textwidth]{AUC-real}}
+\begin{figure}[htpb]
+	\centering
+	\includegraphics[width=0.95\linewidth]{AUC-real}
 	\caption{Two variants of LDpred1 are compared with four variants of LDpred2 (run genome-wide) in the real data applications using published external summary statistics.
 	Bars present AUC values on the test set of UKBB (mean and 95\% CI from 10,000 bootstrap samples).
-	Corresponding values are reported in table S3.
-	}
-	\label{fig:AUC-real}
+	Corresponding values are reported in table S3. \label{fig:AUC-real}}
 \end{figure}
@@
 \begin{figure}[htbp]
-	\centerline{\includegraphics[width=0.95\textwidth]{AUC-all}}
+	\centering
+	\includegraphics[width=0.95\linewidth]{AUC-all}
 	\caption{LDpred2 is compared with C+T, SCT, lassosum, PRS-CS and SBayesR in the real data applications using published external summary statistics.
@@
 
-Figure \ref{fig:AUC-auto} presents the results of real data applications comparing ``grid'' models with their ``auto'' counterparts, i.e.\ models that directly estimate parameters from the data and do not require tuning hyper-parameters from a grid using a validation set.
+Figure S1 presents the results of real data applications comparing ``grid'' models with their ``auto'' counterparts, i.e.\ models that directly estimate parameters from the data and do not require tuning hyper-parameters from a grid using a validation set.
 We remind readers that SBayesR is such an ``auto'' model, but does not have any ``grid'' counterpart.
@@
 Most results show that it is beneficial to run LDpred2 genome-wide instead of per chromosome.
-Figure \ref{fig:AUC-simu2-ldpred2} shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS sample size is small.
+Figure S2 shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome, especially when GWAS sample size is small.
 For example, in the simulation scenario with the sample size of 20,000, switching running LDpred2 per chromosome to running it genome-wide improves AUC of LDpred2-inf from 57.2\% to 57.7\%, LDpred2-grid from 58.6\% to 59.6\%, and LDpred2-auto from 57.3\% to 59.3\%.
-Note that the simulation scenario used to produce results in figure \ref{fig:AUC-simu2-ldpred2} is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
-However, figure \ref{fig:AUC-real-ldpred2} also shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
+Note that the simulation scenario used to produce results in figure S2 is assuming the same genome-wide disease architecture, i.e.\ 3000 causal variants located anywhere on the genome and whose effects are drawn from the same normal distribution.
+However, figure S3 also shows that it can be beneficial to run all LDpred2 models genome-wide rather than per chromosome for real traits as well. 
 Indeed, except for LDpred2-inf applied to T1D, all other cases show similar or better performance when running LDpred2 genome-wide. 
 For example, running LDpred2-grid genome-wide rather than per chromosome is particularly beneficial for CAD (AUC of 63.6\% vs 61.5\%) and PRCA (AUC of 70.2\% vs 68.2\%).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 We show that LDpred2 is more stable and provides higher predictive performance than LDpred1, particularly when handling long-range LD regions, less polygenic traits, and large GWAS sample sizes.
-{\color{red} Now that the Gibbs sampler in LDpred2 is more stable, it seems that the infinitesimal model never outperforms the grid model (which includes an infinitesimal model when using $p=1$).}
+Now that the Gibbs sampler in LDpred2 is more stable, it seems that the infinitesimal model never outperforms the grid model (which includes an infinitesimal model when using $p=1$).
 We hypothesize that LDpred1 does not use a LD window size that is large enough to account for long-range LD such as in the HLA region.
@@
 This new data format is available in R package bigsparser, which has been developed for this paper. 
-{\color{red}It takes between 12 and 36 hours to run all the LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide) using 30 cores; LDpred1 takes a similar time (note that LDpred1 cannot be easily run in parallel).
+It takes between 12 and 36 hours to run all the LDpred2 models over 1.1 million HapMap3 variants at once (genome-wide) using 30 cores; LDpred1 takes a similar time (note that LDpred1 cannot be easily run in parallel).
 It takes a little more than two hours to pre-compute the LD matrix for 10,000 individuals and 1.1 million HapMap3 variants (using 16 cores).
 This pre-computation is fast compared to running the Gibbs sampler in LDpred2 and needs to be performed only once for all traits.
-It can also be further parallelized on different compute nodes for different chromosomes; then it takes only 11 minutes for the largest chromosome.}
+It can also be further parallelized on different compute nodes for different chromosomes; then it takes only 11 minutes for the largest chromosome.
 
@@
 The second extension is LDpred2-auto, which automatically estimates values for hyper-parameters $p$ and $h^2$, which therefore does not require any validation set to tune hyper-parameters from. 
-LDpred2-auto is an attractive option in many applications, especially since we also provide an LD reference that can be used directly {\color{red} (Section ``Software and code availability'')}.
+LDpred2-auto is an attractive option in many applications, especially since we also provide an LD reference that can be used directly (Section ``Software and code availability'').
 LDpred2-auto almost always performs equally well as LDpred2-grid in simulations as well as in real data applications.
@@
 This could break some of the assumptions used in LDpred2-auto and explain why LDpred2-auto underperforms for CAD and Asthma compared to LDpred2-grid. 
-{\color{red} For example, for these two phenotypes, the summary statistics are off compared to the expected identity line in the QC screening procedure presented here (see figures S8 and S9 compared to S7).}
+For example, for these two phenotypes, the summary statistics are off compared to the expected identity line in the QC screening procedure presented here (see figures S8 and S9 compared to S7).
 Future work is needed to study how to make best use of these multiancestry meta-analyses from many small studies in the context of genetic prediction.
@@
 
-In conclusion, we have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
+\section{Conclusion}
+
+We have shown that LDpred2 provides a better, faster and more robust implementation of the LDpred model.
 When compared to recently derived methods that showed higher predictive performance than LDpred1, LDpred2 provides the highest predictive performance for all the real traits tested here.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\begin{methods}
 \section{Methods}
@@
 For the other software that require a PLINK bed file as input, we use function \texttt{snp\_writeBed} that round dosages to write genotype data in bed format.
-{\color{red}
 We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
-These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).}
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.
@@
 
-\begin{table}[ht]
+\begin{table*}[htpb]
 	\centering
@@
 	\caption{The seven simulations scenarios. N: GWAS sample size. \label{tab:simus}}
-\end{table}
+\end{table*}
 
@@
 We defined phenotypes as in \cite{prive2019making}. 
-{\color{red} Briefly, we use self-reported illness codes (field 20001 for cancers and 20002 otherwise) and ICD10 codes (fields 40001, 40002, 41202 and 41204 for all diseases and field 40006 specifically for cancers).}
+Briefly, we use self-reported illness codes (field 20001 for cancers and 20002 otherwise) and ICD10 codes (fields 40001, 40002, 41202 and 41204 for all diseases and field 40006 specifically for cancers).
 Some quality control is applied to summary statistics (see Method section ``Quality control of summary statistics'' below).
@@
 %% TABLE 1 HERE
-\begin{table}[h]
+\begin{table*}[htpb]
 	\centering
@@
 \label{tab:sumstats}}
-\end{table}
+\end{table*}
 
@@
 \[\boldsymbol{\hat{\gamma}_{\text{marg}}} = \dfrac{1}{n-1} \boldsymbol{S}^{-2} \boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{y} ~.\] 
-We further note that the correlation matrix of $\boldsymbol{G}$ is \(\boldsymbol{R} =  \dfrac{1}{n-1} \boldsymbol{S}^{-1} \boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{G}  \boldsymbol{S}^{-1}\).
+We further note that the correlation matrix of $\boldsymbol{G}$ is \[\boldsymbol{R} =  \dfrac{1}{n-1} \boldsymbol{S}^{-1} \boldsymbol{G}^T \boldsymbol{C_n} \boldsymbol{G}  \boldsymbol{S}^{-1} ~.\]
 Then we get 
@@
 Then, the probability that variant $j$ is causal is computed as 
-\[\bar{p}_j = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{n}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{n}}\right\rbrace}~,\]
+\begin{equation*}
+\begin{split}
+\bar{p}_j & = P\left(\beta_j \sim \mathcal N(\cdot,\cdot) ~|~ \tilde{\beta}_j\right) \\
+& = \dfrac{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace}{\phmpn \exp\left\lbrace-\dfrac{1}{2} \betahmpn\right\rbrace + \dfrac{1-p}{\sqrt{\dfrac{1}{n}}} \exp\left\lbrace-\dfrac{1}{2} \dfrac{\tilde{\beta}_j^2}{\dfrac{1}{n}}\right\rbrace}~,
+\end{split}
+\end{equation*}
 which we rewrite as
@@
 In this paper, we investigate if it would be beneficial to run LDpred2 per chromosome, instead of genome-wide.
-{\color{red} 
 It may be beneficial for some phenotypes to assume that their genetic architecture is different for different chromosomes.
@@
 However, for the sake of simplicity, because computations of LDpred2-auto-gwide cannot be easily split like this, we recommend to run all LDpred2-gwide models using the full LD matrix.
-When restricting to HapMap3 variants, the full sparse LD matrix is around 30 GB.}
+When restricting to HapMap3 variants, the full sparse LD matrix is around 30 GB.
 To choose the best LDpred2 model, one can choose the best model according to their preferred criterion (e.g. max AUC). 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\end{methods}
 
-\clearpage
-%\vspace*{5em}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-A tutorial on the steps to run LDpred2 using some small example data is available at \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
-The European LD reference we provide here is available at \url{https://doi.org/10.6084/m9.figshare.13034123}. 
+The newest version of R package bigsnpr can be installed from GitHub (see \href{https://github.com/privefl/bigsnpr}{https://github.com/privefl/bigsnpr}).
+A tutorial on the steps to run LDpred2 using some small example data is available at \href{https://privefl.github.io/bigsnpr/articles/LDpred2.html}{https://privefl.github.io/bigsnpr/articles/LDpred2.html}.
+The European LD reference we provide here is available at \href{https://doi.org/10.6084/m9.figshare.13034123}{https://doi.org/10.6084/m9.figshare.13034123}. 
 We note that this is LD summary data and not individual-level data.
 
-All code used for this paper is available at \url{https://github.com/privefl/paper-ldpred2/tree/master/code}. 
+All code used for this paper is available at \href{https://github.com/privefl/paper-ldpred2/tree/master/code}{https://github.com/privefl/paper-ldpred2/tree/master/code}. 
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
+
+\vspace*{-12pt}
 
@@
 This research has been conducted using the UK Biobank Resource under Application Number 41181.
+
+\vspace*{-12pt}
 
@@
 
-\section*{Declaration of Interests}
+\noindent\emph{Conflict of Interest:} The authors declare no competing interests.
 
-The authors declare no competing interests.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
+%\vspace*{-12pt}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}
